### 2014-10-20 tex/NoSQL_table.tex
tex init

\documentclass{article}
\usepackage{graphicx}
\usepackage{pdflscape}

\begin{document}
Hello
\begin{landscape}
Hello
\end{landscape}
\end{document}

### 2014-10-21 tex/NoSQL_table.tex
NoSQL table

@@
 \usepackage{pdflscape}
+\usepackage{rotating}
+\usepackage[a4paper,margin=1in,landscape]{geometry}
+\usepackage{booktabs}
+\usepackage{array}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
 \begin{document}
-Hello
-\begin{landscape}
-Hello
-\end{landscape}
+%\begin{landscape}
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[h]
+\begin{tabular}{@{}l| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}|@{}}
+\toprule
+                                    & \textbf{MongoDB}                                                                                                                                                & \textbf{Cassandra}                                                                                                       & \textbf{Dynamo}                               & \textbf{BigTable}                                                           & \textbf{HBase}                                                                     & \textbf{CouchBase Server}                                                           & \textbf{Cloudera Impala}            \\ \midrule
+\textbf{NoSQL-type}                 & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & Key  Value                                   & Wide columnar store                                                         & See BigTable                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                \\ \midrule
+\textbf{Distributed}                & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Y, P2P                                        &                                                                             & Masterslave or multi-master, asynchronous replication.
+& Multi-master                                                                        & Y, not 100\% symmetric              \\ \midrule
+\textbf{Storage}                    & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          &                                               & GFS                                                                         & HDFS                                                                               &                                                                                     & HDFS or HBase                       \\ \midrule
+\textbf{Indexing}                   & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                       & Only primary, but secondary in MegaStore                                    & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      \\ \midrule
+\textbf{Query optimization}         & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     &                                               & Bloom filter                                                                & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     \\ \midrule
+\textbf{Partitioning}               & Range partitioning based on shard key                                                                                                                           & Consistent hashing and range partitioning                                                                                & Consistent hashing                            & Range-partitioning                                                          & Range-partitioning                                                                 & Hashing function determines bucket.
+& By default not, but possible{[}8{]} \\ \midrule
+\textbf{Consistency}                & Configurable                                                                                                                                                    & Configurable                                                                                                             & Configurable                                  & Eventual                                                                    & Strong{[}9{]}                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     \\ \midrule
+\textbf{Transactions / Concurrency} & Single document operations are atomic, otherwise multi-document transactions{[}10{]} Concurrent reads, exclusive writes (lock on DB level).
+Sharding possible & Normally only eventual consistency, but CQL plugin offers Lightweight transactions{[}11{]} Row-level atomicity{[}12{]} & Optimistic or pessimistic concurrency control & Single-row transactions, multiversioning, Distributed lock manager (Chubby) & Single-row transactions (ACID), OCC with MVCC for wider scope transactions{[}13{]} & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     \\ \midrule
+\textbf{Open-source?}               & Y                                                                                                                                                               & Y                                                                                                                        & N                                             & N, see HBase                                                                & Y                                                                                  & Y                                                                                   & Y                                   \\ \midrule
+\textbf{Written in}                 & C++                                                                                                                                                             & Java                                                                                                                     & Java                                          &                                                                             & Java                                                                               &                                                                                     & C++                                 \\ \bottomrule
+\end{tabular}
+\end{table}
+
+%\end{landscape}
 \end{document}

### 2014-10-21 tex/NoSQL_table.tex
Dynamo & BigTable removed

--- 2014-10-20 tex/NoSQL_table.tex
@@
 \usepackage{pdflscape}
+\usepackage{rotating}
+\usepackage[a4paper,margin=.5in,landscape]{geometry}
+\usepackage{booktabs}
+\usepackage{array}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
 \begin{document}
-Hello
-\begin{landscape}
-Hello
-\end{landscape}
+%\begin{landscape}
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}l| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}|@{}}
+\toprule
+                                    & \textbf{MongoDB}                                                                                                                                                & \textbf{Cassandra}                                                                                                       & \textbf{HBase}                                                                     & \textbf{CouchBase Server}                                                           & \textbf{Cloudera Impala}            \\ \midrule
+\textbf{NoSQL-type}                 & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & See BigTable                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                \\ \midrule
+\textbf{Distributed}                & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
+& Multi-master                                                                        & Y, not 100\% symmetric              \\ \midrule
+\textbf{Storage}                    & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       \\ \midrule
+\textbf{Indexing}                   & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      \\ \midrule
+\textbf{Query optimization}         & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     \\ \midrule
+\textbf{Partitioning}               & Range partitioning based on shard key                                                                                                                           & Consistent hashing and range partitioning                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
+& By default not, but possible{[}8{]} \\ \midrule
+\textbf{Consistency}                & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong{[}9{]}                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     \\ \midrule
+\textbf{Transactions / Concurrency} & Single document operations are atomic, otherwise multi-document transactions{[}10{]} Concurrent reads, exclusive writes (lock on DB level).
+Sharding possible & Normally only eventual consistency, but CQL plugin offers Lightweight transactions{[}11{]} Row-level atomicity{[}12{]} & Single-row transactions (ACID), OCC with MVCC for wider scope transactions{[}13{]} & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     \\ \midrule
+\textbf{Open-source?}               & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   \\ \midrule
+\textbf{Written in}                 & C++                                                                                                                                                             & Java                                                                                                                     & Java                                                                               &                                                                                     & C++                                 \\ \bottomrule
+\end{tabular}
+\end{table}
+
+%\end{landscape}
 \end{document}

### 2014-10-26 tex/NoSQL_table.tex
VoltDB added to table

--- 2014-10-20 tex/NoSQL_table.tex
@@
 \usepackage{pdflscape}
+\usepackage{rotating}
+\usepackage[a4paper,margin=.5in,landscape]{geometry}
+\usepackage{booktabs}
+\usepackage{array}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
 \begin{document}
-Hello
-\begin{landscape}
-Hello
-\end{landscape}
+%\begin{landscape}
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[h]
+\centering
+\begin{tabular}{@{}L{2.5cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}| L{3cm}|@{}}
+\cmidrule(l){2-7}
+                                                & MongoDB                                                                                                                                                         & Cassandra                                                                                                                & HBase                                                                              & CouchBase Server                                                                    & Cloudera Impala                     & VoltDB                                                                    \\ \midrule
+\multicolumn{1}{|l|}{NoSQL-type}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & See BigTable                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory NewSQL                                                          \\ \midrule
+\multicolumn{1}{|l|}{Distributed}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
+& Multi-master                                                                        & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at thesame time.
+\\ \midrule
+\multicolumn{1}{|l|}{Storage}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{Indexing}                  & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
+\multicolumn{1}{|l|}{Query optimization}        & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{Partitioning}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing and range partitioning                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
+& By default not, but possible{[}8{]} & Consistent hashing                                                        \\ \midrule
+\multicolumn{1}{|l|}{Consistency}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong{[}9{]}                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
+\multicolumn{1}{|l|}{Transactions, Concurrency} & Single document operations are atomic, otherwise multi-document transactions{[}10{]} Concurrent reads, exclusive writes (lock on DB level).
+Sharding possible & Normally only eventual consistency, but CQL plugin offers Lightweight transactions{[}11{]} Row-level atomicity{[}12{]} & Single-row transactions (ACID), OCC with MVCC for wider scope transactions{[}13{]} & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{Open-source?}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y                                                                         \\ \midrule
+\multicolumn{1}{|l|}{Written in}                & C++                                                                                                                                                             & Java                                                                                                                     & Java                                                                               &                                                                                     & C++                                 & Java                                                                      \\ \bottomrule
+\end{tabular}
+\end{table}
+
+%\end{landscape}
 \end{document}

### 2014-11-11 paper_dec/paper_dec.tex
Start 1st paper

\documentclass{article}
\usepackage{cite}

\begin{document}
\subsection{Cassandra}

Apache Cassandra, originally developped at Facebook but later open-sourced, combines the data-model of Google's BigTable system with the architecture and distribution strategy of Amazon's DynamoDB. It is intended for flexible, highly-available storage of very large datasets, running on cheap commodity hardware and offering high write throughput while not sacrificing read efficiency \cite{lakshman2010cassandra}.\\\\

%datamodel en CQL

%distribution + concurrency + CAP dink
In order to be able to scale linearly in the number of nodes to very large datasets, Cassandra operates in a fully masterless fashion. In terms of the CAP-theorem, it focusses on availability and partition-tolerance, rather than immediate consistency (though the user has control over the level of consistency, as will be explained). %TODO concurrency (quorum, CAS, row-atomicity) ; tunable consistency; failure detection/handling

%extra's (Spark, Solr)

%afronden

\bibliography{biblio}{}
\bibliographystyle{plain}
\end{document}

### 2014-11-11 tex/NoSQL_table.tex
Start 1st paper

@@
 \cmidrule(l){2-7}
-                                                & MongoDB                                                                                                                                                         & Cassandra                                                                                                                & HBase                                                                              & CouchBase Server                                                                    & Cloudera Impala                     & VoltDB                                                                    \\ \midrule
-\multicolumn{1}{|l|}{NoSQL-type}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & See BigTable                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory NewSQL                                                          \\ \midrule
-\multicolumn{1}{|l|}{Distributed}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
+                                                & \textbf{MongoDB}                                                                                                                                                         & \textbf{Cassandra}                                                                                                                & \textbf{HBase}                                                                              & \textbf{CouchBase Server}                                                                    & \textbf{Cloudera Impala}                     & \textbf{VoltDB}                                                                    \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Type}}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & See BigTable                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory NewSQL                                                          \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
 & Multi-master                                                                        & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at thesame time.
 \\ \midrule
-\multicolumn{1}{|l|}{Storage}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{Indexing}                  & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
-\multicolumn{1}{|l|}{Query optimization}        & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{Partitioning}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing and range partitioning                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
+\multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing and range partitioning                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
 & By default not, but possible{[}8{]} & Consistent hashing                                                        \\ \midrule
-\multicolumn{1}{|l|}{Consistency}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong{[}9{]}                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{Transactions, Concurrency} & Single document operations are atomic, otherwise multi-document transactions{[}10{]} Concurrent reads, exclusive writes (lock on DB level).
-Sharding possible & Normally only eventual consistency, but CQL plugin offers Lightweight transactions{[}11{]} Row-level atomicity{[}12{]} & Single-row transactions (ACID), OCC with MVCC for wider scope transactions{[}13{]} & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
-\multicolumn{1}{|l|}{Open-source?}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y                                                                         \\ \midrule
-\multicolumn{1}{|l|}{Written in}                & C++                                                                                                                                                             & Java                                                                                                                     & Java                                                                               &                                                                                     & C++                                 & Java                                                                      \\ \bottomrule
+\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong{[}9{]}                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS{[}11{]} Row-level atomicity{[}12{]} & Single-row transactions (ACID), OCC with MVCC for wider scope transactions{[}13{]} & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y                                                                         \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Written in}}                & C++                                                                                                                                                             & Java                                                                                                                     & Java                                                                               &                                                                                     & C++                                 & Java                                                                      \\ \bottomrule
 \end{tabular}

### 2014-11-12 NoSQL_tabel/NoSQL_table.tex
1e sneuvelversie cassandra af

--- 2014-11-11 tex/NoSQL_table.tex
@@
 \multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing and range partitioning                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
+\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
 & By default not, but possible{[}8{]} & Consistent hashing                                                        \\ \midrule

### 2014-11-12 paper_dec/paper_dec.tex
1e sneuvelversie cassandra af

@@
 \usepackage{cite}
+\usepackage{url}
 
 \begin{document}
-\subsection{Cassandra}
+\section{Cassandra}
 
 Apache Cassandra, originally developped at Facebook but later open-sourced, combines the data-model of Google's BigTable system with the architecture and distribution strategy of Amazon's DynamoDB.
-It is intended for flexible, highly-available storage of very large datasets, running on cheap commodity hardware and offering high write throughput while not sacrificing read efficiency \cite{lakshman2010cassandra}.\\\\
-
-%datamodel en CQL
-
-%distribution + concurrency + CAP dink
-In order to be able to scale linearly in the number of nodes to very large datasets, Cassandra operates in a fully masterless fashion.
+It is intended for flexible, highly-available storage of very large datasets, running on cheap commodity hardware and offering high write throughput while not sacrificing read efficiency \cite{lakshman2010cassandra}.
+\\
+Since its inception Cassandra has however diverged slightly from the BigTable data model.
+It now offers tables and composite columns -much like in a conventional schema, and comes with its own query language, CQL\cite{cassandra_then&now}.
+CQL resembles SQL in many ways, but comes with some restrictions.
+For instance, it doesn't feature the JOIN clause \cite{cassandra_cql}.
+It strongly encourages physically collocating data that will be queried together, and supports denormalization with features such as collection types.\\
+In order to scale linearly in the number of nodes to very large datasets, Cassandra operates in a fully masterless fashion.
 In terms of the CAP-theorem, it focusses on availability and partition-tolerance, rather than immediate consistency (though the user has control over the level of consistency, as will be explained).
-%TODO concurrency (quorum, CAS, row-atomicity) ; tunable consistency; failure detection/handling
-
-%extra's (Spark, Solr)
+High availability and partition tolerance are achieved through asynchronous replication of rows over several nodes in the cluster, using consistent hashing and virtual nodes to handle high churn and incremental addition of nodes \cite{decandia2007dynamo} \cite{lakshman2010cassandra} \cite{cassandra_then&now}.
+The amount of replicas can be chosen by the client.
+Furthermore, Cassandra provides cross-datacenter replication to cope with entire datacenter failures.
+\\On reads and writes, the client can specify the desired quorum, that is the number of replicas that acknowledge the operation.
+Although Cassandra was built with eventual consistency in mind, strong consistency can be obtained by choosing the quorum larger than the number of replicas \cite{grolinger2013data}.\\
+In terms of concurrency control, Cassandra offers atomicity for single-row operations and serializable \textit{lightweight transactions}, essentially a compare-and-set functionality for larger operations \cite{cassandra_lightweight_trans}.\\
+Being an open-source project, Cassandra is freely available, but there is an enterprise version with extra features such as integration with Apache Spark and Apache Solr, for complex analytical and search tasks \cite{cassandra_solr} \cite{cassandra_spark}.\\\\
+In conclusion, Cassandra offers flexible data-modeling with decent querying and indexing support through its CQL-interface and scales incrementally to vast datasets, thanks to its extensive replication and failure-handling features.
 

### 2014-11-12 paper_dec/paper_dec.tex
Eerste beetje HBase

@@
 \begin{document}
-\section{Cassandra}
+\section{Columnar stores}
+\subsection{Cassandra}
 
@@
 
-%afronden
+\subsection{HBase}
+
+Apache HBase is an open source datastore using Google BigTable's datamodel, but running on top of the Hadoop Distributed File System (HDFS) instead of the Google File System (GFS).
+\\Since its launch, HBase has adopted several secondary indexing mechanisms.
+HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for Map/Reduce jobs.
+\\The strong points of HBase are its strong consistency, which is rare among NoSQL-stores, and its concurrency model: ACID-compliant single row transactions and optimistic concurrency control using multiversioning for wider scope transactions\cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.
 

### 2014-11-13 NoSQL_tabel/NoSQL_table.tex
VoltDB sneuvelversie

@@
 \multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
-& Multi-master                                                                        & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at thesame time.
+& Multi-master                                                                        & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
 \\ \midrule
@@
 & By default not, but possible{[}8{]} & Consistent hashing                                                        \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong{[}9{]}                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS{[}11{]} Row-level atomicity{[}12{]} & Single-row transactions (ACID), OCC with MVCC for wider scope transactions{[}13{]} & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y                                                                         \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Written in}}                & C++                                                                                                                                                             & Java                                                                                                                     & Java                                                                               &                                                                                     & C++                                 & Java                                                                      \\ \bottomrule
+\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment
+ \\\bottomrule
 \end{tabular}

### 2014-11-13 paper_dec/paper_dec.tex
VoltDB sneuvelversie

@@
 HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for Map/Reduce jobs.
-\\The strong points of HBase are its strong consistency, which is rare among NoSQL-stores, and its concurrency model: ACID-compliant single row transactions and optimistic concurrency control using multiversioning for wider scope transactions\cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.
+\\The strong points of HBase are its strong consistency, which is rare among NoSQL-stores, and its concurrency model: ACID-compliant single row transactions and optimistic multi-version concurrency control for wider scope operations\cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.\\\\
+In conclusion, HBase offers flexible data-modeling and is especially useful when there is already a HDFS dataset and when Map/Reduce compatibility is desired.
+It scales well to very large datasets and offers excellent concurrency control and strong consistency.
+
+\section{NewSQL}
+\subsection{VoltDB}
+
+VoltDB is a relational in-memory distributed database, aiming to offer the guarantees of classical SQL-stores and the performance and scaling of NoSQL systems\cite{stonebraker2013voltdb}.
+%TODO rewrite
+\\VoltDB stores data in the traditional relational model, but replicated and partitioned (using consistent hashing) over several nodes\cite{grolinger2013data}.
+The data is queryable through a (growing) subset of SQL-92 \cite{voltdb2010voltdb}.
+Relying entirely on DRAM makes it expensive to scale to petabyte scale datavolumes, but VoltDB can export data to other, more suited DMBSs such as columnar NoSQL stores.
+\\In terms of concurrency control, VoltDB offers full ACID-transactions which execute simultaneously on all replicas.
+Main memory is divided into chunks and these are statically assigned to individual, single-threaded cores.
+A global controller serializes all multi-node transactions single-node transactions and are inserted into the transactions queues of the respective nodes.
+In this way, VoltDB obviates the need for locking and latching techniques.
+\\\\
+In short, VoltDB provides relational in memory storage for relatively large datasets, with very fast SQL-query capacities and ACID transactions.
+It is thus more suited for computation intensive applications that don't work on exorbitantly large data volumes but require very low latency.
 

### 2014-11-15 paper_dec/paper_dec.tex
Methodology + Impala

@@
 \begin{document}
+
+\section{Methodology}
+
+Because of the enormous choice of NoSQL and NewSQL datastores, reviewing all of them wasn't possible.
+This survey reviews the most popular datastores in a few relevant categories, namely document stores, wide columnar stores and NewSQL stores.
+A selection was then made based on similar criteria as in \cite{grolinger2013data}, following the ranking of DB-Engine Ranking \cite{db_engine_rank} as an indicator of popularity.
+This ranking tries to measure popularity based on a few parameters, such as the number of mentions on Web sites, general interest according to Google Trends, frequency of technical discussions on the Web, number of job offers, and number of professional profiles in which the systems are mentioned.
+The result were the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
+There is already an existing application of MongoDB in the sequencing pipeline, making it even more relevant.
+Lastly, NewSQl query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the lab.
+These 6 systems were then compared on characteristics relevant to high performance computing, such as indexing mechanisms, client interfaces to the data, distribution strategy, concurrency control and consistency models.
+
 \section{Columnar stores}
@@
 
+\subsection{Cloudera Impala}
+
+Cloudera Impala is a distributed SQL-engine running on top of the Hadoop stack, either on HDFS or HBase \cite{cloudera_impala}.
+It is intended specifically for analytical use, focussing on delivering real-time querying capacities and not on high-throughput on writes.
+\\Data is accessible through a form of SQL-92.
+\\Impala has a highly efficient I/O-layer keeping disk- and CPU-utilization high at all times, resulting in considerably faster performance than other SQL-on-Hadoop solutions such as Apache Hive \cite{floratou2014sql}.
+However, this comes with the drawback that the working set of a query has to fit in the aggregate physical memory of the cluster it runs on.
+This puts some restrictions on the size of datasets processible with Impala, not fully using the scaling capacity of the underlying Hadoop layer.
+\\Because of its analytical purposes, Impala doesn't offer extensive concurrency control features, relying on the datalayer underneath.
+With HBase's excellent concurrency control features, this doesn't necessarily pose a great problem.
+
 \bibliography{biblio}{}

### 2014-11-16 paper_dec/paper_dec.tex
update gitignore
Sneuvelversie MongoDB af

@@
 These 6 systems were then compared on characteristics relevant to high performance computing, such as indexing mechanisms, client interfaces to the data, distribution strategy, concurrency control and consistency models.
+
+\section{Document stores}
+
+\subsection{MongoDB}
+
+MongoDB stores data in .BSON (binary JSON) documents.
+It has powerful indexing support, with the ability to define secondary indexes of a wide array of types on all attributes, much like in the traditional relational model \cite{mongodb_indexes}.
+To aid with denormalization, the documents can contain embedded documents and arrays, thus obviating the need for joins in the query language.
+\\MongoDB offers API's in many languages and the functionality to define the equivalent of SQL WHERE-clauses as javascript expressions.
+These are then translated to MongoDB's proprietary internal querying language \cite{grolinger2013data}.
+The MongoDB query optimizer processes queries and chooses the most efficient query plan for a query given the available indexes.
+These plans are cached if there are multiple viable options, and can be reevaluated as the data evolves \cite{mongodb_query_plans}.
+\\The consistency of MongoDB is configurable, strong consistency can be attained in two ways: setting the connection to read-only from the master node (which has the most up-to-date version of the data), or forcing a write to succeed only after all replicas have acknowledged it.
+The former degrades the scaling ability of read requests, the later the latency of write requests \cite{grolinger2013data}.
+For concurrency control, MongoDB offers single-document atomicity and implements reader-writer locks. Having to lock on writes severely impacts performance in write-intensive scenarios.
+\\\\As a wrap-up, MongoDB stores BSON-files accessible through many API's with flexible querying and indexing techniques, but its concurrency and consistency model have some drawbacks.
+
+\subsection{CouchBase Server}
 

### 2014-11-16 paper_dec/paper_dec.tex
Couchbase sneuvelversie

@@
 
-MongoDB stores data in .BSON (binary JSON) documents.
+MongoDB stores data in BSON (binary JSON) documents.
 It has powerful indexing support, with the ability to define secondary indexes of a wide array of types on all attributes, much like in the traditional relational model \cite{mongodb_indexes}.
@@
 \\The consistency of MongoDB is configurable, strong consistency can be attained in two ways: setting the connection to read-only from the master node (which has the most up-to-date version of the data), or forcing a write to succeed only after all replicas have acknowledged it.
-The former degrades the scaling ability of read requests, the later the latency of write requests \cite{grolinger2013data}.
+The former degrades the scaling ability of read requests, the latter the latency of write requests \cite{grolinger2013data}.
 For concurrency control, MongoDB offers single-document atomicity and implements reader-writer locks. Having to lock on writes severely impacts performance in write-intensive scenarios.
@@
 \subsection{CouchBase Server}
+
+CouchBase, result of the merger of CouchDB and Membase, stores data in JSON documents.
+It uses the memcached protocol for distributed caching and is intended for highly interactive applications with low-latency requirements \cite{grolinger2013data}\cite{couchbase_about}.
+\\The JSON documents can be nested and are queryable through a SQL-like language, N1QL (note: the latest version, released March 2014, was still a developer preview)\cite{couchbase_n1ql}.
+\\Within one cluster, transactions are strongly consistent, but between multiple clusters only eventually consistent.
+\\Couchbase lets clients choose between optimistic (using compare-and-swap) and pessimistic (using 'finegrained locking') concurrency control.
+\\\\With its flexible data modelling, caching and concurrency control, Couchbase is a good fit for applications requiring fast and intensive interaction between client and data.
 

### 2014-11-17 NoSQL_tabel/NoSQL_table.tex
Conclusion

@@
                                                 & \textbf{MongoDB}                                                                                                                                                         & \textbf{Cassandra}                                                                                                                & \textbf{HBase}                                                                              & \textbf{CouchBase Server}                                                                    & \textbf{Cloudera Impala}                     & \textbf{VoltDB}                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Type}}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & See BigTable                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory NewSQL                                                          \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Type}}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & Wide columnar store                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory relational NewSQL                                                          \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.

### 2014-11-17 paper_dec/paper_dec.tex
Conclusion

@@
 \usepackage{url}
+%\usepackage{lscape}
+
+\usepackage{graphicx}
+\usepackage{pdflscape}
+\usepackage{rotating}
+%\usepackage[a4paper,margin=.5in,landscape]{geometry}
+\usepackage{booktabs}
+\usepackage{array}
 
 \begin{document}
+
+\title{A survey on NoSQL and NewSQL datastores}
+\author{Brecht Gossel\'e}
+\maketitle
 
@@
 
+%\include{tabel}
 \section{Document stores}
@@
 It is intended specifically for analytical use, focussing on delivering real-time querying capacities and not on high-throughput on writes.
-\\Data is accessible through a form of SQL-92.
+\\Data is accessible through a subset of SQL-92.
 \\Impala has a highly efficient I/O-layer keeping disk- and CPU-utilization high at all times, resulting in considerably faster performance than other SQL-on-Hadoop solutions such as Apache Hive \cite{floratou2014sql}.
@@
 
+\section{Conclusions}
+
+In recent years, progress in microbiology and bioinformatics has driven down the cost of DNA-sequencing.
+As DNA of more and more organisms, people,...
+is being sequenced, the question arises to make this process ever more efficient and to store the generated data in a scalable, performant and accessible way.
+This study has focused on NoSQL and NewSQL datastores to be used in both the sequencing pipeline as in the storage and analysis of the results.
+These solutions offer elastic scalability, flexible data modelling, good behavior in a distributed setting and resilience to failures.\\
+Specifically, this paper has reviewed 6 different datastores, 2 of the most popular ones in three categories, and compared them on a selection of properties relevant to HPC applications.
+
 \bibliography{biblio}{}

### 2014-11-18 paper_dec/paper_dec.tex
dink
abstract

@@
-\documentclass{article}
+\documentclass{IEEEtran}
+%\documentclass{article}
 \usepackage{cite}
@@
 
+\section{Abstract}
+
+Advances in bioinformatics have driven down the cost of genome sequencing dramatically over the past years.
+This raises the question how to efficiently and flexibly handle and store the large amount of data this process generates.
+Similar developments in web technology have over the last decade, spawned numerous NoSQL and NewSQL datastores, which offer robust, highly scalable distributed storage and flexible data modelling.
+This paper reviews 6 such datastores and compares them on characteristics relevant to high performance computing in general and bioinformatics more specifically %TODO bioinformatics ding, is dit wel zo?.
+
 \section{Methodology}
 
-Because of the enormous choice of NoSQL and NewSQL datastores, reviewing all of them wasn't possible.
+Because of the enormous choice of NoSQL and NewSQL datastores, an exhaustive study wasn't possible.
 This survey reviews the most popular datastores in a few relevant categories, namely document stores, wide columnar stores and NewSQL stores.

### 2014-11-18 paper_dec/tabel.tex
tabel in paper + TODOs

--- 2014-11-17 NoSQL_tabel/NoSQL_table.tex
@@
-\documentclass{article}
-\usepackage{graphicx}
-\usepackage{pdflscape}
-\usepackage{rotating}
-\usepackage[a4paper,margin=.5in,landscape]{geometry}
-\usepackage{booktabs}
-\usepackage{array}
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
@@
 
-\begin{document}
-%\begin{landscape}
-% Table generated by Excel2LaTeX from sheet 'Sheet1'
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
+\begin{landscape}
 \begin{table}[h]
@@
 \multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y
  \\\bottomrule
@@
 \end{table}
-
-%\end{landscape}
-\end{document}
+\end{landscape}

### 2014-11-18 paper_dec/paper_dec.tex
tabel in paper + TODOs
aanpassinkjes

@@
 Similar developments in web technology have over the last decade, spawned numerous NoSQL and NewSQL datastores, which offer robust, highly scalable distributed storage and flexible data modelling.
-This paper reviews 6 such datastores and compares them on characteristics relevant to high performance computing in general and bioinformatics more specifically %TODO bioinformatics ding, is dit wel zo?.
+This paper reviews 6 such datastores and compares them on characteristics relevant to high performance computing in general and bioinformatics more specifically.
+%TODO bioinformatics ding, is dit wel zo?
+
+\section{Introduction}
+
+%Getalletjes over DNA-sequencing
+%NoSQL datamodellen, CAP, NewSQL fifolosie
+%Tabel?
 
@@
 There is already an existing application of MongoDB in the sequencing pipeline, making it even more relevant.
-Lastly, NewSQl query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the lab.
+Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the lab.
+%TODO eerste vermelding van lab?
 These 6 systems were then compared on characteristics relevant to high performance computing, such as indexing mechanisms, client interfaces to the data, distribution strategy, concurrency control and consistency models.
+
+\include{tabel}
 
@@
 \section{Columnar stores}
+%TODO LSM trees & bloom filters for both
 \subsection{Cassandra}

### 2014-11-18 paper_dec/tabel.tex
aanpassinkjes

@@
 \multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries                                                                                             & Bloom filter{[}5{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
@@
 \multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y

### 2014-11-18 paper_dec/paper_dec.tex
conclusions uitgebreid

@@
-\documentclass{IEEEtran}
-%\documentclass{article}
+%\documentclass{IEEEtran}
+\documentclass{article}
 \usepackage{cite}
@@
 In recent years, progress in microbiology and bioinformatics has driven down the cost of DNA-sequencing.
-As DNA of more and more organisms, people,...
-is being sequenced, the question arises to make this process ever more efficient and to store the generated data in a scalable, performant and accessible way.
+As DNA of more and more organisms is being sequenced, the question arises to make this process ever more efficient and to store the generated data in a scalable, performant and accessible way.
 This study has focused on NoSQL and NewSQL datastores to be used in both the sequencing pipeline as in the storage and analysis of the results.
@@
 Specifically, this paper has reviewed 6 different datastores, 2 of the most popular ones in three categories, and compared them on a selection of properties relevant to HPC applications.
+Columnar stores like Apache Cassandra and HBase are well suited to the task of storing extremely large datasets in a reliable and performant way, and could handle the storage of already sequenced genomes.
+VoltDB and CouchBase Server offer low latency on reads and writes, albeit (especially VoltDB) on smaller datasets, but would thus be very useful for storing intermediate, rapidly changing, data in the sequencing pipeline.
+%TODO quod Impala & MongoDB?
 

### 2014-11-19 paper_dec/paper_dec.tex
Introduction begonnen

@@
 \usepackage{rotating}
-%\usepackage[a4paper,margin=.5in,landscape]{geometry}
 \usepackage{booktabs}
@@
 
-\section{Abstract}
-
+\begin{abstract}
 Advances in bioinformatics have driven down the cost of genome sequencing dramatically over the past years.
@@
 %TODO bioinformatics ding, is dit wel zo?
+\end{abstract}
 
 \section{Introduction}
+
+Scientific progress has caused the cost of genome sequencing to drop at an exponential rate over the past decade and a half, even outpacing Moore's law since 2008 \cite{wetterstrand_sequencing_cost}.
+As in all sorts of biological, medical and pharmaceutical research, more and more genomes are being sequenced, the amount of data generated increases rapidly.
+The intermediary data generated in the sequencing pipeline for a human genome, ranges in the order of ??
+TB.
+Also, a single human genome takes ??
+MB/GB to store when sequenced,%TODO source
+ so as this scales to genomes of millions of people and other organisms, storage requirements become ever more demanding in terms of scalability, latency and concurrency.
+\\Increasing popularity of web services such as social networks has caused a similar explosion of data for web companies.
+To handle this so-called Big Data\cite{mashey1997big} in an adequate way, traditional relational DBMS no longer suffice.
+Therefor, large web companies such as Google and Amazon have developped new storage solutions that meet the demands for high scalability, low latency and high availability\cite{baker2011megastore}.
+This has spawned many so called NoSQL ('Not only SQL') databases, which loosen the rigid relational datamodel in favour of better scaling and easier distribution of the data.
+\\NoSQL datastores come in multiple flavours and can be divided in a few categories based on the datamodel they use:
+\begin{itemize}
+\item Columnar stores: based on the datamodel pioneered by Google's BigTable, these store data in "a sparse, distributed, persistent multidimensional sorted map" \cite{chang2008bigtable}.
+In BigTable's case this is a map of row key, column key and a timestamp.
+In this way, multiple versions of the data can be stored in chronological order.
+Because the system doesn't interpret the data, relationships can't be modeled.
+This is left to application logic\cite{hecht2011nosql}
+\end{itemize}
 

### 2014-11-26 NoSQL_tabel/NoSQL_table.tex
LSM tree & Bloom filters

@@
 \multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary (single field, compound, multikey, geospatial, text, hashed){[}2{]}                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries{[}4{]}                                                                                             & Bloom filter{[}5{]} + No JOIN{[}6{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute, B-tree                                                                  & Y, primary \& secondary, LSM-tree                                                                                           & Y, primary \& secondary, LSM-tree                                                     & Y, primary \& secondary, B-tree                                                      & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries                                                                                             & Bloom filter{[}5{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
@@
 \multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit{[}10{]} Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y
  \\\bottomrule

### 2014-11-26 paper_dec/paper_dec.tex
LSM tree & Bloom filters

@@
 Because the system doesn't interpret the data, relationships can't be modeled.
-This is left to application logic\cite{hecht2011nosql}
+This is left to application logic\cite{hecht2011nosql}.
+\item Document stores: these encapsulate key-value pairs in JSON or JSON-like documents.
 \end{itemize}
 
-%Getalletjes over DNA-sequencing
 %NoSQL datamodellen, CAP, NewSQL fifolosie
@@
 The result were the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
-There is already an existing application of MongoDB in the sequencing pipeline, making it even more relevant.
+There is already an existing application of MongoDB in DNA sequencing pipelines, making it even more relevant \cite{elprep_mongo}.
 Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the lab.
@@
 These 6 systems were then compared on characteristics relevant to high performance computing, such as indexing mechanisms, client interfaces to the data, distribution strategy, concurrency control and consistency models.
-
-\include{tabel}
 
@@
 It now offers tables and composite columns -much like in a conventional schema, and comes with its own query language, CQL\cite{cassandra_then&now}.
-CQL resembles SQL in many ways, but comes with some restrictions.
+CQL resembles SQL in many ways, albeit with some restrictions.
 For instance, it doesn't feature the JOIN clause \cite{cassandra_cql}.
 It strongly encourages physically collocating data that will be queried together, and supports denormalization with features such as collection types.\\
+Cassandra has primary and secondary indexing mechanisms and implements these - like BigTable \cite{chang2008bigtable}, using log-structured merge trees (or LSM trees).
+These allow for deferring updates and flushing them to disk once enough have accumulated, thus reducing disk I/O.
+This greatly benefits write throughput \cite{o1996log}\cite{sears2012blsm}\cite{lakshman2010cassandra}.
+Cassandra, like BigTable, also offers Bloom filters\cite{mullin1983second}.
+These are an efficient probabilistic mechanism to predict whether an item is in a set (in this case, whether a key is in a table) and can thus significantly reduce unnecessary table scans\cite{lakshman2010cassandra}.\\
 In order to scale linearly in the number of nodes to very large datasets, Cassandra operates in a fully masterless fashion.
@@
 \\Since its launch, HBase has adopted several secondary indexing mechanisms.
+For indexing, HBase also uses LSM trees\cite{borthakur2011apache}\cite{sears2012blsm}, and it also offers Bloom filters\cite{hbase_schema}.
 HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for Map/Reduce jobs.

### 2014-11-26 paper_dec/tabel.tex
LSM tree & Bloom filters

@@
 \multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute, B-tree                                                                  & Y, primary \& secondary, LSM-tree                                                                                           & Y, primary \& secondary, LSM-tree                                                     & Y, primary \& secondary, B-tree                                                      & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries                                                                                             & Bloom filter{[}5{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule

### 2014-11-27 paper_dec/paper_dec.tex
introduction gefixed

@@
-%\documentclass{IEEEtran}
-\documentclass{article}
+\documentclass{IEEEtran}
+%\documentclass{article}
 \usepackage{cite}
@@
 This is left to application logic\cite{hecht2011nosql}.
-\item Document stores: these encapsulate key-value pairs in JSON or JSON-like documents.
+\item Document stores: these store data as key-value pairs in documents.
+Values can be of a wide variety of types, such as nested documents, lists or scalar values.
+Attribute names can be dynamically specified at runtime and need not adhere to a fixed schema\cite{cattell2011scalable}.
+This is well suited for modelling complex data structures.
+Many document stores use the JSON-fileformat (or some derived form).
+In contrast to columnar stores, the values in documents are not opaque to the system and can thus be queried and indexed\cite{hecht2011nosql}.\\
+The term NewSQL data stores is being used to classify a set of solutions that aim to combine the horizontal scalability and fault tolerance of NoSQL stores with the relational data model.
+Though they all use the relational model and offer SQL, NewSQL stores vary greatly under the hood, depending on the architecture they are built on\cite{grolinger2013data}.
+This paper proceeds first by elaborating on the methodology used to select and compare the chosen databases and then by reviewing each of the six datastores in detail.
 \end{itemize}

### 2014-11-29 paper_dec/paper_dec.tex
mongodb couchbase index trees

@@
 \documentclass{IEEEtran}
-%\documentclass{article}
 \usepackage{cite}
@@
 MongoDB stores data in BSON (binary JSON) documents.
-It has powerful indexing support, with the ability to define secondary indexes of a wide array of types on all attributes, much like in the traditional relational model \cite{mongodb_indexes}.
+It has powerful indexing support, with the ability to define secondary indexes of a wide array of types on all attributes, much like in the traditional relational model.
+These are implemented using B-trees \cite{mongodb_indexes}.
 To aid with denormalization, the documents can contain embedded documents and arrays, thus obviating the need for joins in the query language.
@@
 \\The JSON documents can be nested and are queryable through a SQL-like language, N1QL (note: the latest version, released March 2014, was still a developer preview)\cite{couchbase_n1ql}.
+Primary and secondary indexes can be defined on the data and are implemented using B-trees\cite{couchbase_index}.
 \\Within one cluster, transactions are strongly consistent, but between multiple clusters only eventually consistent.
@@
 \section{Columnar stores}
-%TODO LSM trees & bloom filters for both
 \subsection{Cassandra}

### 2014-12-02 paper_dec/paper_dec.tex
feedback Roel

@@
 This raises the question how to efficiently and flexibly handle and store the large amount of data this process generates.
-Similar developments in web technology have over the last decade, spawned numerous NoSQL and NewSQL datastores, which offer robust, highly scalable distributed storage and flexible data modelling.
+Similar developments in web technology have spawned numerous NoSQL and NewSQL datastores in the last decade, which offer robust, highly scalable distributed storage and flexible data modelling.
 This paper reviews 6 such datastores and compares them on characteristics relevant to high performance computing in general and bioinformatics more specifically.
@@
 Scientific progress has caused the cost of genome sequencing to drop at an exponential rate over the past decade and a half, even outpacing Moore's law since 2008 \cite{wetterstrand_sequencing_cost}.
-As in all sorts of biological, medical and pharmaceutical research, more and more genomes are being sequenced, the amount of data generated increases rapidly.
-The intermediary data generated in the sequencing pipeline for a human genome, ranges in the order of ??
-TB.
-Also, a single human genome takes ??
-MB/GB to store when sequenced,%TODO source
- so as this scales to genomes of millions of people and other organisms, storage requirements become ever more demanding in terms of scalability, latency and concurrency.
+As in all sorts of biological, medical and pharmaceutical research, more and more genomes are being sequenced the amount of data generated increases rapidly.
+For instance, the whole genome sequencing pipeline of the Broad Institute\cite{broad_institute}, a reference in the field, generates in the order of 3 TB of intermediary data when sequencing a single human genome.
+Also, when sequenced with 50x coverage (resolution) a single human genome takes 50 GB to store in compressed format,
+%TODO source
+so as this scales to genomes of millions of people and other organisms, storage and processing requirements become ever more demanding in terms of scalability, latency and concurrency.
+\\A logical evolution has been to tackle these problems with high performance computing systems.
+The Exascience Life Lab of imec, Intel, Janssen Pharmaceutica and 5 Flemish universities, actively researches the application of supercomputers for processing whole genome sequences \cite{exascience_life_lab}\cite{lifelab_BWA}.
 \\Increasing popularity of web services such as social networks has caused a similar explosion of data for web companies.
 To handle this so-called Big Data\cite{mashey1997big} in an adequate way, traditional relational DBMS no longer suffice.
-Therefor, large web companies such as Google and Amazon have developped new storage solutions that meet the demands for high scalability, low latency and high availability\cite{baker2011megastore}.
+Therefore, large web companies such as Google and Amazon have developped new storage solutions that meet the demands for high and incremental scalability, low latency and high availability\cite{baker2011megastore}.
 This has spawned many so called NoSQL ('Not only SQL') databases, which loosen the rigid relational datamodel in favour of better scaling and easier distribution of the data.
@@
 In contrast to columnar stores, the values in documents are not opaque to the system and can thus be queried and indexed\cite{hecht2011nosql}.\\
-The term NewSQL data stores is being used to classify a set of solutions that aim to combine the horizontal scalability and fault tolerance of NoSQL stores with the relational data model.
+The term NewSQL data stores is being used to classify a set of solutions that aim to combine the scalability, distribution and fault tolerance of NoSQL stores with the relational data model.
 Though they all use the relational model and offer SQL, NewSQL stores vary greatly under the hood, depending on the architecture they are built on\cite{grolinger2013data}.
@@
 
-Because of the enormous choice of NoSQL and NewSQL datastores, an exhaustive study wasn't possible.
+Because of the enormous choice of NoSQL and NewSQL datastores, an exhaustive study wasn't feasible.
 This survey reviews the most popular datastores in a few relevant categories, namely document stores, wide columnar stores and NewSQL stores.
@@
 The result were the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
-There is already an existing application of MongoDB in DNA sequencing pipelines, making it even more relevant \cite{elprep_mongo}.
-Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the lab.
+There is already an existing application of MongoDB in the DNA sequencing pipeline used in the ExaScience Life Lab\cite{elprep_mongo}.
+Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the aforementioned lab.
 %TODO eerste vermelding van lab?
@@
 In order to scale linearly in the number of nodes to very large datasets, Cassandra operates in a fully masterless fashion.
-In terms of the CAP-theorem, it focusses on availability and partition-tolerance, rather than immediate consistency (though the user has control over the level of consistency, as will be explained).
+In terms of the CAP-theorem, it focusses on availability and partition-tolerance, rather than immediate consistency (though the user has control over the level of consistency, as will be explained)\cite{brewer2000towards}.
 High availability and partition tolerance are achieved through asynchronous replication of rows over several nodes in the cluster, using consistent hashing and virtual nodes to handle high churn and incremental addition of nodes \cite{decandia2007dynamo} \cite{lakshman2010cassandra} \cite{cassandra_then&now}.
@@
 In terms of concurrency control, Cassandra offers atomicity for single-row operations and serializable \textit{lightweight transactions}, essentially a compare-and-set functionality for larger operations \cite{cassandra_lightweight_trans}.\\
-Being an open-source project, Cassandra is freely available, but there is an enterprise version with extra features such as integration with Apache Spark and Apache Solr, for complex analytical and search tasks \cite{cassandra_solr} \cite{cassandra_spark}.\\\\
+Being an open-source project, Cassandra is freely available, but there is an enterprise version with extra features such as integration with the data processing engine Apache Spark and the distributed search platform Apache Solr, for complex analytical and search tasks \cite{cassandra_solr} \cite{cassandra_spark} \cite{zaharia2010spark}\cite{apache_solr}.\\\\
 In conclusion, Cassandra offers flexible data-modeling with decent querying and indexing support through its CQL-interface and scales incrementally to vast datasets, thanks to its extensive replication and failure-handling features.
@@
 Main memory is divided into chunks and these are statically assigned to individual, single-threaded cores.
-A global controller serializes all multi-node transactions single-node transactions and are inserted into the transactions queues of the respective nodes.
+A global controller serializes all multi-node transactions to a sequence of single-node transactions and inserts those into the transactions queues of the respective nodes.
 In this way, VoltDB obviates the need for locking and latching techniques.
@@
 In short, VoltDB provides relational in memory storage for relatively large datasets, with very fast SQL-query capacities and ACID transactions.
-It is thus more suited for computation intensive applications that don't work on exorbitantly large data volumes but require very low latency.
+It is thus more suited for computate intensive applications that don't work on exorbitantly large data volumes but require very low latency.
 
@@
 \\Because of its analytical purposes, Impala doesn't offer extensive concurrency control features, relying on the datalayer underneath.
-With HBase's excellent concurrency control features, this doesn't necessarily pose a great problem.
+With HBase's excellent concurrency control features, this doesn't necessarily pose a problem.
 
@@
 Columnar stores like Apache Cassandra and HBase are well suited to the task of storing extremely large datasets in a reliable and performant way, and could handle the storage of already sequenced genomes.
-VoltDB and CouchBase Server offer low latency on reads and writes, albeit (especially VoltDB) on smaller datasets, but would thus be very useful for storing intermediate, rapidly changing, data in the sequencing pipeline.
+VoltDB and CouchBase Server offer low latency on reads and writes, albeit (especially VoltDB) on smaller datasets, but would thus be very useful for storing rapidly changing intermediate data in the sequencing pipeline.
 %TODO quod Impala & MongoDB?

### 2014-12-03 paper_dec/paper_dec.tex
typos
Query optimization, indexing & partitioning in tabel geupdated; couchbase & impala in paper geupdated

@@
 
-%NoSQL datamodellen, CAP, NewSQL fifolosie
 %Tabel?
@@
 Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the aforementioned lab.
-%TODO eerste vermelding van lab?
 These 6 systems were then compared on characteristics relevant to high performance computing, such as indexing mechanisms, client interfaces to the data, distribution strategy, concurrency control and consistency models.
@@
 
-VoltDB is a relational in-memory distributed database, aiming to offer the guarantees of classical SQL-stores and the performance and scaling of NoSQL systems\cite{stonebraker2013voltdb}.
-%TODO rewrite
+VoltDB is a relational in-memory distributed database, aiming to offer the guarantees of classical SQL-stores and the scaling of NoSQL systems at very high speeds\cite{stonebraker2013voltdb}.
 \\VoltDB stores data in the traditional relational model, but replicated and partitioned (using consistent hashing) over several nodes\cite{grolinger2013data}.
@@
 \\Data is accessible through a subset of SQL-92.
+\\Contrary to the options discussed before, Impala doesn't partition rows by default.
+It does offer the possibility and leaves the user free to partition data should the data size call for it\cite{impala_partitioning}.
 \\Impala has a highly efficient I/O-layer keeping disk- and CPU-utilization high at all times, resulting in considerably faster performance than other SQL-on-Hadoop solutions such as Apache Hive \cite{floratou2014sql}.

### 2014-12-03 paper_dec/tabel.tex
Query optimization, indexing & partitioning in tabel geupdated; couchbase & impala in paper geupdated

--- 2014-11-18 paper_dec/tabel.tex
@@
 \begin{landscape}
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 \multicolumn{1}{|l|}{\textbf{Type}}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & Wide columnar store                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory relational NewSQL                                                          \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
+\multicolumn{1}{|l|}{\textbf{Querying}}                &                                                                                                                                            &                                                  &                                        &                                                &                &                                                          \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Master-slave or multi-master, asynchronous replication.
 & Multi-master                                                                        & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
 \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries                                                                                             & Bloom filter{[}5{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
-& By default not, but possible{[}8{]} & Consistent hashing                                                        \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       & Cassandra File System (HDFS compatible)                                                                                      & HDFS                                                                               & Standard Unix \& Windows FS's                                                                                  & HDFS or HBase                       &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute, B-tree                                                                  & Y, primary \& secondary, LSM-tree                                                                                           & Y, primary \& secondary, LSM-tree                                                     & Y, primary \& secondary, B-tree                                                      & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary.
+Hash- \& tree-indexes                             \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries                                                                                             & Bloom filter                                                                                     & Bloom filter                                                                & memcached                                                                           & Partition pruning using partition key &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing                                                                                & Range-partitioning                                                                 & Hashing function                                                 & By default not, but possible & Consistent hashing                                                        \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule

### 2014-12-04 paper_dec/paper_dec.tex
Querying & API in table

@@
 
-%Tabel?
-
 \section{Methodology}
@@
 The result were the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
-There is already an existing application of MongoDB in the DNA sequencing pipeline used in the ExaScience Life Lab\cite{elprep_mongo}.
-Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interested from researchers in the aforementioned lab.
+There already exists an extension of the DNA sequencing pipeline used in the ExaScience Life Lab that allows for using MongoDB databases as input and/or output targets, making MongoDB even more relevant\cite{elprep_mongo}.
+Lastly, NewSQL query engine Cloudera Impala was also taken into consideration because of explicit interest from researchers in the aforementioned lab.
 These 6 systems were then compared on characteristics relevant to high performance computing, such as indexing mechanisms, client interfaces to the data, distribution strategy, concurrency control and consistency models.
@@
 For indexing, HBase also uses LSM trees\cite{borthakur2011apache}\cite{sears2012blsm}, and it also offers Bloom filters\cite{hbase_schema}.
-HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for Map/Reduce jobs.
+HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for MapReduce jobs.
 \\The strong points of HBase are its strong consistency, which is rare among NoSQL-stores, and its concurrency model: ACID-compliant single row transactions and optimistic multi-version concurrency control for wider scope operations\cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.\\\\
-In conclusion, HBase offers flexible data-modeling and is especially useful when there is already a HDFS dataset and when Map/Reduce compatibility is desired.
+In conclusion, HBase offers flexible data-modeling and is especially useful when there is already a HDFS dataset and when MapReduce compatibility is desired.
 It scales well to very large datasets and offers excellent concurrency control and strong consistency.

### 2014-12-05 paper_dec/paper_dec.tex
typos

@@
 For instance, the whole genome sequencing pipeline of the Broad Institute\cite{broad_institute}, a reference in the field, generates in the order of 3 TB of intermediary data when sequencing a single human genome.
-Also, when sequenced with 50x coverage (resolution) a single human genome takes 50 GB to store in compressed format,
+Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read in the sequencing process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format,
 %TODO source
@@
 This is left to application logic\cite{hecht2011nosql}.
-\item Document stores: these store data as key-value pairs in documents.
+\item Document stores: these store data as key-value pairs encapsulated in documents.
 Values can be of a wide variety of types, such as nested documents, lists or scalar values.
@@
 This ranking tries to measure popularity based on a few parameters, such as the number of mentions on Web sites, general interest according to Google Trends, frequency of technical discussions on the Web, number of job offers, and number of professional profiles in which the systems are mentioned.
-The result were the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
+The resulting selection consists of the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
 There already exists an extension of the DNA sequencing pipeline used in the ExaScience Life Lab that allows for using MongoDB databases as input and/or output targets, making MongoDB even more relevant\cite{elprep_mongo}.
@@
 These are then translated to MongoDB's proprietary internal querying language \cite{grolinger2013data}.
-The MongoDB query optimizer processes queries and chooses the most efficient query plan for a query given the available indexes.
-These plans are cached if there are multiple viable options, and can be reevaluated as the data evolves \cite{mongodb_query_plans}.
+The MongoDB query optimizer processes queries and chooses the most efficient plan for a query given the available indexes.
+These plans are cached if there are multiple viable options and can be reevaluated as the data evolves \cite{mongodb_query_plans}.
 \\The consistency of MongoDB is configurable, strong consistency can be attained in two ways: setting the connection to read-only from the master node (which has the most up-to-date version of the data), or forcing a write to succeed only after all replicas have acknowledged it.
@@
 It uses the memcached protocol for distributed caching and is intended for highly interactive applications with low-latency requirements \cite{grolinger2013data}\cite{couchbase_about}.
-\\The JSON documents can be nested and are queryable through a SQL-like language, N1QL (note: the latest version, released March 2014, was still a developer preview)\cite{couchbase_n1ql}.
+\\The JSON documents can be nested and are queryable through a SQL-like language, N1QL (note: the most recent version at the time of writing, released in March 2014, is still a developer preview)\cite{couchbase_n1ql}.
 Primary and secondary indexes can be defined on the data and are implemented using B-trees\cite{couchbase_index}.
@@
 Since its inception Cassandra has however diverged slightly from the BigTable data model.
-It now offers tables and composite columns -much like in a conventional schema, and comes with its own query language, CQL\cite{cassandra_then&now}.
+It now provides tables and composite columns -much like in a conventional schema, and comes with its own query language, CQL\cite{cassandra_then&now}.
 CQL resembles SQL in many ways, albeit with some restrictions.
@@
 Cassandra has primary and secondary indexing mechanisms and implements these - like BigTable \cite{chang2008bigtable}, using log-structured merge trees (or LSM trees).
-These allow for deferring updates and flushing them to disk once enough have accumulated, thus reducing disk I/O.
+These allow for deferring updates and flushing them to disk in batches as soon as enough have accumulated, thus reducing disk I/O.
 This greatly benefits write throughput \cite{o1996log}\cite{sears2012blsm}\cite{lakshman2010cassandra}.
@@
 \\Since its launch, HBase has adopted several secondary indexing mechanisms.
-For indexing, HBase also uses LSM trees\cite{borthakur2011apache}\cite{sears2012blsm}, and it also offers Bloom filters\cite{hbase_schema}.
+For indexing, HBase also uses LSM trees\cite{borthakur2011apache}\cite{sears2012blsm}, and it also provides Bloom filters\cite{hbase_schema}.
 HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for MapReduce jobs.
@@
 In short, VoltDB provides relational in memory storage for relatively large datasets, with very fast SQL-query capacities and ACID transactions.
-It is thus more suited for computate intensive applications that don't work on exorbitantly large data volumes but require very low latency.
+It is thus more suited for compute intensive applications that don't work on exorbitantly large data volumes but require very low latency.
 

### 2014-12-06 NoSQL_table/NoSQL_table.tex
tabel -> table
Querying & API in table
couchbase next to mongo in table

--- 2014-11-18 paper_dec/tabel.tex
@@
+\documentclass{article}
+\usepackage{graphicx}
+\usepackage{pdflscape}
+\usepackage{rotating}
+\usepackage[a4paper,margin=.5in,landscape]{geometry}
+\usepackage{booktabs}
+\usepackage{array}
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
@@
 
-\begin{landscape}
-\begin{table}[h]
+\begin{document}
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[ht]
 \centering
@@
 \cmidrule(l){2-7}
-                                                & \textbf{MongoDB}                                                                                                                                                         & \textbf{Cassandra}                                                                                                                & \textbf{HBase}                                                                              & \textbf{CouchBase Server}                                                                    & \textbf{Cloudera Impala}                     & \textbf{VoltDB}                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Type}}                & Document-store (BSON)                                                                                                                                           & Wide columnar store                                                                                                      & Wide columnar store                                                                       & Document-store (JSON)                                                               & SQL on top of Hadoop                & In-memory relational NewSQL                                                          \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication                                                                                                                          & Masterless, asynchronous replication                                                                                     & Masterslave or multi-master, asynchronous replication.
-& Multi-master                                                                        & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
+                                         & \textbf{MongoDB}       & \textbf{CouchBase Server}    & \textbf{Cassandra}     & \textbf{HBase}     & \textbf{Cloudera Impala}    & \textbf{VoltDB}  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Type}}      & Document-store (BSON)  & Document-store (JSON)        & Wide columnar store    & Wide columnar store   & SQL on top of Hadoop     & In-memory relational NewSQL          \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, JDBC, MapReduce                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication   & Multi-master  & Masterless, asynchronous replication     & Master-slave or multi-master, asynchronous replication.
+& Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
 \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS                                                                                                                                       &                                                                                                                          & HDFS                                                                               &                                                                                     & HDFS or HBase                       &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & primary \& secondary on every attribute                                                                        & Y, primary \& secondary                                                                                                  & Y, primary \& secondary                                                            & Y, primary \& secondary                                                             & Y, primary \& secondary{[}3{]}      & Y, primary \& secondary                                                  \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries                                                                                             & Bloom filter{[}5{]}                                                                                     & Bloom filter{[}7{]}                                                                & memcached                                                                           &                                     &                                                                           \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key                                                                                                                           & Consistent hashing                                                                                & Range-partitioning                                                                 & Hashing function determines bucket.
-& By default not, but possible{[}8{]} & Consistent hashing                                                        \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable                                                                                                                                                    & Configurable                                                                                                             & Strong                                                                      & Within cluster: strong; within multiple clusters: eventual                          &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & CQL offers CAS, row-level atomicity & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations & Application can implement optimistic (using CAS) or pessimistic concurrency control &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS    & Standard Unix \& Windows FS's  & Cassandra File System (HDFS compatible)     & HDFS & HDFS or HBase & \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & Primary \& secondary on every attribute, B-tree    & Primary \& secondary, B-tree  & Primary \& secondary, LSM-tree                                                                                           & Primary \& secondary, LSM-tree                                                                                                         & Primary \& secondary{[}3{]}      & Primary \& secondary.
+Hash- \& tree-indexes                             \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries         & memcached                                                                                         & Bloom filter                                                                                     & Bloom filter                                                                                                                                      & Partition pruning using partition key &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key         & Hashing function                                                                                                                     & Consistent hashing                                                                                & Range-partitioning                                                                                                               & By default not, but possible & Consistent hashing                                                        \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable     & Within cluster: strong; within multiple clusters: eventual                                                                                                                                                       & Configurable                                                                                                             & Strong                                                                                        &                                     & Strong                                                                    \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & Application can implement optimistic (using CAS) or pessimistic concurrency control& CQL offers CAS, row-level atomicity & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations  &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y
@@
 \end{table}
-\end{landscape}
+
+\end{document}

### 2014-12-06 NoSQL_table/NoSQL_table.tex
Mongo replication & storage, hbase api & replication

@@
 \multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, JDBC, MapReduce                                                  \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication   & Multi-master  & Masterless, asynchronous replication     & Master-slave or multi-master, asynchronous replication.
+\multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication   & Multi-master, asynchronous replication  & Masterless, asynchronous replication     & Master-slave or multi-master, asynchronous replication.
 & Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
 \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Storage}}                   & 16MB file limit or GridFS    & Standard Unix \& Windows FS's  & Cassandra File System (HDFS compatible)     & HDFS & HDFS or HBase & \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Storage}}                   & Standard FS's, 16MB document limit    & Standard Unix \& Windows FS's  & Cassandra File System (HDFS compatible)     & HDFS & HDFS or HBase & \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Indexing}}                  & Primary \& secondary on every attribute, B-tree    & Primary \& secondary, B-tree  & Primary \& secondary, LSM-tree                                                                                           & Primary \& secondary, LSM-tree                                                                                                         & Primary \& secondary{[}3{]}      & Primary \& secondary.

### 2014-12-06 paper_dec/paper_dec.tex
Mongo replication & storage, hbase api & replication

@@
 Scientific progress has caused the cost of genome sequencing to drop at an exponential rate over the past decade and a half, even outpacing Moore's law since 2008 \cite{wetterstrand_sequencing_cost}.
-As in all sorts of biological, medical and pharmaceutical research, more and more genomes are being sequenced the amount of data generated increases rapidly.
+Because in all sorts of biological, medical and pharmaceutical research, more and more genomes are being sequenced, the amount of generated data increases rapidly.
 For instance, the whole genome sequencing pipeline of the Broad Institute\cite{broad_institute}, a reference in the field, generates in the order of 3 TB of intermediary data when sequencing a single human genome.
-Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read in the sequencing process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format,
+Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read during the process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format,
 %TODO source
@@
 
-%\include{tabel}
 \section{Document stores}
@@
 To aid with denormalization, the documents can contain embedded documents and arrays, thus obviating the need for joins in the query language.
+\\The document size is limited to 16 MB, to help ensure that a single document cannot take up excessive amounts of RAM or bandwidth.
+To store and retrieve larger files, the built-in tool GridFS (which is not an actual file system) can split up files in smaller chunks and store these chunks as separate documents\cite{mongodb_gridfs}.
 \\MongoDB offers API's in many languages and the functionality to define the equivalent of SQL WHERE-clauses as javascript expressions.
@@
 These plans are cached if there are multiple viable options and can be reevaluated as the data evolves \cite{mongodb_query_plans}.
-\\The consistency of MongoDB is configurable, strong consistency can be attained in two ways: setting the connection to read-only from the master node (which has the most up-to-date version of the data), or forcing a write to succeed only after all replicas have acknowledged it.
+\\The consistency of MongoDB is configurable.
+Strong consistency can be attained in two ways: setting the connection to read-only from the master node (which has the most up-to-date version of the data), or forcing a write to succeed only after all replicas have acknowledged it.
 The former degrades the scaling ability of read requests, the latter the latency of write requests \cite{grolinger2013data}.
+\\Data is replicated asynchronously using range-partitioning: nodes are responsible for ranges of keys.
+This speeds up range queries, but can create hotspots and load-balancing issues.
+To route updates to the right replicas, MongoDB operates in a master-slave setting.\\
 For concurrency control, MongoDB offers single-document atomicity and implements reader-writer locks. Having to lock on writes severely impacts performance in write-intensive scenarios.
@@
 For indexing, HBase also uses LSM trees\cite{borthakur2011apache}\cite{sears2012blsm}, and it also provides Bloom filters\cite{hbase_schema}.
-HBase doesn't offer an SQL-like advanced querying language, but because of its HDFS underpinnings it can function as both input and output for MapReduce jobs.
+HBase comes with a Java API but without an SQL-like advanced querying language, though a workaround via Apache Hive, another data warehousing and querying project\cite{apache_hive}, and its query language HiveQL is possible.
+Because of its HDFS underpinnings it can function as both input and output for MapReduce jobs.
+\\HBase partitions data in ranges like BigTable and replicates updates in master-slave or multi-master fashion.
+Read-requests are not distributed however, as rows are only serviced by one server.
+The replicas are solely intended for failure recovery.
 \\The strong points of HBase are its strong consistency, which is rare among NoSQL-stores, and its concurrency model: ACID-compliant single row transactions and optimistic multi-version concurrency control for wider scope operations\cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.\\\\

### 2014-12-06 paper_dec/NoSQL_table.tex
Impala components + table in paper (decently\!)

--- 2014-12-06 NoSQL_table/NoSQL_table.tex
@@
 \begin{document}
+\pagenumbering{gobble}
 % Table generated by Excel2LaTeX from sheet 'Sheet1'
@@
 \multicolumn{1}{|l|}{\textbf{Type}}      & Document-store (BSON)  & Document-store (JSON)        & Wide columnar store    & Wide columnar store   & SQL on top of Hadoop     & In-memory relational NewSQL          \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, JDBC, MapReduce                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, rich API, MapReduce                                                  \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication   & Multi-master, asynchronous replication  & Masterless, asynchronous replication     & Master-slave or multi-master, asynchronous replication.
-& Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
+& Masterless              & Masterless, updates executed on all replicas at the same time.
 \\ \midrule

### 2014-12-06 paper_dec/paper_dec.tex
Impala components + table in paper (decently\!)

@@
 \usepackage{url}
-%\usepackage{lscape}
-
+\usepackage{lscape}
+\usepackage[final]{pdfpages}
 \usepackage{graphicx}
@@
 
+\begin{landscape}
+\includepdf[angle=90]{NoSQL_table.pdf}
+\end{landscape}
 \section{Document stores}
@@
 \\Data is accessible through a subset of SQL-92.
+\\Impala's architecture is almost perfectly symmetrically distributed: all nodes execute the same \texttt{impalad}-process, which is responsible for the main database functionality.
+Queries can be sent to any node which will then function as coordinator for the specific query.
+There are however two processes that run on only one (not necessarily the same) node in the cluster, performing bookkeeping tasks and relaying metadata changes through the cluster\cite{impala_components}.
 \\Contrary to the options discussed before, Impala doesn't partition rows by default.

### 2014-12-07 paper_dec/NoSQL_table.tex
less offering, more style

@@
 \multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable     & Within cluster: strong; within multiple clusters: eventual                                                                                                                                                       & Configurable                                                                                                             & Strong                                                                                        &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & Application can implement optimistic (using CAS) or pessimistic concurrency control& CQL offers CAS, row-level atomicity & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations  &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & Application can implement optimistic (using CAS) or pessimistic concurrency control& row-level atomicity, CAS & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations  &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y

### 2014-12-07 paper_dec/paper_dec.tex
less offering, more style

@@
 The term NewSQL data stores is being used to classify a set of solutions that aim to combine the scalability, distribution and fault tolerance of NoSQL stores with the relational data model.
-Though they all use the relational model and offer SQL, NewSQL stores vary greatly under the hood, depending on the architecture they are built on\cite{grolinger2013data}.
+Though they all use the relational model and supply SQL-querying capacities, NewSQL stores vary greatly under the hood, depending on the architecture they are built on\cite{grolinger2013data}.
 This paper proceeds first by elaborating on the methodology used to select and compare the chosen databases and then by reviewing each of the six datastores in detail.
@@
 A selection was then made based on similar criteria as in \cite{grolinger2013data}, following the ranking of DB-Engine Ranking \cite{db_engine_rank} as an indicator of popularity.
-This ranking tries to measure popularity based on a few parameters, such as the number of mentions on Web sites, general interest according to Google Trends, frequency of technical discussions on the Web, number of job offers, and number of professional profiles in which the systems are mentioned.
+This ranking tries to measure popularity based on a few parameters, such as the number of mentions on Web sites, general interest according to Google Trends, frequency of technical discussions on the Web, number of job listings, and number of professional profiles in which the systems are mentioned.
 The resulting selection consists of the document stores MongoDB and CouchBase Server, wide columnar stores Cassandra and HBase and NewSQL database VoltDB.
@@
 Although Cassandra was built with eventual consistency in mind, strong consistency can be obtained by choosing the quorum larger than the number of replicas \cite{grolinger2013data}.\\
-In terms of concurrency control, Cassandra offers atomicity for single-row operations and serializable \textit{lightweight transactions}, essentially a compare-and-set functionality for larger operations \cite{cassandra_lightweight_trans}.\\
+In terms of concurrency control, Cassandra supports atomicity for single-row operations and serializable \textit{lightweight transactions}, essentially a compare-and-set functionality for larger operations \cite{cassandra_lightweight_trans}.\\
 Being an open-source project, Cassandra is freely available, but there is an enterprise version with extra features such as integration with the data processing engine Apache Spark and the distributed search platform Apache Solr, for complex analytical and search tasks \cite{cassandra_solr} \cite{cassandra_spark} \cite{zaharia2010spark}\cite{apache_solr}.\\\\
@@
 For indexing, HBase also uses LSM trees\cite{borthakur2011apache}\cite{sears2012blsm}, and it also provides Bloom filters\cite{hbase_schema}.
-HBase comes with a Java API but without an SQL-like advanced querying language, though a workaround via Apache Hive, another data warehousing and querying project\cite{apache_hive}, and its query language HiveQL is possible.
-Because of its HDFS underpinnings it can function as both input and output for MapReduce jobs.
+HBase comes with a Java API but without an SQL-like advanced querying language, though a workaround via Apache Hive, another data warehousing and analytics project\cite{apache_hive}, and its query language HiveQL is possible.
+Because of its HDFS underpinnings it can easily function as both input and output for MapReduce jobs.
 \\HBase partitions data in ranges like BigTable and replicates updates in master-slave or multi-master fashion.
 Read-requests are not distributed however, as rows are only serviced by one server.
-The replicas are solely intended for failure recovery.
+The replicas are intended solely for failure recovery.
 \\The strong points of HBase are its strong consistency, which is rare among NoSQL-stores, and its concurrency model: ACID-compliant single row transactions and optimistic multi-version concurrency control for wider scope operations\cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.\\\\
-In conclusion, HBase offers flexible data-modeling and is especially useful when there is already a HDFS dataset and when MapReduce compatibility is desired.
-It scales well to very large datasets and offers excellent concurrency control and strong consistency.
+In conclusion, HBase allows users to flexibly model data and is especially useful when there is already a HDFS dataset and when MapReduce compatibility is a priority.
+It scales well to very large datasets and features excellent concurrency control and strong consistency.
 
@@
 
-VoltDB is a relational in-memory distributed database, aiming to offer the guarantees of classical SQL-stores and the scaling of NoSQL systems at very high speeds\cite{stonebraker2013voltdb}.
+VoltDB is a relational in-memory distributed database, aiming to couple the guarantees of classical SQL-stores with the scaling of NoSQL systems while performing at very high speeds\cite{stonebraker2013voltdb}.
 \\VoltDB stores data in the traditional relational model, but replicated and partitioned (using consistent hashing) over several nodes\cite{grolinger2013data}.
@@
 Relying entirely on DRAM makes it expensive to scale to petabyte scale datavolumes, but VoltDB can export data to other, more suited DMBSs such as columnar NoSQL stores.
-\\In terms of concurrency control, VoltDB offers full ACID-transactions which execute simultaneously on all replicas.
+\\In terms of concurrency control, VoltDB supports full ACID-transactions which execute simultaneously on all replicas.
 Main memory is divided into chunks and these are statically assigned to individual, single-threaded cores.
@@
 \\Contrary to the options discussed before, Impala doesn't partition rows by default.
-It does offer the possibility and leaves the user free to partition data should the data size call for it\cite{impala_partitioning}.
+It does give the user the possibility to partition data should the data size call for it\cite{impala_partitioning}.
 \\Impala has a highly efficient I/O-layer keeping disk- and CPU-utilization high at all times, resulting in considerably faster performance than other SQL-on-Hadoop solutions such as Apache Hive \cite{floratou2014sql}.
@@
 This puts some restrictions on the size of datasets processible with Impala, not fully using the scaling capacity of the underlying Hadoop layer.
-\\Because of its analytical purposes, Impala doesn't offer extensive concurrency control features, relying on the datalayer underneath.
+\\Because of its analytical purposes, Impala doesn't have extensive concurrency control features, relying instead on the underlying storage layer.
 With HBase's excellent concurrency control features, this doesn't necessarily pose a problem.

### 2014-12-09 paper_dec/NoSQL_table.tex
filled out table, voltdb stored procedures + query plans

@@
 \multicolumn{1}{|l|}{\textbf{Type}}      & Document-store (BSON)  & Document-store (JSON)        & Wide columnar store    & Wide columnar store   & SQL on top of Hadoop     & In-memory relational NewSQL          \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, rich API, MapReduce                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, Java stored procedures, rich API, MapReduce                                                  \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication   & Multi-master, asynchronous replication  & Masterless, asynchronous replication     & Master-slave or multi-master, asynchronous replication.
@@
 \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Storage}}                   & Standard FS's, 16MB document limit    & Standard Unix \& Windows FS's  & Cassandra File System (HDFS compatible)     & HDFS & HDFS or HBase & \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & Primary \& secondary on every attribute, B-tree    & Primary \& secondary, B-tree  & Primary \& secondary, LSM-tree                                                                                           & Primary \& secondary, LSM-tree                                                                                                         & Primary \& secondary{[}3{]}      & Primary \& secondary.
+\multicolumn{1}{|l|}{\textbf{Storage}}                   & Standard FS's, 16MB document limit    & Standard Unix \& Windows FS's  & Cassandra File System (HDFS compatible)     & HDFS & HDFS or HBase & Main memory\\ \midrule
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & Primary \& secondary on every attribute, B-tree    & Primary \& secondary, B-tree  & Primary \& secondary, LSM-tree                                                                                           & Primary \& secondary, LSM-tree                                                                                                         & Primary \& secondary      & Primary \& secondary.
 Hash- \& tree-indexes                             \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries         & memcached                                                                                         & Bloom filter                                                                                     & Bloom filter                                                                                                                                      & Partition pruning using partition key &                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries         & memcached                                                                                         & Bloom filter                                                                                     & Bloom filter                                                                                                                                      & Partition pruning using partition key & Queries in stored procedures planned at compile time                                                                           \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key         & Hashing function                                                                                                                     & Consistent hashing                                                                                & Range-partitioning                                                                                                               & By default not, but possible & Consistent hashing                                                        \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable     & Within cluster: strong; within multiple clusters: eventual                                                                                                                                                       & Configurable                                                                                                             & Strong                                                                                        &                                     & Strong                                                                    \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & Application can implement optimistic (using CAS) or pessimistic concurrency control& row-level atomicity, CAS & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations  &                                     & ACID + data access serialized and executed in single-threaded environment \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Consistency}}               & Configurable     & Within cluster: strong; within multiple clusters: eventual                                                                                                                                                       & Configurable                                                                                                             & Strong                                                                                        & Relies on underlying storage layer                                    & Strong                                                                    \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Concurrency control}} & Atomic single document operations, otherwise 2-phase commit; Concurrent reads, exclusive writes (lock on DB level) & Application can implement optimistic (using CAS) or pessimistic concurrency control& Row-level atomicity, CAS & Single-row transactions (ACID possible), OCC with MVCC for wider scope operations  &                                     Relies on underlying storage layer & ACID + data access serialized and executed in single-threaded environment \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Open-source?}}              & Y                                                                                                                                                               & Y                                                                                                                        & Y                                                                                  & Y                                                                                   & Y                                   & Y
@@
 \end{tabular}
+\caption{An overview of the compared datastores and their features}
 \end{table}

### 2014-12-09 paper_dec/paper_dec.tex
VoltDB indexes
filled out table, voltdb stored procedures + query plans

@@
 This paper reviews 6 such datastores and compares them on characteristics relevant to high performance computing in general and bioinformatics more specifically.
-%TODO bioinformatics ding, is dit wel zo?
 \end{abstract}
@@
 For instance, the whole genome sequencing pipeline of the Broad Institute\cite{broad_institute}, a reference in the field, generates in the order of 3 TB of intermediary data when sequencing a single human genome.
-Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read during the process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format,
-%TODO source
-so as this scales to genomes of millions of people and other organisms, storage and processing requirements become ever more demanding in terms of scalability, latency and concurrency.
+Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read during the process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format\footnote{This information was gathered through personal interaction with researchers at the Exascience Life Lab}, so as this scales to genomes of millions of people and other organisms, storage and processing requirements become ever more demanding in terms of scalability, latency and concurrency.
 \\A logical evolution has been to tackle these problems with high performance computing systems.
-The Exascience Life Lab of imec, Intel, Janssen Pharmaceutica and 5 Flemish universities, actively researches the application of supercomputers for processing whole genome sequences \cite{exascience_life_lab}\cite{lifelab_BWA}.
+The Exascience Life Lab of imec, Intel, Janssen Pharmaceutica and 5 Flemish universities, actively researches the application of supercomputers for accelerating the processing of whole genome sequences \cite{exascience_life_lab}\cite{lifelab_BWA}.
 \\Increasing popularity of web services such as social networks has caused a similar explosion of data for web companies.
@@
 The data is queryable through a (growing) subset of SQL-92 \cite{voltdb2010voltdb}.
+Queries are preferably defined as stored procedures written in Java, in which the SQL statements are embedded.
+VoltDB supports primary and secondary indexing, and leaves the user the choice between hash- and tree-indexes\cite{voltdb_indexes}.
+VoltDB plans and optimizes queries in stored procedures optimized at compile time\cite{voltdb_query_plans}.\\
 Relying entirely on DRAM makes it expensive to scale to petabyte scale datavolumes, but VoltDB can export data to other, more suited DMBSs such as columnar NoSQL stores.

### 2014-12-09 paper_dec/paper_dec.tex
added Mongo & Impala to conclusion

@@
 Queries are preferably defined as stored procedures written in Java, in which the SQL statements are embedded.
-VoltDB supports primary and secondary indexing, and leaves the user the choice between hash- and tree-indexes\cite{voltdb_indexes}.
+VoltDB supports primary and secondary indexing, and gives the user the choice between hash- and tree-indexes\cite{voltdb_indexes}.
 VoltDB plans and optimizes queries in stored procedures optimized at compile time\cite{voltdb_query_plans}.\\
@@
 However, this comes with the drawback that the working set of a query has to fit in the aggregate physical memory of the cluster it runs on.
-This puts some restrictions on the size of datasets processible with Impala, not fully using the scaling capacity of the underlying Hadoop layer.
+This puts some restrictions on the size of datasets processable with Impala, not fully using the scaling capacity of the underlying Hadoop layer.
 \\Because of its analytical purposes, Impala doesn't have extensive concurrency control features, relying instead on the underlying storage layer.
@@
 Columnar stores like Apache Cassandra and HBase are well suited to the task of storing extremely large datasets in a reliable and performant way, and could handle the storage of already sequenced genomes.
+MongoDB sets itself apart with its extensive API, flexible data modeling, querying and indexing features.
+It doesn't have quite the write-handling capacities of the columnar stores, but is nevertheless a strong candidate for storing very large datasets such as the sequenced genomes.
 VoltDB and CouchBase Server offer low latency on reads and writes, albeit (especially VoltDB) on smaller datasets, but would thus be very useful for storing rapidly changing intermediate data in the sequencing pipeline.
-%TODO quod Impala & MongoDB?
+Lastly, Cloudera Impala lends itself well to situations where fast read, but not write queries on large datasets are required.
 

### 2014-12-10 paper_dec/paper_dec.tex
key-value & graph DBs in introduction & methodology

@@
 \begin{itemize}
+\item Key-value stores: much like dictionaries and associative maps, these map unique keys to values.
+Values are uninterpreted byte arrays, and the only way to access is by their key.
+This means modeling relations and complex structures between data, rich querying and indexing is not possible\cite{hecht2011nosql}\cite{grolinger2013data}.
 \item Columnar stores: based on the datamodel pioneered by Google's BigTable, these store data in "a sparse, distributed, persistent multidimensional sorted map" \cite{chang2008bigtable}.
@@
 Many document stores use the JSON-fileformat (or some derived form).
-In contrast to columnar stores, the values in documents are not opaque to the system and can thus be queried and indexed\cite{hecht2011nosql}.\\
+In contrast to columnar stores, the values in documents are not opaque to the system and can thus be queried and indexed\cite{hecht2011nosql}.
+\item Graph databases: as the name suggests, these originate from graph theory and use graphs as their data model.
+They are especially useful to manage highly interconnected data coming from sources such as social networks or location based services, replacing costly operations like recursive joins with efficient graph traversals\cite{hecht2011nosql}.
+\end{itemize}
 The term NewSQL data stores is being used to classify a set of solutions that aim to combine the scalability, distribution and fault tolerance of NoSQL stores with the relational data model.
@@
 This paper proceeds first by elaborating on the methodology used to select and compare the chosen databases and then by reviewing each of the six datastores in detail.
-\end{itemize}
+
+\begin{landscape}
+\includepdf[angle=90]{NoSQL_table.pdf}
+\end{landscape}
 
@@
 This survey reviews the most popular datastores in a few relevant categories, namely document stores, wide columnar stores and NewSQL stores.
+Key-value stores and graph databases were not taken into consideration, as their respective data models are not suited for the application at hand.
 A selection was then made based on similar criteria as in \cite{grolinger2013data}, following the ranking of DB-Engine Ranking \cite{db_engine_rank} as an indicator of popularity.
@@
 
-\begin{landscape}
-\includepdf[angle=90]{NoSQL_table.pdf}
-\end{landscape}
 \section{Document stores}

### 2014-12-13 paper_dec/NoSQL_table.tex
moved table up a bit. Handing in.

@@
 Hash- \& tree-indexes                             \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer + Shard-keys to speed up distributed queries         & memcached                                                                                         & Bloom filter                                                                                     & Bloom filter                                                                                                                                      & Partition pruning using partition key & Queries in stored procedures planned at compile time                                                                           \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Query optimization}}        & Query optimizer, shard-keys to speed up distributed queries         & memcached                                                                                         & Bloom filter                                                                                     & Bloom filter                                                                                                                                      & Partition pruning using partition key & Queries in stored procedures planned at compile time                                                                           \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Partitioning}}              & Range partitioning based on shard key         & Hashing function                                                                                                                     & Consistent hashing                                                                                & Range-partitioning                                                                                                               & By default not, but possible & Consistent hashing                                                        \\ \midrule

### 2014-12-13 paper_dec/nosql_newsql_survey.tex
moved table up a bit. Handing in.

--- 2014-12-10 paper_dec/paper_dec.tex
@@
 For instance, the whole genome sequencing pipeline of the Broad Institute\cite{broad_institute}, a reference in the field, generates in the order of 3 TB of intermediary data when sequencing a single human genome.
-Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read during the process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format\footnote{This information was gathered through personal interaction with researchers at the Exascience Life Lab}, so as this scales to genomes of millions of people and other organisms, storage and processing requirements become ever more demanding in terms of scalability, latency and concurrency.
+Also, when sequenced with 50x mean coverage (denoting the average number of times a base has been read during the process\cite{coverage_definition}) a single human genome takes 50 GB to store in compressed format\footnote{This information was gathered during conversations with researchers at the Exascience Life Lab}, so as this scales to genomes of millions of people and other organisms, storage and processing requirements become ever more demanding in terms of scalability, latency and concurrency.
 \\A logical evolution has been to tackle these problems with high performance computing systems.
@@
 \item Key-value stores: much like dictionaries and associative maps, these map unique keys to values.
-Values are uninterpreted byte arrays, and the only way to access is by their key.
+Values are uninterpreted byte arrays, and the only way to access them is by their key.
 This means modeling relations and complex structures between data, rich querying and indexing is not possible\cite{hecht2011nosql}\cite{grolinger2013data}.
@@
 
-\begin{landscape}
-\includepdf[angle=90]{NoSQL_table.pdf}
-\end{landscape}
-
 \section{Methodology}
@@
 A selection was then made based on similar criteria as in \cite{grolinger2013data}, following the ranking of DB-Engine Ranking \cite{db_engine_rank} as an indicator of popularity.
+\begin{landscape}
+\includepdf[angle=90]{NoSQL_table.pdf}
+\end{landscape}
 This ranking tries to measure popularity based on a few parameters, such as the number of mentions on Web sites, general interest according to Google Trends, frequency of technical discussions on the Web, number of job listings, and number of professional profiles in which the systems are mentioned.

### 2015-01-05 paper_dec/nosql_newsql_survey.tex
hyphenation

@@
 This has spawned many so called NoSQL ('Not only SQL') databases, which loosen the rigid relational datamodel in favour of better scaling and easier distribution of the data.
-\\NoSQL datastores come in multiple flavours and can be divided in a few categories based on the datamodel they use:
+\\NoSQL datastores come in multiple flavours and can be divided in a few cate\-gories based on the datamodel they use:
 \begin{itemize}
@@
 Because of the enormous choice of NoSQL and NewSQL datastores, an exhaustive study wasn't feasible.
-This survey reviews the most popular datastores in a few relevant categories, namely document stores, wide columnar stores and NewSQL stores.
+This survey reviews the most popular datastores in a few relevant cate\-gories, namely document stores, wide columnar stores and NewSQL stores.
 Key-value stores and graph databases were not taken into consideration, as their respective data models are not suited for the application at hand.

### 2015-01-28 db_schemes/gemini_schemas.tex
gemini schema layouts comparison

\documentclass{article}
\usepackage{booktabs}
\usepackage{color}

\setlength\parindent{0pt}

\begin{document}

\section{Nu}

1 \texttt{variants} tabel met array columns.\\
\texttt{SELECT * FROM variants WHERE (gts.brecht == hom\_ref \&\& gts.aaron == hom\_alt)}
\begin{table}[h]
\begin{tabular}{@{}|l|l|l|l|l|l|l|@{}}
\toprule
variant\_id & ref & alt & \ldots & gts & gt\_types & \ldots \\ \bottomrule
\end{tabular}
\end{table}

\section{Alternatieven}

\subsection{Super-\texttt{variants}-tabel}

1 \texttt{variants} tabel zonder array columns, met kolom voor elke \texttt{individual}.\\
\begin{table}[!htbp]
\begin{tabular}{@{}|l|l|l|l|l|l|l|@{}}
\toprule
variant\_id & ref & alt & \ldots & aaron & brecht & \ldots \\ \bottomrule
\end{tabular}
\end{table}
\texttt{SELECT * FROM variants WHERE brecht == hom\_ref \&\& aaron == hom\_alt}

\subsection{\texttt{genotypes} tabel}

1 \texttt{variants} tabel zonder array columns

\begin{table}[!htbp]
\begin{tabular}{@{}|l|l|l|l|@{}}
\toprule
variant\_id & ref & alt & \ldots \\ \bottomrule
\end{tabular}
\end{table}

1 \texttt{genotypes} tabel met rij voor elke \texttt{(variant, sample, genotype)}

\begin{table}[!htbp]
\begin{tabular}{@{}|l|l|l|l|l|@{}}
\toprule
variant\_id & sample & genotype & \color{red} individual\_id & \ldots \\ \bottomrule
\end{tabular}
\end{table}

Query onmogelijk zonder \texttt{individual\_id}. Anders:\\
\texttt{SELECT * FROM ((SELECT variant\_id FROM (variants JOIN (SELECT * FROM genotypes WHERE individual\_id == brecht)) WHERE hom\_ref) JOIN (SELECT * FROM genotypes WHERE individual\_id == aaron)) WHERE hom\_alt}

\subsection{\texttt{individual} tabellen}

1 \texttt{variants} tabel zonder array columns.

\begin{table}[!htbp]
\begin{tabular}{@{}|l|l|l|l|@{}}
\toprule
variant\_id & ref & alt & \ldots \\ \bottomrule
\end{tabular}
\end{table}

1 \texttt{individual} tabel per individu met rij voor elke \texttt{variant}

\begin{table}[!htbp]
\begin{tabular}{@{}|l|l|l|l|@{}}
\toprule
variant\_id & sample & genotype & \ldots \\ \bottomrule
\end{tabular}
\end{table}

\texttt{SELECT * FROM aaron JOIN (SELECT variant\_id FROM (brecht JOIN variants) WHERE hom\_ref) WHERE hom\_alt }

\end{document}

### 2015-01-30 db_schemas/gemini_schemas.tex
tabbing in gemini query examples

--- 2015-01-28 db_schemes/gemini_schemas.tex
@@
 1 \texttt{variants} tabel met array columns.\\
-\texttt{SELECT * FROM variants WHERE (gts.brecht == hom\_ref \&\& gts.aaron == hom\_alt)}
+
+\texttt{SELECT * \\FROM variants \\WHERE (gts.brecht == hom\_ref \&\& gts.aaron == hom\_alt)}
 \begin{table}[h]
@@
 1 \texttt{variants} tabel zonder array columns, met kolom voor elke \texttt{individual}.\\
+
+\texttt{SELECT * \\
+FROM variants \\
+WHERE brecht == hom\_ref \&\& aaron == hom\_alt}
 \begin{table}[!htbp]
@@
 \end{table}
-\texttt{SELECT * FROM variants WHERE brecht == hom\_ref \&\& aaron == hom\_alt}
 
+\subsection{\texttt{individual} tabellen}
+
+1 \texttt{variants} tabel zonder array columns.
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|l|l|@{}}
+\toprule
+variant\_id & ref & alt & \ldots \\ \bottomrule
+\end{tabular}
+\end{table}
+
+1 \texttt{individual} tabel per individu met rij voor elke \texttt{variant}
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|l|l|@{}}
+\toprule
+variant\_id & sample & genotype & \ldots \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{tabbing}
+\texttt{SELECT *}
+\\\texttt{FROM aaron JOIN (}\=\texttt{SELECT variant\_id}
+\\ \> \texttt{FROM (brecht JOIN variants)}
+\\ \> \texttt{WHERE hom\_ref) }
+\\ \texttt{WHERE hom\_alt }
+\end{tabbing}
+
+\newpage
 \subsection{\texttt{genotypes} tabel}
@@
 Query onmogelijk zonder \texttt{individual\_id}. Anders:\\
-\texttt{SELECT * FROM ((SELECT variant\_id FROM (variants JOIN (SELECT * FROM genotypes WHERE individual\_id == brecht)) WHERE hom\_ref) JOIN (SELECT * FROM genotypes WHERE individual\_id == aaron)) WHERE hom\_alt}
 
-\subsection{\texttt{individual} tabellen}
+\begin{tabbing}
+\texttt{SELECT *} 
+\\\texttt{FROM ((}\=\texttt{SELECT variant\_id }
+\\ \> \texttt{FROM (variants JOIN (}\=\texttt{SELECT * }
+\\ \>\>\texttt{FROM genotypes }
+\\ \>\>\texttt{WHERE individual\_id == brecht)) }
+\\ \> \texttt{WHERE hom\_ref) JOIN (}\=\= \texttt{SELECT * }
+					\\ \>\>\> \texttt{FROM genotypes }
+					\\ \>\>\> \texttt{WHERE individual\_id == aaron))}
+\\ \texttt{WHERE hom\_alt}
 
-1 \texttt{variants} tabel zonder array columns.
-
-\begin{table}[!htbp]
-\begin{tabular}{@{}|l|l|l|l|@{}}
-\toprule
-variant\_id & ref & alt & \ldots \\ \bottomrule
-\end{tabular}
-\end{table}
-
-1 \texttt{individual} tabel per individu met rij voor elke \texttt{variant}
-
-\begin{table}[!htbp]
-\begin{tabular}{@{}|l|l|l|l|@{}}
-\toprule
-variant\_id & sample & genotype & \ldots \\ \bottomrule
-\end{tabular}
-\end{table}
-
-\texttt{SELECT * FROM aaron JOIN (SELECT variant\_id FROM (brecht JOIN variants) WHERE hom\_ref) WHERE hom\_alt }
+\end{tabbing}
 

### 2015-02-19 probleemstelling/NoSQL_table.tex
probleemstelling

--- 2014-12-06 NoSQL_table/NoSQL_table.tex
@@
 \multicolumn{1}{|l|}{\textbf{Type}}      & Document-store (BSON)  & Document-store (JSON)        & Wide columnar store    & Wide columnar store   & SQL on top of Hadoop     & In-memory relational NewSQL          \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, JDBC, MapReduce                                                  \\ \midrule
+\multicolumn{1}{|l|}{\textbf{Querying \& API}} & Proprietary language, dynamic queries with JS, rich API, MapReduce    & N1QL, memcached API, MapReduce          & CQL, rich API, MapReduce            & No query language (Hive via workaround), Java API, MapReduce                                                   & SQL-92, MapReduce               & SQL-92, rich API, MapReduce                                                  \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Distributed}}               & Master-slave, asynchronous replication   & Multi-master, asynchronous replication  & Masterless, asynchronous replication     & Master-slave or multi-master, asynchronous replication.
-& Y, not 100\% symmetric              & Masterless, updates executed on all replicas at the same time.
+& Masterless              & Masterless, updates executed on all replicas at the same time.
 \\ \midrule
 \multicolumn{1}{|l|}{\textbf{Storage}}                   & Standard FS's, 16MB document limit    & Standard Unix \& Windows FS's  & Cassandra File System (HDFS compatible)     & HDFS & HDFS or HBase & \\ \midrule
-\multicolumn{1}{|l|}{\textbf{Indexing}}                  & Primary \& secondary on every attribute, B-tree    & Primary \& secondary, B-tree  & Primary \& secondary, LSM-tree                                                                                           & Primary \& secondary, LSM-tree                                                                                                         & Primary \& secondary{[}3{]}      & Primary \& secondary.
+\multicolumn{1}{|l|}{\textbf{Indexing}}                  & Primary \& secondary on every attribute, B-tree    & Primary \& secondary, B-tree  & Primary \& secondary, LSM-tree                                                                                           & Primary \& secondary, LSM-tree                                                                                                         & Primary \& secondary      & Primary \& secondary.
 Hash- \& tree-indexes                             \\ \midrule

### 2015-02-19 probleemstelling/rapport.tex
probleemstelling

\documentclass{article}
\usepackage{cite}
\usepackage{url}
\usepackage{lscape}
\usepackage[final]{pdfpages}
\usepackage{graphicx}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{array}
\usepackage[dutch]{babel}

\begin{document}

\title{Optimized genome sequencing: onderzoeksprobleem}

\author{Brecht Gossel\'e}
\maketitle

\section{Probleemschets}

\subsection{Algemeen}

Wetenschappelijke vooruitgang heeft ervoor gezorgd dat de kost om genomen te sequencen het afgelopen decennium exponentieel gedaald is, sinds 2008 zelfs aan een hogere snelheid dan de evolutie volgens de wet van Moore \cite{wetterstrand_sequencing_cost}. In allerhande soorten biologisch, medisch en pharmaceutisch onderzoek worden dan ook genomen van meer en meer organismen gesequenced en dit genereert enorme hoeveelheden data. Ter illustratie: de \textit{whole genome sequencing pipeline} van het Broad Institute \cite{broad_institute}, een referentie in het veld, genereert bij het sequencen van 1 volledig menselijk genoom in de orde van 3TB aan tussentijdse data. Het eindresultaat is 50 GB gecomprimeerde data voor 1 menselijk genoom bij 50x \textit{coverage} (een maat voor de resolutie \cite{coverage_definition}). Naarmate genomen van miljoenen mensen en andere levende wezens geanalyseerd en opgeslagen worden, vereist deze evolutie steeds betere schaalbaarheid, responstijd, en parallellisering voor de opslag en verwerking van deze data.\\
Een logische stap is om deze problemen aan te pakken met grote verdeelde computersystemen, zogenaamde \textit{high performance computing systems} of \textit{supercomputers}. Het Exascience Life Lab van imec, Intel, Janssen Pharmaceutica en de 5 Vlaamse universiteiten verricht specifiek onderzoek naar de toepassing van supercomputers om het genoomsequencingproces te versnellen \cite{lifelab_bwa}\cite{exascience_life_lab}.\\
De snel toegenomen populariteit van webdiensten als sociale netwerken zadelde webbedrijven op met een gelijkaardige explosie aan data. Om deze zogenaamde Big Data \cite{mashey1997big} adequaat te beheren, volstaan traditionele relationele DBMS niet meer. Daarom hebben grote webbedrijven zoals Google, Amazon en Facebook nieuwe opslagtechnieken ontwikkeld die voldoen aan de vereisten qua incrementele schaalbaarheid, lage responstijden en hoge beschikbaarheid \cite{baker2011megastore}. Dit heeft vele zogenaamde NoSQL ('Not only SQL') databases voortgebracht, die het rigide relationele datamodel inruilen voor betere schaalbaarheid en gemakkelijkere distributie van de data. Daarnaast is er ook de recentere opkomst van NewSQL-systemen: deze trachten de schaalbaarheid, distributie en fouttolerantie van NoSQL-systemen te combineren met het relationele datamodel en de bijhorende SQL-query interface en sterke garanties op gebied van concurrency en consistentie.

\subsection{GEMINI}

GEMINI (van GEnome MINIng) is een tool voor flexibele analyse van genetische variaties in grote datasets van genomen, ontwikkeld aan de University of Virginia \cite{10.1371/journal.pcbi.1003153}. GEMINI vertrekt daarbij van VCF (Variant Call Format) bestanden, laadt deze in in een database, en biedt dan de mogelijkheid queries uit te voeren op deze databank. De huidige operationele versie van GEMINI draait op de eenvoudige SQL-databank SQLite, maar deze laat op gebied van schaalbaarheid en performantie te wensen over. Een eerste oplossing om de performantie te verhogen was om over te schakelen op PostgreSQL, en hiervan is intussen al met succes een eerste versie ge\"implementeerd. Het resultaat is een verhoging van de querysnelheden van 5 tot 20x, zonder verregaande optimalisaties. De verwachtingen zijn echter ook hier dat PostgreSQL op langere termijn niet zal kunnen schalen naar datasets van (honderd)duizenden genomen.\\
GEMINI bewaart de data over genetische varianten in enkele zeer grote tabellen. Een eerste vereiste voor een database is dus om deze tabulaire data goed te kunnen voorstellen en beheren, bij voorkeur d.m.v. automatische verspreiding over verschillende nodes in een cluster. Bovendien voert GEMINI na het inladen van de VCF bestanden enkel nog read-queries uit, dus zijn de belangrijkste verdere vereisten voor een database hoge read-throughput, goede query-mogelijkheden en indexeringsmechanismes. Omdat GEMINI in Python ge\"implementeerd is, is een Python-API ook nuttig.

\section{Doelstellingen}

Het doel van mijn masterproef is te onderzoeken hoe en welke NoSQL- of NewSQL-systemen kunnen helpen om de genoomsequencing pipeline te versnellen. Daarbij komt bovendien kijken dat verschillende delen van de pipeline verschillende eigenschappen en dus uiteenlopende vereisten hebben.\\
Een eerste stap in mijn onderzoek is dan ook een literatuurstudie om het probleemdomein te leren kennen, verschillende NoSQL- en NewSQL-systemen met elkaar te vergelijken en in te schatten welke het meest geschikt zouden zijn voor welk facet van het sequencingproces.\\
De volgende stap is om met de opgedane kennis te besluiten welke data store het meest geschikt is om GEMINI te verbeteren, dit te implementeren en door testen te verifiëren of de performantie van GEMINI in combinatie met een NoSQL- of NewSQL-systeem aan de verwachtingen voldoet, dit te vergelijken met de bestaande implementaties en te onderzoeken vanaf welke grootte van dataset het de moeite loont om van PostgreSQL over te schakelen op een NoSQL- of NewSQL-oplossing.

\section{Aanpak}

\subsection{Literatuurstudie}

De literatuurstudie heb ik intussen afgerond; de resultaten hiervan heb ik beschreven in een eerste wetenschappelijke paper \cite{gossele_survey}. Ik heb enkele van de populairste datastores in 3 relevante categorieën bestudeerd: de NoSQL document stores MongoDB en Couchbase Server, de NoSQL columnar stores Cassandra en HBase en de NewSQL-systemen Cloudera Impala en VoltDB. Als maat voor de populariteit heb ik de website DB-Engine Ranking \cite{db_engine_rank} gevolgd, die de populariteit van databases inschat op basis van een aantal parameters zoals aantal vermeldingen op het web, zoekfrequentie in verschillende zoekmachines en jobaanbiedingen die de databases vermelden. Deze 6 systemen heb ik vervolgens vergeleken op een aantal eigenschappen die van toepassing zijn voor high performance computing, zoals indexeringsmechanismes, client interfaces naar de data, distributiestrategie, concurrency controle en consistentiemodellen. De resultaten van de vergelijkende studie zijn samengevat in de onderaan bijgevoegde tabel.

\subsection{Implementatie}

De eerste prioriteit in de implementatie is kiezen welke van de bestudeerde databases het meest geschikt is voor GEMINI. Een voor de hand liggende keuze is Cassandra: dit is een matuur product dat ervoor bekend staat te schalen tot enorme datasets, biedt de nodige query- en indexeringsmechanismes, een Python API en bovendien mapt de huidige data layout van GEMINI goed op het columnaire model van Cassandra.\\
De volgende stap is om dit daadwerkelijk te implementeren: uitdokteren hoe de data uit GEMINI exact in Cassandra voor te stellen en vervolgens de connectie tussen GEMINI en Cassandra te programmeren. Dit is vooral het inladen van de data uit VCF files, en het omzetten van queries uit de GEMINI command line interface naar geldige queries in de query taal van Cassandra (CQL of Cassandra Query Language, een sterk op SQL lijkende taal). GEMINI heeft reeds een uitgebreide suite unit-tests die de implementatie zullen vergemakkelijken.
Eenmaal de implementatie werkt en een Cassandra cluster met een testdataset aan genoom informatie operationeel is, kan de combinatie GEMINI-Cassandra gebenchmarkt worden door de prestaties op een aantal veelgebruikte queries te meten.\\\\
Als er tijd overblijft, kan ik nog een tweede database uitproberen en implementeren. Couchbase Server is een andere interessante optie: deze staat dankzij zijn memcached-protocol bekend om z'n lage latency, biedt ook query- en indexing-mechanismes aan (maar iets minder uitgebreid dan Cassandra), heeft eveneens een Python API en het datamodel van GEMINI mapt vrij goed op de JSON-documenten architectuur van Couchbase.

\begin{landscape}
\includepdf[angle=90]{NoSQL_table.pdf}
\end{landscape}

\bibliography{biblio}{}
\bibliographystyle{plain}
\end{document}

### 2015-02-19 gemini/gemini/GeminiQuery.py
gemini src

@@ -1,0 +1,1247 @@
+#!/usr/bin/env python
+
+import os
+import sys
+import sqlite3
+import re
+import itertools
+import collections
+import json
+import abc
+import re
+import numpy as np
+
+# gemini imports
+import gemini_utils as util
+from gemini_constants import *
+from gemini_utils import (OrderedSet, OrderedDict, itersubclasses, partition,
+                          partition_by_fn)
+import compression
+from sql_utils import ensure_columns, get_select_cols_and_rest
+from gemini_subjects import get_subjects
+
+class RowFormat:
+    """A row formatter to output rows in a custom format.  To provide
+    a new output format 'foo', implement the class methods and set the
+    name field to foo.  This will automatically add support for 'foo' to
+    anything accepting the --format option via --format foo.
+    """
+
+    __metaclass__ = abc.ABCMeta
+
+    @abc.abstractproperty
+    def name(self):
+        return
+
+    @abc.abstractmethod
+    def format(self, row):
+        """ return a string representation of a GeminiRow object
+        """
+        return '\t'.join([str(row.row[c]) for c in row.row])
+
+    @abc.abstractmethod
+    def format_query(self, query):
+        """ augment the query with columns necessary for the format or else just
+        return the untouched query
+        """
+        return query
+
+    @abc.abstractmethod
+    def predicate(self, row):
+        """ the row must pass this additional predicate to be output. Just
+        return True if there is no additional predicate"""
+        return True
+
+    @abc.abstractmethod
+    def header(self, fields):
+        """ return a header for the row """
+        return "\t".join(fields)
+
+
+class DefaultRowFormat(RowFormat):
+
+    name = "default"
+
+    def __init__(self, args):
+        pass
+
+    def format(self, row):
+        return '\t'.join([str(row.row[c]) for c in row.row])
+
+    def format_query(self, query):
+        return query
+
+    def predicate(self, row):
+        return True
+
+    def header(self, fields):
+        """ return a header for the row """
+        return "\t".join(fields)
+
+class CarrierSummary(RowFormat):
+    """
+    Generates a count of the carrier/noncarrier status of each feature in a given
+    column of the sample table
+
+    Assumes None == unknown.
+    """
+    name = "carrier_summary"
+
+    def __init__(self, args):
+        subjects = get_subjects(args)
[…]

### 2015-02-19 gemini/gemini/annotations.py

@@ -1,0 +1,941 @@
+#!/usr/bin/env python
+import pysam
+import sqlite3
+import os
+import sys
+import collections
+import re
+from unidecode import unidecode
+from bx.bbi.bigwig_file import BigWigFile
+from gemini.config import read_gemini_config
+
+# dictionary of anno_type -> open Tabix file handles
+annos = {}
+
+def get_anno_files( args ):
+    config = read_gemini_config( args = args )
+    anno_dirname = config["annotation_dir"]
+    # Default annotations -- always found
+    annos = {
+     'pfam_domain': os.path.join(anno_dirname, 'hg19.pfam.ucscgenes.bed.gz'),
+     'cytoband': os.path.join(anno_dirname, 'hg19.cytoband.bed.gz'),
+     'dbsnp': os.path.join(anno_dirname, 'dbsnp.b141.20140813.hg19.vcf.gz'),
+     'clinvar': os.path.join(anno_dirname, 'clinvar_20140807.vcf.gz'),
+     'gwas': os.path.join(anno_dirname, 'hg19.gwas.bed.gz'),
+     'rmsk': os.path.join(anno_dirname, 'hg19.rmsk.bed.gz'),
+     'segdup': os.path.join(anno_dirname, 'hg19.segdup.bed.gz'),
+     'conserved': os.path.join(anno_dirname, '29way_pi_lods_elements_12mers.chr_specific.fdr_0.1_with_scores.txt.hg19.merged.bed.gz'),
+     'cpg_island': os.path.join(anno_dirname, 'hg19.CpG.bed.gz'),
+     'dgv': os.path.join(anno_dirname, 'hg19.dgv.bed.gz'),
+     'esp': os.path.join(anno_dirname,
+                         'ESP6500SI.all.snps_indels.vcf.gz'),
+     '1000g': os.path.join(anno_dirname,
+                           'ALL.autosomes.phase3_shapeit2_mvncall_integrated_v5.20130502.sites.vcf.gz'),
+     'recomb': os.path.join(anno_dirname,
+                            'genetic_map_HapMapII_GRCh37.gz'),
+     'gms': os.path.join(anno_dirname,
+                         'GRCh37-gms-mappability.vcf.gz'),
+     'grc': os.path.join(anno_dirname, 'GRC_patch_regions.bed.gz'),
+     'cse': os.path.join(anno_dirname, "cse-hiseq-8_4-2013-02-20.bed.gz"),
+     'encode_tfbs': os.path.join(anno_dirname,
[…]

### 2015-02-19 gemini/gemini/gemini_annotate.py

@@ -1,0 +1,390 @@
+    #!/usr/bin/env python
+
+import os
+import sys
+import sqlite3
+from collections import defaultdict, namedtuple
+import json
+import subprocess
+
+import numpy as np
+from scipy.stats import mode
+import pysam
+
+from gemini.annotations import annotations_in_region, guess_contig_naming
+from database import database_transaction
+
+def add_requested_columns(args, update_cursor, col_names, col_types=None):
+    """
+    Attempt to add new, user-defined columns to the
+    variants table.  Warn if the column already exists.
+    """
+
+    if args.anno_type in ["count", "boolean"]:
+
+        col_name = col_names[0]
+        col_type = "integer"
+
+        try:
+            alter_qry = "ALTER TABLE variants ADD COLUMN " \
+                        + col_name \
+                        + " " \
+                        + col_type \
+                        + " " \
+                        + "DEFAULT NULL"
+            update_cursor.execute(alter_qry)
+        except sqlite3.OperationalError:
+            sys.stderr.write("WARNING: Column \"("
+                             + col_name
+                             + ")\" already exists in variants table. Overwriting values.\n")
+    elif args.anno_type == "extract":
+
+        for col_name, col_type in zip(col_names, col_types):
+
+            try:
+                alter_qry = "ALTER TABLE variants ADD COLUMN " \
+                            + col_name \
+                            + " " \
+                            + col_type \
+                            + " " \
+                            + "DEFAULT NULL"
+                update_cursor.execute(alter_qry)
+            except sqlite3.OperationalError:
+                sys.stderr.write("WARNING: Column \"("
+                                 + col_name
+                                 + ")\" already exists in variants table. Overwriting values.\n")
+    else:
+        sys.exit("Unknown annotation type: %s\n" % args.anno_type)
+
+
+def _annotate_variants(args, conn, get_val_fn, col_names=None, col_types=None, col_ops=None):
+    """Generalized annotation of variants with a new column.
+
+    get_val_fn takes a list of annotations in a region and returns
+    the value for that region to update the database with.
+
+    Separates selection and identification of values from update,
+    to avoid concurrent database access errors from sqlite3, especially on
+    NFS systems. The retained to_update list is small, but batching
+    could help if memory issues emerge.
+    """
+    # For each, use Tabix to detect overlaps with the user-defined
[…]

### 2015-02-19 gemini/gemini/gemini_load.py

@@ -1,0 +1,422 @@
+#!/usr/bin/env python
+
+# native Python imports
+import os.path
+import shutil
+import sys
+import sqlite3
+
+import annotations
+import subprocess
+from cluster_helper.cluster import cluster_view
+import database as gemini_db
+from gemini_load_chunk import GeminiLoader
+import gemini_annotate
+import uuid
+import time
+import datetime
+
+
+def load(parser, args):
+    if (args.db is None or args.vcf is None):
+        parser.print_help()
+        exit("ERROR: load needs both a VCF file and a database file\n")
+
+    annos = annotations.get_anno_files( args )
+    # force skipping CADD and GERP if the data files have not been installed
+    if args.skip_cadd is False:
+        if 'cadd_score' not in annos:
+            sys.stderr.write("\nCADD scores are not being loaded because the"
+            " annotation file could not be found.\n"
+            "`Run gemini update --dataonly --extra cadd_score`"
+            " to install the annotation file.\n\n")
+            args.skip_cadd = True
+        else:
+            sys.stderr.write("CADD scores are being loaded (to skip use:--skip-cadd).\n")
+    if args.skip_gerp_bp is False:
+        if 'gerp_bp' not in annos:
+            sys.stderr.write("\nGERP per bp is not being loaded because the annotation file"
+                        " could not be found.\n    Run `gemini update --dataonly --extra gerp_bp`"
+                        " to install the annotation file.\n\n")
+            args.skip_gerp_bp = True
+        else:
+            sys.stderr.write("GERP per bp is being loaded (to skip use:--skip-gerp-bp).\n")
+    # collect of the the add'l annotation files
+    annotations.load_annos( args )
+
+    if args.scheduler:
+        load_ipython(args)
+    elif args.cores > 1:
+        load_multicore(args)
+    else:
+        load_singlecore(args)
+
+def load_singlecore(args):
+    # create a new gemini loader and populate
+    # the gemini db and files from the VCF
+    gemini_loader = GeminiLoader(args)
+    gemini_loader.store_resources()
+    gemini_loader.store_version()
+    gemini_loader.store_vcf_header()
[…]

### 2015-02-19 gemini/gemini/gemini_load_chunk.py

@@ -1,0 +1,690 @@
+#!/usr/bin/env python
+
+# native Python imports
+import os.path
+import sys
+import sqlite3
+import numpy as np
+from itertools import repeat
+import json
+
+# third-party imports
+import cyvcf as vcf
+
+# gemini modules
+import version
+from ped import default_ped_fields, load_ped_file
+import gene_table
+import infotag
+import database
+import annotations
+import func_impact
+import severe_impact
+import popgen
+import structural_variants as svs
+from gemini_constants import *
+from compression import pack_blob
+from gemini.config import read_gemini_config
+
+
+class GeminiLoader(object):
+    """
+    Object for creating and populating a gemini
+    database and auxillary data files.
+    """
+    def __init__(self, args, buffer_size=10000):
+        self.args = args
+
+        # create the gemini database
+        self._create_db()
+        # create a reader for the VCF file
+        self.vcf_reader = self._get_vcf_reader()
+        # load sample information
+
+        if not self.args.no_genotypes and not self.args.no_load_genotypes:
+            # load the sample info from the VCF file.
+            self._prepare_samples()
+            # initialize genotype counts for each sample
+            self._init_sample_gt_counts()
+            self.num_samples = len(self.samples)
+        else:
+            self.num_samples = 0
+
+        self.buffer_size = buffer_size
+        self._get_anno_version()
+        
+        if not args.skip_gene_tables:
+            self._get_gene_detailed()
+            self._get_gene_summary()
+
+        if self.args.anno_type == "VEP":
+            self._effect_fields = self._get_vep_csq(self.vcf_reader)
+        else:
+            self._effect_fields = []
+
+    def store_vcf_header(self):
+        """Store the raw VCF header.
+        """
+        database.insert_vcf_header(self.c, self.vcf_reader.raw_header)
+
+    def store_resources(self):
+        """Create table of annotation resources used in this gemini database.
+        """
+        database.insert_resources(self.c, annotations.get_resources( self.args ))
+
+    def store_version(self):
+        """Create table documenting which gemini version was used for this db.
+        """
+        database.insert_version(self.c, version.__version__)
+
+    def _get_vid(self):
+        if hasattr(self.args, 'offset'):
+            v_id = int(self.args.offset)
+        else:
+            v_id = 1
+        return v_id
+
+    def populate_from_vcf(self):
+        """
+        """
+        import gemini_annotate  # avoid circular dependencies
+        self.v_id = self._get_vid()
+        self.counter = 0
[…]

### 2015-02-19 gemini/gemini/gemini_main.py

@@ -1,0 +1,1143 @@
+#!/usr/bin/env python
+import os.path
+import sys
+import argparse
+import gemini.version
+
+
+def examples(parser, args):
+
+    print
+    print "[load] - load a VCF file into a gemini database:"
+    print "   gemini load -v my.vcf my.db"
+    print "   gemini load -v my.vcf -t snpEff my.db"
+    print "   gemini load -v my.vcf -t VEP my.db"
+    print
+
+    print "[stats] - report basic statistics about your variants:"
+    print "   gemini stats --tstv my.db"
+    print "   gemini stats --tstv-coding my.db"
+    print "   gemini stats --sfs my.db"
+    print "   gemini stats --snp-counts my.db"
+    print
+
+    print "[query] - explore the database with ad hoc queries:"
+    print "   gemini query -q \"select * from variants where is_lof = 1 and aaf <= 0.01\" my.db"
+    print "   gemini query -q \"select chrom, pos, gt_bases.NA12878 from variants\" my.db"
+    print "   gemini query -q \"select chrom, pos, in_omim, clin_sigs from variants\" my.db"
+    print
+
+    print "[dump] - convenient \"data dumps\":"
+    print "   gemini dump --variants my.db"
+    print "   gemini dump --genotypes my.db"
+    print "   gemini dump --samples my.db"
+    print
+
+    print "[region] - access variants in specific genomic regions:"
+    print "   gemini region --reg chr1:100-200 my.db"
+    print "   gemini region --gene TP53 my.db"
+    print
+
+    print "[tools] - there are also many specific tools available"
+    print "   1. Find compound heterozygotes."
+    print "     gemini comp_hets my.db"
+    print
+
+    exit()
+
+def main():
+    #########################################
+    # create the top-level parser
+    #########################################
+    parser = argparse.ArgumentParser(prog='gemini', formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument("-v", "--version", help="Installed gemini version",
+                        action="version",
+                        version="%(prog)s " + str(gemini.version.__version__))
+    parser.add_argument('--annotation-dir', dest='annotation_dir',
+                             help='Path to the annotation database.\n'
+                                'This argument is optional and if given will take precedence over the default location stored in the gemini config file.')
+    subparsers = parser.add_subparsers(title='[sub-commands]', dest='command')
+
+    #########################################
+    # $ gemini examples
+    #########################################
+    parser_examples = subparsers.add_parser('examples',
+                                            help='show usage examples')
[…]

### 2015-02-19 gemini/gemini/gemini_query.py

@@ -1,0 +1,165 @@
+#!/usr/bin/env python
+import os
+import sys
+from itertools import tee, ifilterfalse
+from collections import defaultdict
+
+# gemini imports
+import GeminiQuery
+from GeminiQuery import select_formatter
+from gemini_constants import *
+from gemini_region import add_region_to_query
+from gemini_subjects import (Subject, get_subjects, get_subjects_in_family,
+                             get_family_dict)
+from gemini_utils import itersubclasses
+from dgidb import query_dgidb
+
+def all_samples_predicate(args):
+    """ returns a predicate that returns True if, for a variant,
+    the only samples that have the variant have a given phenotype
+    """
+    subjects = get_subjects(args).values()
+    return select_subjects_predicate(subjects, args)
+
+def family_wise_predicate(args):
+    formatter = select_formatter(args)
+    families = get_family_dict(args)
+    gq = GeminiQuery.GeminiQuery(args.db, out_format=formatter)
+    predicates = []
+    for f in families.values():
+        family_names = [x.name for x in f]
+        subjects = get_subjects_in_family(args, f).values()
+        predicates.append(select_subjects_predicate(subjects, args,
+                                                    family_names))
+    def predicate(row):
+        return sum([p(row) for p in predicates]) >= args.min_kindreds
+    return predicate
+
+def select_subjects_predicate(subjects, args, subset=None):
+    subjects = set([s.name for s in subjects])
+    predicates = []
+    if "all" in args.in_subject:
+        predicates.append(variant_in_all_subjects(subjects))
+    if "none" in args.in_subject:
+        predicates.append(variant_not_in_subjects(subjects))
+    if "only" in args.in_subject:
+        predicates.append(variant_only_in_subjects(subjects, subset))
+    if "any" in args.in_subject:
+        predicates.append(variant_in_any_subject(subjects))
+    if "not" in args.in_subject:
+        def predicate(row):
+            return not all([p(row) for p in predicates])
+    else:
+        def predicate(row):
+            return all([p(row) for p in predicates])
+    return predicate
+
+def variant_in_any_subject(subjects):
+    def predicate(row):
+        return subjects.intersection(samples_with_variant(row)) != set()
+    return predicate
+
+def variant_in_all_subjects(subjects):
+    def predicate(row):
+        return subjects.issubset(samples_with_variant(row))
+    return predicate
+
+def variant_only_in_subjects(subjects, subset=None):
[…]

### 2015-02-19 gemini/gemini/gemini_subjects.py

@@ -1,0 +1,733 @@
+#!/usr/bin/env python
+import sqlite3
+import os
+import sys
+import numpy as np
+from collections import defaultdict
+
+from gemini_constants import *
+import GeminiQuery
+
+class Subject(object):
+
+    """
+    Describe a single subject in the the samples table.
+    """
+    def __init__(self, row):
+        self._set_fields_from_row(row)
+
+    def __repr__(self):
+        return "\t".join(map(str, [self.name, self.paternal_id,
+                                   self.maternal_id, self.phenotype]))
+
+    def set_father(self):
+        self.father = True
+
+    def set_mother(self):
+        self.mother = True
+
+    def _set_fields_from_row(self, row):
+        [setattr(self, k, v) for (k, v) in zip(row.keys(), list(row))]
+        self.phenotype = int(self.phenotype) if self._has_phenotype() else None
+        self._set_affected_status(row)
+
+    def _has_phenotype(self):
+        if hasattr(self, 'phenotype') and self.phenotype is not None:
+            return True
+
+    def _set_affected_status(self, row):
+        # 1 = unaffected
+        # 2 = affected
+        # 0 or -9 is unknown.
+        # http://pngu.mgh.harvard.edu/~purcell/plink/data.shtml#ped
+        if str(self.phenotype) == "2":
+            self.affected = True
+        elif str(self.phenotype) == "1":
+            self.affected = False
+        # distinguish unknown from known to be unaffected.
+        else:
+            self.affected = None
+
+
+class Family(object):
+
+    """
+    Describe the relationships among multiple subjects in a family.
+    """
+    def __init__(self, subjects):
+        self.subjects = subjects
+        self.father = None
+        self.mother = None
+        self.family_id = self.subjects[0].family_id
+        self.children = []
+        self.is_constructed = False
+
+        self.find_parents()
+
+    def has_an_affected(self):
+        """
+        Return True if the Family has at least one affected individual.
+        Otherwise return False.
+        """
+        for subject in self.subjects:
+            if subject.affected:
+                return True
+        return False
+
+    def has_an_affected_child(self):
+        """
+        Return True if the Family has at least one affected child.
+        Otherwise return False.
+        """
+        if not self.is_constructed:
+            self.find_parents()
+
+        for child in self.children:
+            if child.affected:
+                return True
+        return False
+
+    def find_parents(self):
+        """
+        Screen for children with parental ids so that
+        we can identify the parents in this family.
+
[…]

### 2015-02-19 gemini/gemini/scripts/gemini_install.py

@@ -1,0 +1,342 @@
+#!/usr/bin/env python
+"""Installer for gemini: a lightweight db framework for disease and population genetics.
+
+https://github.com/arq5x/gemini
+
+Handles installation of:
+
+- Required third party software
+- Required Python libraries
+- Gemini application
+- Associated data files
+
+Requires: Python 2.7 (or 2.6 and argparse), git, and compilers (gcc, g++)
+
+Run gemini_install.py -h for usage.
+"""
+import argparse
+import platform
+import os
+import shutil
+import subprocess
+import sys
+import urllib2
+
+remotes = {"requirements_pip":
+           "https://raw.github.com/arq5x/gemini/master/requirements.txt",
+            "requirements_conda":
+           "",
+            "versioned_installations":
+            "https://raw.githubusercontent.com/arq5x/gemini/master/versioning/",
+           "cloudbiolinux":
+           "https://github.com/chapmanb/cloudbiolinux.git",
+           "gemini":
+           "https://github.com/arq5x/gemini.git",
+           "anaconda":
+           "http://repo.continuum.io/miniconda/Miniconda-3.7.0-%s-x86_64.sh"}
+
+def main(args):
+    check_dependencies()
+    work_dir = os.path.join(os.getcwd(), "tmpgemini_install")
+    if not os.path.exists(work_dir):
+        os.makedirs(work_dir)
+    os.chdir(work_dir)
+
+    if args.gemini_version != 'latest':
+        requirements_pip = os.path.join(remotes['versioned_installations'],
+                                        args.gemini_version, 'requirements_pip.txt')
+        requirements_conda = os.path.join(remotes['versioned_installations'],
+                                          args.gemini_version, 'requirements_conda.txt')
+        try:
+            urllib2.urlopen(requirements_pip)
+        except:
+            sys.exit('Gemini version %s could not be found. Try the latest version.' % args.gemini_version)
+        remotes.update({'requirements_pip': requirements_pip, 'requirements_conda': requirements_conda})
+
+    print "Installing isolated base python installation"
+    make_dirs(args)
+    anaconda = install_anaconda_python(args, remotes)
+    print "Installing gemini..."
+    install_conda_pkgs(anaconda, remotes, args)
+    gemini = install_gemini(anaconda, remotes, args.datadir, args.tooldir, args.sudo)
[…]

### 2015-02-19 gemini/gemini/version.py

@@ -1,0 +1,1 @@
+__version__="0.11.1a"

### 2015-02-19 gemini/test/test-exac.sh

@@ -1,0 +1,22 @@
+#############################################
+# 1. Test ExAC allele frequencies
+#############################################
+echo "    exac.t01...\c"
+echo "chrom	start	end	ref	alt	in_exac	aaf_exac_all	aaf_adj_exac_all	aaf_adj_exac_afr	aaf_adj_exac_amr	aaf_adj_exac_eas	aaf_adj_exac_fin	aaf_adj_exac_nfe	aaf_adj_exac_oth	aaf_adj_exac_sas
+chr1	985954	985955	G	C	0	None	None	None	None	None	None	None	None	None
+chr1	1199488	1199489	G	A	0	None	None	None	None	None	None	None	None	None
+chr1	1959698	1959699	G	A	1	0.017	0.0175534349801	0.00481877226063	0.00973012667523	0	0.0410552061495	0.0237265929571	0.0232843137255	0.00655049605698
+chr1	161276552	161276553	G	T	0	None	None	None	None	None	None	None	None	None
+chr1	247587092	247587093	C	T	0	None	None	None	None	None	None	None	None	None
+chr9	112185055	112185056	C	G	1	1.626e-05	1.62665105082e-05	0.000189000189	0	0	0	0	0	0
+chr12	121432116	121432118	GC	G	1	0.0002284	0.000234270414993	0.000539374325782	0.000371839365394	0.000825354902608	0	9.89472017731e-05	0	0.000300571085062
+chr14	105420589	105420590	C	T	1	0.001746	0.00175203039036	0.000301083902047	0.000775862068966	0	0.000296384113811	0.0024867521241	0.00437636761488	0.00168431183831" > exp
+
+gemini query --header -q "select chrom, start, end, ref, alt, in_exac, aaf_exac_all, aaf_adj_exac_all, \
+	                        aaf_adj_exac_afr, aaf_adj_exac_amr, aaf_adj_exac_eas, aaf_adj_exac_fin, \
+		                        aaf_adj_exac_nfe, aaf_adj_exac_oth, aaf_adj_exac_sas from variants" test.exac.db > obs
+
+check obs exp
+rm obs exp
+
+#############################################

### 2015-02-23 thesis/thesis.tex
thesis text template dink

\documentclass[master=cws,masteroption=gs]{kulemt}
\setup{title={Master thesis Gemini Cassandra},
		author={Brecht Gossel\'e},
		promotor={Prof. Roel Wuyts},
		assessor={},
		assistant={Prof. Roel Wuyts}}
\setup{filingcard,
		translatedtitle=,
		udc=,
		shortabstract={Here comes a very short abstract, containing no more than 500 words. \LaTeX\ commands can be used here.}}
\begin{document}
\mainmatter
Oi
\end{document}

### 2015-03-16 gemini/gemini/GeminiQuery.py
query expression: NOT, AND, OR

@@ -792,7 +792,7 @@ class GeminiQuery(object):
         and sample names so that the wildcard
         query can be applied to the gt_* columns.
         """
-        query = 'SELECT sample_id, name FROM samples '
+        query = 'SELECT name FROM samples '
         if wildcard.strip() != "*":
             query += ' WHERE ' + wildcard
 

### 2015-03-16 gemini/gemini/database_cassandra.py

@@ -303,20 +303,14 @@ def create_variants_table(session, gt_column_names):
                         sub_type text, \
                         call_rate float, \
                         PRIMARY KEY (sub_type, call_rate, variant_id))''')  
-    
-    session.execute('''CREATE TABLE if not exists variants_by_chrom_depth ( \
-                        variant_id int, \
-                        chrom text, \
-                        depth int, \
-                        PRIMARY KEY (chrom, depth, variant_id))''')
 
 def create_variants_by_samples_tables(session):  
     
-    session.execute('''CREATE TABLE if not exists variants_by_sample_gt_types (\
+    session.execute('''CREATE TABLE if not exists variants_by_samples_gt_type (\
                         sample_name text, gt_type int, variant_id int, \
-                        primary key (sample_name, gt_type, variant_id))''')
+                        primary key ((sample_name, gt_type), variant_id))''')
     
-    session.execute('''CREATE TABLE if not exists variants_by_sample_gt_depths(\
+    session.execute('''CREATE TABLE if not exists variants_by_samples_gt_depth(\
                         sample_name text, gt_depth int, variant_id int, \
                         primary key (sample_name, gt_depth, variant_id))''')         
 

### 2015-03-16 gemini/gemini/gemini_load_chunk.py

@@ -120,7 +120,6 @@ class GeminiLoader(object):
         self.var_buffer = blist([])
         self.var_impacts_buffer = blist([])
         self.var_subtypes_buffer = blist([])
-        self.var_chroms_buffer = blist([])
         buffer_count = 0
         self.skipped = 0
         extra_file, extraheader_file = gemini_annotate.get_extra_files(self.args.db)
@@ -138,7 +137,6 @@ class GeminiLoader(object):
                 # add the core variant info to the variant buffer
                 self.var_buffer.append(variant)
                 self.var_subtypes_buffer.append([variant[4], variant[11], variant[12]])
-                self.var_chroms_buffer.append([variant[4], variant[0], variant[75]])
                 # add each of the impact for this variant (1 per gene/transcript)
                 for var_impact in variant_impacts:
                     self.var_impacts_buffer.append(var_impact)
@@ -154,13 +152,10 @@ class GeminiLoader(object):
                                                       self.var_impacts_buffer)
                     database_cassandra.batch_insert(self.session, 'variants_by_sub_type_call_rate',\
                                                      get_column_names('variants_by_sub_type_call_rate'), self.var_subtypes_buffer)
-                    database_cassandra.batch_insert(self.session, 'variants_by_chrom_depth', \
-                                                    get_column_names('variants_by_chrom_depth'), self.var_chroms_buffer)
                     # binary.genotypes.append(var_buffer)
                     # reset for the next batch
                     self.var_buffer = blist([])
                     self.var_subtypes_buffer = blist([])
-                    self.var_chroms_buffer = blist([])
                     self.var_impacts_buffer = blist([])
                     buffer_count = 0
                 self.v_id += 1
@@ -176,8 +171,6 @@ class GeminiLoader(object):
         database_cassandra.batch_insert(self.session, 'variant_impacts', get_column_names('variant_impacts'), self.var_impacts_buffer)
         database_cassandra.batch_insert(self.session, 'variants_by_sub_type_call_rate',\
                                             get_column_names('variants_by_sub_type_call_rate'), self.var_subtypes_buffer)
-        database_cassandra.batch_insert(self.session, 'variants_by_chrom_depth', \
-                                            get_column_names('variants_by_chrom_depth'), self.var_chroms_buffer)
         sys.stderr.write("pid " + str(os.getpid()) + ": " +
[…]

### 2015-03-16 gemini/gemini/query_expressions.py

@@ -1,0 +1,72 @@
+'''
+Created on Mar 16, 2015
+
+@author: brecht
+'''
+
+class Expression(object):
+    
+    def __init__(self, from_table, select_column, where_columns, where_exps):
+        self.table = from_table
+        self.select_column = select_column
+        self.where_columns = where_columns
+        self.where_exps = where_exps
+        
+    def evaluate(self, session, starting_set):
+        where_clause = " AND ".join(zipwith(lambda x, y: x + y, self.where_columns, self.where_exps))
+        query = "SELECT %s FROM %s WHERE %s" % (self.select_column, self.table, where_clause)
+        
+        if not starting_set == "*":
+            in_clause = ",".join(map(lambda x: str(x), starting_set))            
+            query += " AND %s IN (%s)" % (self.select_column, in_clause)
+            
+        print query
+        rows = session.execute(query)
+        return rows_as_list(rows)
+    
+class AND_expression(object):
+    
+    def __init__(self, left, right):
+        self.left = left
+        self.right = right
+        
+    def evaluate(self, session, starting_set):
+        temp = self.left.evaluate(session, starting_set)
+        return self.right.evaluate(session, temp)
+    
+class OR_expression(object):
+    
+    def __init__(self, left, right):
+        self.left = left
+        self.right = right
+        
+    def evaluate(self, session, starting_set):
+        return union(self.left.evaluate(session, starting_set), self.right.evaluate(session, starting_set))
+    
+class NOT_expression(object):
+    
+    def __init__(self, exp, table, select_column):
+        self.exp = exp
+        self.table = table
+        self.select_column = select_column
+        
+    def evaluate(self, session, starting_set):
+        if starting_set == '*':
+            correct_starting_set = rows_as_list(session.execute("SELECT %s FROM %s" % (self.select_column, self.table)))
+        else:
+            correct_starting_set = starting_set
+            
+        return diff(correct_starting_set, self.exp.evaluate(session, correct_starting_set))
+    
+def zipwith(f, list_a, list_b):
+    return [ f(a,b) for (a,b) in zip(list_a,list_b) ]
+
+def diff(list1, list2):
+    return filter(lambda x: not x in list2, list1)
+
+def union(list1, list2):
+    return set(list1 + list2)
+    
+def rows_as_list(rows):
+    return map(lambda x: x[0], rows)
+    

### 2015-03-16 thesis/cha1.tex

--- 2015-02-19 probleemstelling/rapport.tex
@@
-\documentclass{article}
-\usepackage{cite}
-\usepackage{url}
-\usepackage{lscape}
-\usepackage[final]{pdfpages}
-\usepackage{graphicx}
-\usepackage{pdflscape}
-\usepackage{rotating}
-\usepackage{booktabs}
-\usepackage{array}
-\usepackage[dutch]{babel}
+\chapter{Inleiding}
+\label{cha:1}
 
-\begin{document}
-
-\title{Optimized genome sequencing: onderzoeksprobleem}
-
-\author{Brecht Gossel\'e}
-\maketitle
-
-\section{Probleemschets}
-
-\subsection{Algemeen}
+\section{Probleemstelling}
 
@@
 
-\subsection{GEMINI}
+\section{GEMINI}
 
@@
 Omdat GEMINI in Python ge\"implementeerd is, is een Python-API ook nuttig.
-
-\section{Doelstellingen}
-
-Het doel van mijn masterproef is te onderzoeken hoe en welke NoSQL- of NewSQL-systemen kunnen helpen om de genoomsequencing pipeline te versnellen.
-Daarbij komt bovendien kijken dat verschillende delen van de pipeline verschillende eigenschappen en dus uiteenlopende vereisten hebben.\\
-Een eerste stap in mijn onderzoek is dan ook een literatuurstudie om het probleemdomein te leren kennen, verschillende NoSQL- en NewSQL-systemen met elkaar te vergelijken en in te schatten welke het meest geschikt zouden zijn voor welk facet van het sequencingproces.\\
-De volgende stap is om met de opgedane kennis te besluiten welke data store het meest geschikt is om GEMINI te verbeteren, dit te implementeren en door testen te verifiëren of de performantie van GEMINI in combinatie met een NoSQL- of NewSQL-systeem aan de verwachtingen voldoet, dit te vergelijken met de bestaande implementaties en te onderzoeken vanaf welke grootte van dataset het de moeite loont om van PostgreSQL over te schakelen op een NoSQL- of NewSQL-oplossing.
-
-\section{Aanpak}
-
-\subsection{Literatuurstudie}
-
-De literatuurstudie heb ik intussen afgerond; de resultaten hiervan heb ik beschreven in een eerste wetenschappelijke paper \cite{gossele_survey}.
-Ik heb enkele van de populairste datastores in 3 relevante categorieën bestudeerd: de NoSQL document stores MongoDB en Couchbase Server, de NoSQL columnar stores Cassandra en HBase en de NewSQL-systemen Cloudera Impala en VoltDB.
-Als maat voor de populariteit heb ik de website DB-Engine Ranking \cite{db_engine_rank} gevolgd, die de populariteit van databases inschat op basis van een aantal parameters zoals aantal vermeldingen op het web, zoekfrequentie in verschillende zoekmachines en jobaanbiedingen die de databases vermelden.
-Deze 6 systemen heb ik vervolgens vergeleken op een aantal eigenschappen die van toepassing zijn voor high performance computing, zoals indexeringsmechanismes, client interfaces naar de data, distributiestrategie, concurrency controle en consistentiemodellen.
-De resultaten van de vergelijkende studie zijn samengevat in de onderaan bijgevoegde tabel.
-
-\subsection{Implementatie}
-
-De eerste prioriteit in de implementatie is kiezen welke van de bestudeerde databases het meest geschikt is voor GEMINI.
-Een voor de hand liggende keuze is Cassandra: dit is een matuur product dat ervoor bekend staat te schalen tot enorme datasets, biedt de nodige query- en indexeringsmechanismes, een Python API en bovendien mapt de huidige data layout van GEMINI goed op het columnaire model van Cassandra.\\
-De volgende stap is om dit daadwerkelijk te implementeren: uitdokteren hoe de data uit GEMINI exact in Cassandra voor te stellen en vervolgens de connectie tussen GEMINI en Cassandra te programmeren.
-Dit is vooral het inladen van de data uit VCF files, en het omzetten van queries uit de GEMINI command line interface naar geldige queries in de query taal van Cassandra (CQL of Cassandra Query Language, een sterk op SQL lijkende taal).
-GEMINI heeft reeds een uitgebreide suite unit-tests die de implementatie zullen vergemakkelijken.
-Eenmaal de implementatie werkt en een Cassandra cluster met een testdataset aan genoom informatie operationeel is, kan de combinatie GEMINI-Cassandra gebenchmarkt worden door de prestaties op een aantal veelgebruikte queries te meten.\\\\
-Als er tijd overblijft, kan ik nog een tweede database uitproberen en implementeren.
-Couchbase Server is een andere interessante optie: deze staat dankzij zijn memcached-protocol bekend om z'n lage latency, biedt ook query- en indexing-mechanismes aan (maar iets minder uitgebreid dan Cassandra), heeft eveneens een Python API en het datamodel van GEMINI mapt vrij goed op de JSON-documenten architectuur van Couchbase.
-
-\begin{landscape}
-\includepdf[angle=90]{NoSQL_table.pdf}
-\end{landscape}
-
-\bibliography{biblio}{}
-\bibliographystyle{plain}
-\end{document}

### 2015-03-17 gemini/gemini/GeminiQuery.py
build query exps from gt-filter wildcards

@@ -785,6 +785,7 @@ class GeminiQuery(object):
                                    if not description_tuple[0].startswith("gt")]
             self.report_cols = self.all_query_cols
 
+    #TODO: convert wildcard to query expression on correct samples table(s)	
     def _get_matching_sample_ids(self, wildcard):
         """
         Helper function to convert a sample wildcard
@@ -799,8 +800,7 @@ class GeminiQuery(object):
         sample_info = [] # list of sample_id/name tuples
         self.session.execute(query)
         for row in self.session:
-            # sample_ids are 1-based but gt_* indices are 0-based
-            sample_info.append(str(row['name']))
+            sample_info.append(row[0])
         return sample_info
 
     def _correct_genotype_filter(self):
@@ -897,32 +897,38 @@ class GeminiQuery(object):
 
                 # build the rule based on the wildcard the user has supplied.
                 if wildcard_op == "all":
-                    rule = "("
-                    for idx, sample in enumerate(self.sample_info[token_idx]):
-                        rule += column + '_' + str(sample) + wildcard_rule
-                        if idx < len(self.sample_info[token_idx]) - 1:
-                            rule += ' and '
-                    rule += ")"
+                    rule = Expression('variants_by_samples_' + column, 'variant_id', \
+				'sample_id = ' + self.sample_info[token_idx][0] + \
+				'AND ' + column + wildcard_rule)
+		    for i in range(1, len(self.sample_info[token_idx]):
+			left = Expression('variants_by_samples_' + column, 'variant_id', \
+				'sample_id = ' + self.sample_info[token_idx][i] + \
+				'AND ' + column + wildcard_rule)
+			rule = AND_expression(left, rule)
+				
                 elif wildcard_op == "any":
-                    rule = "("
-                    for idx, sample in enumerate(self.sample_info[token_idx]):
-                        rule += column + '_' + str(sample) + wildcard_rule
-                        if idx < len(self.sample_info[token_idx]) - 1:
-                            rule += ' or '
-                    rule += ")"
+                    rule = Expression('variants_by_samples_' + column, 'variant_id', \
+				'sample_id = ' + self.sample_info[token_idx][0] + \
+				'AND ' + column + wildcard_rule)
+		    for i in range(1, len(self.sample_info[token_idx]):
+			left = Expression('variants_by_samples_' + column, 'variant_id', \
+				'sample_id = ' + self.sample_info[token_idx][i] + \
[…]

### 2015-03-17 gemini/gemini/gemini_load_chunk.py
query exps with one where clause i.s.o. columns and clauses

@@ -14,7 +14,7 @@ import version
 from ped import default_ped_fields, load_ped_file
 import gene_table
 import infotag
-import database_cassandra
+from database_cassandra import insert, batch_insert, create_tables, create_variants_table, create_samples_table
 import annotations
 import func_impact
 import severe_impact
@@ -76,17 +76,17 @@ class GeminiLoader(object):
     def store_vcf_header(self):
         """Store the raw VCF header.
         """
-        database_cassandra.insert(self.session, 'vcf_header', get_column_names('vcf_header'), [self.vcf_reader.raw_header])
+        insert(self.session, 'vcf_header', get_column_names('vcf_header'), [self.vcf_reader.raw_header])
 
     def store_resources(self):
         """Create table of annotation resources used in this gemini database.
         """
-        database_cassandra.batch_insert(self.session, 'resources', get_column_names('resources'), annotations.get_resources( self.args ))
+        batch_insert(self.session, 'resources', get_column_names('resources'), annotations.get_resources( self.args ))
 
     def store_version(self):
         """Create table documenting which gemini version was used for this db.
         """
-        database_cassandra.insert(self.session, 'version', get_column_names('version'), [version.__version__])
+        insert(self.session, 'version', get_column_names('version'), [version.__version__])
 
     def _get_vid(self):
         if hasattr(self.args, 'offset'):
@@ -147,10 +147,10 @@ class GeminiLoader(object):
                     sys.stderr.write("pid " + str(os.getpid()) + ": " +
                                      str(self.counter) + " variants processed.\n")
                     #TODO: column_names cachen?
-                    database_cassandra.batch_insert(self.session, 'variants', get_column_names('variants') + self.gt_column_names, self.var_buffer)
-                    database_cassandra.batch_insert(self.session, 'variant_impacts', get_column_names('variant_impacts'),
+                    batch_insert(self.session, 'variants', get_column_names('variants') + self.gt_column_names, self.var_buffer)
+                    batch_insert(self.session, 'variant_impacts', get_column_names('variant_impacts'),
                                                       self.var_impacts_buffer)
-                    database_cassandra.batch_insert(self.session, 'variants_by_sub_type_call_rate',\
+                    batch_insert(self.session, 'variants_by_sub_type_call_rate',\
[…]

### 2015-03-17 gemini/gemini/query_expressions.py

@@ -6,15 +6,13 @@ Created on Mar 16, 2015
 
 class Expression(object):
     
-    def __init__(self, from_table, select_column, where_columns, where_exps):
+    def __init__(self, from_table, select_column, where_clause):
         self.table = from_table
         self.select_column = select_column
-        self.where_columns = where_columns
-        self.where_exps = where_exps
+        self.where_clause = where_clause
         
     def evaluate(self, session, starting_set):
-        where_clause = " AND ".join(zipwith(lambda x, y: x + y, self.where_columns, self.where_exps))
-        query = "SELECT %s FROM %s WHERE %s" % (self.select_column, self.table, where_clause)
+        query = "SELECT %s FROM %s WHERE %s" % (self.select_column, self.table, self.where_clause)
         
         if not starting_set == "*":
             in_clause = ",".join(map(lambda x: str(x), starting_set))            
@@ -57,9 +55,6 @@ class NOT_expression(object):
             correct_starting_set = starting_set
             
         return diff(correct_starting_set, self.exp.evaluate(session, correct_starting_set))
-    
-def zipwith(f, list_a, list_b):
-    return [ f(a,b) for (a,b) in zip(list_a,list_b) ]
 
 def diff(list1, list2):
     return filter(lambda x: not x in list2, list1)

### 2015-03-17 gemini/gemini/GeminiQuery.py
pretty printing query exps

@@ -13,6 +13,7 @@ import compression
 from gemini_constants import HOM_ALT, HOM_REF, HET, UNKNOWN
 from gemini_subjects import get_subjects
 from gemini_utils import (OrderedSet, OrderedDict, itersubclasses, partition_by_fn)
+from query_expressions import Expression, AND_expression, OR_expression, NOT_expression
 import gemini_utils as util
 import numpy as np
 from sql_utils import ensure_columns, get_select_cols_and_rest
@@ -335,7 +336,7 @@ class GeminiRow(object):
                  variant_samples=None,
                  HET_samples=None, HOM_ALT_samples=None,
                  HOM_REF_samples=None, UNKNOWN_samples=None,
-                 info=None,formatter=DefaultRowFormat(None)):
+                 info=None, formatter=DefaultRowFormat(None)):
         self.row = row
         self.gts = gts
         self.info = info
@@ -690,9 +691,9 @@ class GeminiQuery(object):
         # open up a new database
         if os.path.exists(self.db):
             self.cluster = Cluster(self.db)
-            #self.cluster.isolation_level = None
+            # self.cluster.isolation_level = None
             # allow us to refer to columns by name
-            #self.cluster.row_factory = sqlite3.Row
+            # self.cluster.row_factory = sqlite3.Row
             self.session = self.cluster.cursor()
 
 
@@ -785,7 +786,7 @@ class GeminiQuery(object):
                                    if not description_tuple[0].startswith("gt")]
             self.report_cols = self.all_query_cols
 
-    #TODO: convert wildcard to query expression on correct samples table(s)	
+    # TODO: convert wildcard to query expression on correct samples table(s)	
     def _get_matching_sample_ids(self, wildcard):
         """
         Helper function to convert a sample wildcard
@@ -797,7 +798,7 @@ class GeminiQuery(object):
         if wildcard.strip() != "*":
             query += ' WHERE ' + wildcard
 
-        sample_info = [] # list of sample_id/name tuples
+        sample_info = []  # list of sample_id/name tuples
         self.session.execute(query)
         for row in self.session:
             sample_info.append(row[0])
@@ -859,9 +860,9 @@ class GeminiQuery(object):
                     if len(t) == 0:
                         continue
                     if (t.find("gt") >= 0 or t.find("GT") >= 0):
-                        #TODO: niet meer nodig, denk ik?
-                        #corrected = self._correct_genotype_col(t)
-                        #corrected_gt_filter.append(corrected)
[…]

### 2015-03-17 gemini/gemini/query_expressions.py

@@ -22,6 +22,9 @@ class Expression(object):
         rows = session.execute(query)
         return rows_as_list(rows)
     
+    def to_string(self):
+        return "(" + self.where_clause +")"
+    
 class AND_expression(object):
     
     def __init__(self, left, right):
@@ -32,6 +35,10 @@ class AND_expression(object):
         temp = self.left.evaluate(session, starting_set)
         return self.right.evaluate(session, temp)
     
+    def to_string(self):
+        res = "(" + self.left.to_string() + " AND " + self.right.to_string() + ")"
+        return res
+    
 class OR_expression(object):
     
     def __init__(self, left, right):
@@ -40,6 +47,10 @@ class OR_expression(object):
         
     def evaluate(self, session, starting_set):
         return union(self.left.evaluate(session, starting_set), self.right.evaluate(session, starting_set))
+    
+    def to_string(self):
+        res = "(" + self.left.to_string() + " OR " + self.right.to_string() + ")"
+        return res
     
 class NOT_expression(object):
     
@@ -55,6 +66,9 @@ class NOT_expression(object):
             correct_starting_set = starting_set
             
         return diff(correct_starting_set, self.exp.evaluate(session, correct_starting_set))
+    
+    def to_string(self):
+        return "NOT " + self.exp.to_string()
 
 def diff(list1, list2):
     return filter(lambda x: not x in list2, list1)

### 2015-03-17 gemini/gemini/GeminiQuery.py
gt-filters and wildcards work, but not yet together

@@ -457,12 +457,12 @@ class GeminiQuery(object):
         # list of samples ids for each clause in the --gt-filter
         self.sample_info = collections.defaultdict(list)
 
-        # map sample names to indices. e.g. self.sample_to_idx[NA20814] -> 323
+        '''# map sample names to indices. e.g. self.sample_to_idx[NA20814] -> 323
         self.sample_to_idx = util.map_samples_to_indices(self.session)
         # and vice versa. e.g., self.idx_to_sample[323] ->  NA20814
         self.idx_to_sample = util.map_indices_to_samples(self.session)
         self.idx_to_sample_object = util.map_indices_to_sample_objects(self.session)
-        self.sample_to_sample_object = util.map_samples_to_sample_objects(self.session)
+        self.sample_to_sample_object = util.map_samples_to_sample_objects(self.session)'''
         self.formatter = out_format
         self.predicates = [self.formatter.predicate]
 
@@ -484,8 +484,10 @@ class GeminiQuery(object):
         self.query = self.formatter.format_query(query)
         self.gt_filter = gt_filter
         if self._is_gt_filter_safe() is False:
-            sys.exit("ERROR: invalid --gt-filter command.")
-
+            sys.exit("ERROR: unsafe --gt-filter command.")
+        if not self.gt_filter == None:
+            self.gt_filter = self.gt_filter.replace('==','=')
+        
         self.show_variant_samples = show_variant_samples
         self.variant_samples_delim = variant_samples_delim
 
@@ -689,20 +691,17 @@ class GeminiQuery(object):
         Establish a connection to the requested Gemini database.
         """
         # open up a new database
-        if os.path.exists(self.db):
-            self.cluster = Cluster(self.db)
-            # self.cluster.isolation_level = None
-            # allow us to refer to columns by name
-            # self.cluster.row_factory = sqlite3.Row
-            self.session = self.cluster.cursor()
+        
+        self.cluster = Cluster()
+        self.session = self.cluster.connect('gemini_keyspace')
 
 
     def _collect_sample_table_columns(self):
         """
         extract the column names in the samples table into a list
         """
-        self.session.execute('select * from samples limit 1')
-        self.sample_column_names = [tup[0] for tup in self.session.description]
+        column_names = self.cluster.metadata.keyspaces['gemini_keyspace'].tables['samples'].columns.keys()
+        self.sample_column_names = column_names
 
     def _is_gt_filter_safe(self):
         """
@@ -725,10 +724,10 @@ class GeminiQuery(object):
             return False
 
[…]

### 2015-03-17 gemini/update_query.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+
+sudo python setup.py install
+
+gemini query -q "select variant_id from variants" \
+		--gt-filter "gt_type.child_1==HET" \
+		test/my.db

### 2015-03-17 gemini/gemini/row_formats.py
row_formats module

@@ -1,0 +1,336 @@
+import abc
+from gemini.gemini_subjects import get_subjects
+from gemini.GeminiQuery import GeminiQuery
+import re
+from gemini.sql_utils import ensure_columns
+from gemini_constants import HET, HOM_ALT, HOM_REF, UNKNOWN
+
+import json
+import os
+import sys
+import collections
+
+
+# gemini imports
+class RowFormat:
+    """A row formatter to output rows in a custom format.  To provide
+    a new output format 'foo', implement the class methods and set the
+    name field to foo.  This will automatically add support for 'foo' to
+    anything accepting the --format option via --format foo.
+    """
+
+    __metaclass__ = abc.ABCMeta
+
+    @abc.abstractproperty
+    def name(self):
+        return
+
+    @abc.abstractmethod
+    def format(self, row):
+        """ return a string representation of a GeminiRow object
+        """
+        return '\t'.join([str(row.row[c]) for c in row.row])
+
+    @abc.abstractmethod
+    def format_query(self, query):
+        """ augment the query with columns necessary for the format or else just
+        return the untouched query
+        """
+        return query
+
+    @abc.abstractmethod
+    def predicate(self, row):
+        """ the row must pass this additional predicate to be output. Just
+        return True if there is no additional predicate"""
+        return True
+
+    @abc.abstractmethod
+    def header(self, fields):
+        """ return a header for the row """
+        return "\t".join(fields)
+
+class DefaultRowFormat(RowFormat):
+
+    name = "default"
+
+    def __init__(self, args):
+        pass
+
+    def format(self, row):
+        return '\t'.join([str(row.row[c]) for c in row.row])
+
+    def format_query(self, query):
+        return query
+
+    def predicate(self, row):
+        return True
+
+    def header(self, fields):
+        """ return a header for the row """
+        return "\t".join(fields)
+
+class CarrierSummary(RowFormat):
+    """
+    Generates a count of the carrier/noncarrier status of each feature in a given
+    column of the sample table
+
+    Assumes None == unknown.
+    """
+    name = "carrier_summary"
+
+    def __init__(self, args):
+        subjects = get_subjects(args)
+        self.carrier_summary = args.carrier_summary
+
+        # get the list of all possible values in the column
+        # but don't include None, since we are treating that as unknown.
[…]

### 2015-03-22 gemini/droptables.py
snpEff compatible, async insert instead of batchstmts

@@ -5,7 +5,7 @@ from cassandra.query import SimpleStatement
 
 answer = raw_input('Drop all gemini tables (y | n)? ')
 
-if answer.lower().startswith(('ja', 'yes', 'oui', 'si')):
+if answer.lower().startswith(('j', 'y', 'oui', 'si')):
 	print 'Dropping gemini tables...'
 	cluster = Cluster()
 	session = cluster.connect('gemini_keyspace')

### 2015-03-22 gemini/gemini/GeminiQuery.py

@@ -20,7 +20,7 @@ from gemini.query_expressions import Simple_expression, AND_expression,\
     NOT_expression, OR_expression
 from gemini.sql_utils import get_query_parts
 from types import UnicodeType
-from cassandra.query import dict_factory, ordered_dict_factory
+from cassandra.query import ordered_dict_factory
 
 
 # gemini imports

### 2015-03-22 gemini/gemini/database_cassandra.py

@@ -2,7 +2,8 @@
 
 from itertools import repeat
 
-from cassandra.query import BatchStatement
+from cassandra.query import BatchStatement, SimpleStatement
+import Queue
 
 def drop_tables(session):
     session.execute("DROP TABLE IF EXISTS variants")
@@ -12,11 +13,12 @@ def drop_tables(session):
     session.execute("DROP TABLE IF EXISTS gene_detailed")
     session.execute("DROP TABLE IF EXISTS gene_summary")
 
-def create_tables(session):
+def create_tables(session, gt_column_names, extra_sample_columns):
     """
     Create our master DB tables
     """
-    session.execute('''CREATE TABLE if not exists variant_impacts  (   \
+        
+    session.execute(SimpleStatement('''CREATE TABLE if not exists variant_impacts  (   \
                     variant_id uuid,                               \
                     anno_id int,                                  \
                     gene text,                                        \
@@ -36,16 +38,16 @@ def create_tables(session):
                     polyphen_score float,                             \
                     sift_pred text,                                   \
                     sift_score float,                                 \
-                    PRIMARY KEY((variant_id, anno_id)))''')
+                    PRIMARY KEY((variant_id, anno_id)))'''))
 
-    session.execute('''CREATE TABLE if not exists resources ( \
+    session.execute(SimpleStatement('''CREATE TABLE if not exists resources ( \
                      name text PRIMARY KEY,                  \
-                     resource text)''')
+                     resource text)'''))
 
-    session.execute('''CREATE TABLE if not exists version ( \
-                     version text PRIMARY KEY)''')
+    session.execute(SimpleStatement('''CREATE TABLE if not exists version ( \
+                     version text PRIMARY KEY)'''))
     
-    session.execute('''CREATE TABLE if not exists gene_detailed (       \
+    session.execute(SimpleStatement('''CREATE TABLE if not exists gene_detailed (       \
                    uid int PRIMARY KEY,                                \
                    chrom text,                                         \
                    gene text,                                          \
@@ -64,9 +66,9 @@ def create_tables(session):
                    strand text,                                        \
                    synonym text,                                       \
                    rvis_pct float,                                     \
-                   mam_phenotype_id text)''')
+                   mam_phenotype_id text)'''))
     
[…]

### 2015-03-22 gemini/gemini/gemini_load.py

@@ -13,6 +13,7 @@ import gemini_annotate
 import uuid
 import time
 import datetime
+from cluster_helper.cluster import cluster_view
 
 
 def load(parser, args):
@@ -44,7 +45,7 @@ def load(parser, args):
     annotations.load_annos( args )
     
     gemini_loader = GeminiLoader(args)
-    gemini_loader.create_db()
+    gemini_loader.setup_db()
     gemini_loader.single_core_stuff()
 
     if args.scheduler:
@@ -60,7 +61,6 @@ def load_singlecore(args):
     # the gemini db and files from the VCF
     gemini_loader = GeminiLoader(args)
     gemini_loader.connect_to_db()
-    gemini_loader.single_core_stuff()
     gemini_loader.populate_from_vcf()
 
 
@@ -79,14 +79,13 @@ def load_multicore(args):
     load_chunks_multicore(grabix_file, args)
     # gemini_annotate.add_extras(args.db, chunks)
 
-'''
-    def load_ipython(args):
+
+def load_ipython(args):
     grabix_file = bgzip(args.vcf)
     with cluster_view(*get_ipython_args(args)) as view:
         chunks = load_chunks_ipython(grabix_file, args, view)
-        merge_chunks_ipython(chunks, args.db, view)
-    gemini_annotate.add_extras(args.db, chunks)
-'''
+    # gemini_annotate.add_extras(args.db, chunks)
+
 
 def get_chunk_name(chunk):
     return "--chunkdb " + chunk

### 2015-03-22 gemini/gemini/gemini_load_chunk.py

@@ -14,7 +14,7 @@ import version
 from ped import load_ped_file
 import gene_table
 import infotag
-from database_cassandra import insert, batch_insert, create_tables, create_variants_table, create_samples_tables
+from database_cassandra import insert, batch_insert, create_tables
 import annotations
 import func_impact
 import severe_impact
@@ -28,20 +28,21 @@ from blist import blist
 from table_schemes import get_column_names
 from gemini.ped import get_ped_fields
 from uuid import uuid4
+import time
 
 class GeminiLoader(object):
     """
     Object for creating and populating a gemini
     database and auxillary data files.
     """
-    def __init__(self, args, buffer_size=10000):
+    def __init__(self, args, buffer_size=1000):
         self.args = args
-        self.buffer_size = buffer_size
-        self._get_anno_version()
-        
         
         # create a reader for the VCF file
         self.vcf_reader = self._get_vcf_reader()
+        
+        self.buffer_size = buffer_size
+        self._get_anno_version()        
         
         if not self.args.no_genotypes:
             self.samples = self.vcf_reader.samples
@@ -110,16 +111,15 @@ class GeminiLoader(object):
         """
         """
         import gemini_annotate  # avoid circular dependencies
-        self.counter = 0
         self.var_buffer = blist([])
         self.var_impacts_buffer = blist([])
         self.var_subtypes_buffer = blist([])
-        self.var_sample_gt_types_buffer = blist([])
-        self.var_sample_gt_depths_buffer = blist([])
         buffer_count = 0
         self.skipped = 0
         extra_file, extraheader_file = gemini_annotate.get_extra_files(self.args.db)
         extra_headers = {}
+        self.counter = 0
+        start_time = time.clock()
         with open(extra_file, "w") as extra_handle:
             # process and load each variant in the VCF file
             for var in self.vcf_reader:
@@ -133,33 +133,34 @@ class GeminiLoader(object):
                 # add the core variant info to the variant buffer
                 self.var_buffer.append(variant)
                 self.var_subtypes_buffer.append([variant[4], variant[11], variant[12]])
+                
+                var_sample_gt_types_buffer = blist([])
+                var_sample_gt_depths_buffer = blist([])
+                
                 for sample in sample_info:
-                    self.var_sample_gt_types_buffer.append([variant_id, sample[0], sample[1]])
[…]

### 2015-03-22 gemini/update_load.sh

@@ -2,6 +2,6 @@
 
 sudo python setup.py install 
 
-python droptables.py
-gemini load -v test/test.comp_het.vcf -p test/test.comp_het.ped --cores 4 --skip-cadd --skip-gerp-bp test/my.db
+#python droptables.py
+gemini load --skip-gene-tables --test-mode -v test/test.query.vcf --cores 4 --skip-cadd --skip-gerp-bp -t snpEff test/my.db
 

### 2015-03-23 gemini/droptables.py
timing gemini load

@@ -15,4 +15,4 @@ if answer.lower().startswith(('j', 'y', 'oui', 'si')):
 	for table in tables:
 		session.execute('DROP TABLE IF EXISTS %s' % table )
 
-
+print "Dropping tables finished."

### 2015-03-23 gemini/gemini/gemini_load.py

@@ -45,9 +45,12 @@ def load(parser, args):
     annotations.load_annos( args )
     
     gemini_loader = GeminiLoader(args)
+    start_time = time.time()
     gemini_loader.setup_db()
+    time_2 = time.time()
     gemini_loader.single_core_stuff()
-
+    time_3 = time.time()
+    
     if args.scheduler:
         #load_ipython(args)
         sys.stdout.write("Just testing, no fancy scheduler stuff available yet")
@@ -55,6 +58,11 @@ def load(parser, args):
         load_multicore(args)
     else:
         load_singlecore(args)
+    end_time = time.time()
+    print "Finished loading in %s s" % (end_time - start_time)
+    print "Creating db tables took %s s" % (time_2 - start_time)
+    print "Populating samples tables (single-core) took %s s" % (time_3 - time_2)
+    print "Populating variants tables (parallelised) took %s s" % (end_time - time_3)
 
 def load_singlecore(args):
     # create a new gemini loader and populate

### 2015-03-23 gemini/gemini/gemini_load_chunk.py

@@ -119,7 +119,7 @@ class GeminiLoader(object):
         extra_file, extraheader_file = gemini_annotate.get_extra_files(self.args.db)
         extra_headers = {}
         self.counter = 0
-        start_time = time.clock()
+        start_time = time.time()
         with open(extra_file, "w") as extra_handle:
             # process and load each variant in the VCF file
             for var in self.vcf_reader:
@@ -138,6 +138,7 @@ class GeminiLoader(object):
                 var_sample_gt_depths_buffer = blist([])
                 
                 for sample in sample_info:
+                    #TODO: check if gt_type is None. Compare if faster to check client side i.s.o. Cassandra - side
                     var_sample_gt_depths_buffer.append([variant_id, sample[0], sample[2]])
                     var_sample_gt_types_buffer.append([variant_id, sample[0], sample[1]])
                                     
@@ -174,7 +175,7 @@ class GeminiLoader(object):
         batch_insert(self.session, 'variants_by_sub_type_call_rate',\
                                             get_column_names('variants_by_sub_type_call_rate'), self.var_subtypes_buffer)
         
-        end_time = time.clock()
+        end_time = time.time()
         elapsed_time = end_time - start_time            
         sys.stderr.write("pid " + str(os.getpid()) + ": " +
                          str(self.counter) + " variants processed in %s s.\n" % elapsed_time)

### 2015-03-23 gemini/update_load.sh

@@ -2,6 +2,7 @@
 
 sudo python setup.py install 
 
-#python droptables.py
-gemini load --skip-gene-tables --test-mode -v test/test.query.vcf --cores 4 --skip-cadd --skip-gerp-bp -t snpEff test/my.db
+
+python droptables.py
+gemini load --skip-gene-tables --test-mode -v test/test.query.vcf --cores 5 --skip-cadd --skip-gerp-bp -t snpEff test/my.db
 

### 2015-03-26 thesis/Inleiding.tex
Overzicht NoSQL & NewSQL

--- 2015-03-16 thesis/cha1.tex
@@
 GEMINI (van GEnome MINIng) is een tool voor flexibele analyse van genetische variaties in grote datasets van genomen, ontwikkeld aan de University of Virginia \cite{10.1371/journal.pcbi.1003153}.
-GEMINI vertrekt daarbij van VCF (Variant Call Format) bestanden, laadt deze in in een database, en biedt dan de mogelijkheid queries uit te voeren op deze databank.
+GEMINI vertrekt daarbij van VCF (Variant Call Format) tekst-bestanden, laadt deze in in een database, voegt allerlei annotaties toe met bijkomende informatie van enkele vermaarde onderzoeksinstituten en biedt dan de mogelijkheid queries uit te voeren op deze databank.
 De huidige operationele versie van GEMINI draait op de eenvoudige SQL-databank SQLite, maar deze laat op gebied van schaalbaarheid en performantie te wensen over.
@@
 De verwachtingen zijn echter ook hier dat PostgreSQL op langere termijn niet zal kunnen schalen naar datasets van (honderd)duizenden genomen.\\
-GEMINI bewaart de data over genetische varianten in enkele zeer grote tabellen.
-Een eerste vereiste voor een database is dus om deze tabulaire data goed te kunnen voorstellen en beheren, bij voorkeur d.m.v.
-automatische verspreiding over verschillende nodes in een cluster.
-Bovendien voert GEMINI na het inladen van de VCF bestanden enkel nog read-queries uit, dus zijn de belangrijkste verdere vereisten voor een database hoge read-throughput, goede query-mogelijkheden en indexeringsmechanismes.
-Omdat GEMINI in Python ge\"implementeerd is, is een Python-API ook nuttig.
+Naar de toekomst toe is het dus noodzakelijk uit kijken naar andere technologie\"en en types databanken om GEMINI ook bruikbaar te maken voor grootschaligere experimenten.

### 2015-03-26 thesis/NoSQL_overzicht.tex
Overzicht NoSQL & NewSQL

\chapter{Achtergrond databanken}
\label{cha2}

\section{NoSQL}

NoSQL databanken zijn er in verschillende soorten en kunnen op basis van hun datamodel in een aantal categorie\"en onderverdeeld worden:

\begin{itemize}
\item[Key-Value stores] Deze zijn vergelijkbaar met dictionaries en mappen unieke keys op values. Deze values zijn voor de databank volledig betekenisloze byte-arrays en de enige manier om ze op te vragen, is via de bijhoren key. Voor zeer eenvoudige toepassingen resulteert dit in hoge lees- en schrijf-throughput, maar meer geavanceerde features zoals indexing, queries, en het modelleren van relaties binnen de data zijn hierdoor niet mogelijk\cite{hecht2011nosql}\cite{grolinger2013data}.

\item[Columnar stores] Deze zijn gebaseerd op het datamodel dat Google's BigTable heeft ge\"introduceerd en slaan data op in een spaarse, gedistribueerde, persistente en multidimensionele gesorteerde map\cite{chang2008bigtable}. In het geval van BigTable zijn dit drie dimensies: row key, column key en een timestamp. Omdat ook hier het systeem de opgeslagen data niet interpreteert, is het modelleren van relaties niet po een effici\"ente manier mogelijk. Dit wordt overgelaten aan de bovenliggende applicatie\cite{hecht2011nosql}.

\item[Document stores] Deze bewaren data als key-value paren en encapsuleren deze in documenten. Values kunnen van een brede waaier aan types zijn, zoals geneste documenten, lijsten of scalars. De namen van attributen kunnen dynamisch gespecifieerd worden tijdens runtime en moeten geen vooraf vastgelegd schema volgen\cite{cattell2011scalable}. Dit is geschikt voor het modelleren van ingewikkelde datastructuren. Vele document stores gebruiken het JSON-bestandsformaa (of een daarvan afgeleide vorm). In tegenstelling tot columnar stores, zijn de waarden in documenten niet betekenisloos voor het document store systeem en het is dus mogelijk hier indexen op te defini\"eren en queries op uit te voeren\cite{hecht2011nosql}.

\item[Graph databases] Zoals de naam doet vermoeden, stammen deze uit grafentheorie en maken ze gebruik van grafen als datamodel. Ze zijn uitermate geschikt om sterk verweven data te beheren, van bronnen zoals sociale netwerken of location based services. Hiervoor doen ze beroep op effici\"ente mechanismen om grafen te doorlopen waar andere systemen kostelijke operaties als recursieve joins gebruiken\cite{hecht2011nosql}.

\end{itemize}

De term NewSQL slaat op een verzameling systemen die ambi\"eren het klassieke relationele datamodel te combineren met de schaalbaarheid, distributie en fouttolerantie van NoSQL systemn. Hoewel ze allen de gebruiker het relationele model en SQL-achtige query mogelijkheden bieden, verschillen NewSQL stores onderling grondig, afhankelijk van de onderliggende architectuur. Zo zijn er onder meer systemen die gebouwd zijn bovenop bestaande NoSQL databanken, en andere die alle data in main memory opslaan.\cite{grolinger2013data}

### 2015-03-26 thesis/gemini_beschrijving.tex
GEMINI datamodel & inladen

\chapter{GEMINI overzicht}
\label{cha3}

Zoals eerder vermeld is GEMINI een applicatie voor de flexibele analyse van genoomdata van populaties van menselijke individu\"en. Deze sectie gaat dieper in op de belangrijkste features en het onderliggende datamodel van GEMINI in zijn oorspronkelijke vorm.

\section{Datamodel}

GEMINI importeert genetische variants en genotypes van alle gesampelde individu\"en (ook 'samples') vanuit een VCF file in een relationele database.
Daarnaast kan extra informatie over de samples, zoals geslacht, phenotype en onderlinge verwantschappen, meegegeven worden in een PED-file (van pedigree) om latere analyse te vergemakkelijken.\\

Elke variant in een input VCF file wordt uitvoerig geannoteerd na automatische vergelijking met bestaande of door de gebruiker gedefinieerde genoom-annotatiebestanden. De geannoteerde variants vormen de rijen van de hoofdtabel van de database, de \texttt{variants}-tabel. Deze tabel bevat ook voor elke variant informatie over elke sample, zoals diens genotype, de kwaliteit en diepte van de meting voor de variant in kwestie. In de SQLite-versie van GEMINI wordt dit opgeslagen als een gecomprimeerde array per variant, 1 voor elke sample-eigenschap: zo is er een \texttt{gt\_type}-kolom met arrays met de genotypes, en een \texttt{gt\_depth}-kolom met arrays met de diepte van de meting van elke sample voor elke variant. Samen met de \texttt{samples}-tabel, die voor elke sample zaken als het geslacht, phenotype en familierelaties bijhoudt, ligt de \texttt{variants}-tabel aan de basis van de uitgebreide query-mogelijkheden die GEMINI biedt.\\

Daarnaast zijn er nog tabellen zoals de \texttt{variant\_impacts}- en \texttt{gene\_detailed}-tabellen die respectievelijk extra informatie over de variants en het menselijk genoom bevatten. Deze informatie komt in het eerste geval eveneens uit de annotatiebestanden, en in het tweede uit tekstbestanden met referentie-informatie over het menselijk genoom die GEMINI, indien gewenst mee inlaadt.\\
Ten laatste zijn er nog enkele kleine tabellen met meta-informatie, zoals de \texttt{resources}-tabel die de gebruikte annotatie-files bevat, en de \texttt{version}-tabel die bevat doorn welke versie van GEMINI de data ingeladen is.\\\\

Een belangrijke troef van GEMINI en z'n datamodel is de flexibiliteit die het laat naar de gebruiker. Zo kan de gebruiker zelfgedefinieerde annotatie-files gebruiken, en zelf kolommen toevoegen aan de PED-files met informatie over de samples. Deze extra informatie zal GEMINI automatisch in de resp. \texttt{variants}- en \texttt{samples}-tabel opnemen en kan de gebruiker later ook doorzoeken.

\begin{figure}[h]
\includegraphics[scale=0.8]{gemini_schema}
\caption{Een overzicht van GEMINI's schema}
\label{gemini_schema_pic}
\end{figure}

\section{Inladen}
%TODO ref nr IPython dink
Het inladen van de data uit VCF-bestanden is een computationeel intensieve operatie, enerzijds omwille van de enorme grootte van deze bestanden, en anderzijds omdat in deze fase ook alle variants geannoteerd moeten worden. Om het proces te versnellen, biedt GEMINI de mogelijkheid het werk te paralleliseren door het VCF-bestand de comprimeren, op het bestaande bestand een index te defini\"eren en zo het werk te verdelen. Dit kan simpelweg over meerdere processoren binnen 1 computer zijn, maar via de IPython-interface ook over volledige clusters van computers.\\

### 2015-03-26 thesis/gemini_beschrijving.tex
gemini querying beschrijving (begin)

@@
 Om het proces te versnellen, biedt GEMINI de mogelijkheid het werk te paralleliseren door het VCF-bestand de comprimeren, op het bestaande bestand een index te defini\"eren en zo het werk te verdelen.
-Dit kan simpelweg over meerdere processoren binnen 1 computer zijn, maar via de IPython-interface ook over volledige clusters van computers.\\
+Dit kan over meerdere processoren binnen 1 computer zijn, maar via de IPython-interface ook over volledige clusters van computers.\\
+
+\section{Querying}
+
+GEMINI laat de gebruiker toe de opgeslagen genoomdata te doorzoeken.
+Enerzijds via gewone SQL-queries, maar omdat het onderzoeken van individuele genotypes van cruciaal belang is in het onderzoek naar ziektes en SQL standaard niet de individuele genotypes in array-kolommen kan opvragen, biedt GEMINI bovendien een verrijkte SQL-syntax.\\
+Zo laat het de gebruiker toe via filters en wildcards de gewenste genotype-eigenschappen en samples te defini\"eren.
+
+\subsection{Sample filters/wildcards} 
+In SQL \texttt{SELECT}-statements kan de gebruiker met een filter van de vorm column.sample of een wilcard van de vorm (column).(wildcard) uitdrukken welke genotype-kolommen van welke samples hij/zij wilt zien.
+Wanneer de gebruiker ge\"interesseerd is in bijvoorbeeld het genotype van sample Jan, wordt dit:\\
+\texttt{SELECT gt\_types.Jan FROM variants}\\\\
+Als de gebruiker de diepte van de genotypes alle vrouwelijke samples wil zien, wordt dit:\\\\
+\texttt{SELECT (gt\_depths).(sex = 'female')}.\\
+GEMINI vertaalt de wildcard automatisch naar een query op de \texttt{samples}-tabel, en kan vervolgens voor alle samples die aan de voorwaarden voldoen, de waarde uit de gevraagde kolom tonen.
+
+\subsection{Genotype filters}
+Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL WHERE-clauses uitbreiden met zogenaamde genotype filters.
+Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor sample Jan heterozygoot is en Mieke allesbehalve heterozygoot is, kan dit met:\\\\
+
+%TODO aanhalingstekens fixen
+\texttt{gemini query -q "SELECT * FROM variants" --gt-filter "gt\_types.Jan == HET and gt\_types.Mieke != HET"}

### 2015-03-26 thesis/gemini_beschrijving.tex
beschrijving gemini wildcards

@@
 Zoals eerder vermeld is GEMINI een applicatie voor de flexibele analyse van genoomdata van populaties van menselijke individu\"en.
-Deze sectie gaat dieper in op de belangrijkste features en het onderliggende datamodel van GEMINI in zijn oorspronkelijke vorm.
+Deze sectie gaat dieper in op de belangrijkste features en het onderliggende datamodel van GEMINI in zijn oorspronkelijke vorm.\cite{10.1371/journal.pcbi.1003153}\cite{gemini_docs}
 
@@
 In SQL \texttt{SELECT}-statements kan de gebruiker met een filter van de vorm column.sample of een wilcard van de vorm (column).(wildcard) uitdrukken welke genotype-kolommen van welke samples hij/zij wilt zien.
-Wanneer de gebruiker ge\"interesseerd is in bijvoorbeeld het genotype van sample Jan, wordt dit:\\
+Wanneer de gebruiker ge\"interesseerd is in bijvoorbeeld het genotype van Jan, wordt dit:\\
 \texttt{SELECT gt\_types.Jan FROM variants}\\\\
@@
 
-\subsection{Genotype filters}
+\subsection{Genotype filters/wildcards}
 Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL WHERE-clauses uitbreiden met zogenaamde genotype filters.
-Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor sample Jan heterozygoot is en Mieke allesbehalve heterozygoot is, kan dit met:\\\\
+Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor Jan heterozygoot is en Mieke alles behalve heterozygoot is, kan dit met:\\
 
-%TODO aanhalingstekens fixen
-\texttt{gemini query -q "SELECT * FROM variants" --gt-filter "gt\_types.Jan == HET and gt\_types.Mieke != HET"}
+\noindent\texttt{\$ gemini query -q} "\texttt{SELECT * FROM variants}"\textbackslash \\\texttt{--gt-filter }"\texttt{gt\_types.Jan == HET and gt\_types.Mieke != HET}"\\
+
+\noindent Vaak is het de bedoeling om eenzelfde beperking op te leggen aan meerdere samples.
+Op bovenstaande manier wordt dit al monikenwerk.
+Om dit te verhinderen, zijn er genotype wildcards van volgend formaat:\\
+\texttt{(column).(sample\_wildcard).(gt\_wildcard\_rule).(rule\_enforcement)}.
+\begin{description}
+\item[column] Staat voor de genotype kolom waarop de beperking slaat.
+\item[sample\_wildcard] Een wildcard om aan te duiden voor welke samples de beperking moet gelden. Analoog met de eerder vermelde sample wildcards.
+\item[gt\_wildcard\_rule] De beperking op de genotype kolom.
+\item[rule\_enforcement] Duidt aan hoeveel van de geselecteerde samples aan de opgelegde gt\_wildcard\_rule moeten voldoen. Dit kan \texttt{all, any, none, of count <comparison>} zijn.
+\end{description}
+
+\noindent Alle variants opvragen waarvoor alle mannen heterozygoot zijn, gaat bijvoorbeeld met:
+
+\noindent\texttt{\$ gemini query -q "}\texttt{SELECT * FROM variants"}\textbackslash \\
+\texttt{--gt-filter" }\texttt{(gt\_types).(sex = 'male').(=HET).(all)"}\\
+
+\noindent Alle variants opvragen waarvoor minstens 1 bruinharig individu een genotype van diepte minstens 100 heeft, met:
+
+\noindent\texttt{\$ gemini query -q "}\texttt{SELECT * FROM variants"}\textbackslash \\
+\texttt{--gt-filter" }\texttt{(gt\_depths).(hair\_colour = 'brown').(>=100).(any)"}\\
+
+\noindent Alle variants opvragen waarvoor minder dan 10 individu\"en een genotype met diepte minder dan 50 hebben, met:
+
+\noindent\texttt{\$ gemini query -q "}\texttt{SELECT * FROM variants"}\textbackslash \\
+\texttt{--gt-filter" }\texttt{(gt\_depths).(*).(=HET).(count < 10)"}\\
+
+Al de bovenstaande filters kunnen met elkaar gecombineerd worden, evenals met gewone WHERE-clauses op de andere kolommen van de variants-tabel.

### 2015-03-27 gemini/gemini/GeminiQuery.py
iterating over results, header, formatting etc

@@ -13,14 +13,13 @@ import compression
 from gemini_constants import HOM_ALT, HOM_REF, HET, UNKNOWN
 from gemini_subjects import get_subjects
 from gemini_utils import (OrderedSet, OrderedDict, itersubclasses, partition_by_fn)
-import numpy as np
 from sql_utils import ensure_columns, get_select_cols_and_rest
 from collections import namedtuple
 from gemini.query_expressions import Simple_expression, AND_expression,\
-    NOT_expression, OR_expression
+    NOT_expression, OR_expression, rows_as_list
 from gemini.sql_utils import get_query_parts
 from types import UnicodeType
-from cassandra.query import ordered_dict_factory
+from cassandra.query import ordered_dict_factory, tuple_factory
 
 
 # gemini imports
@@ -60,7 +59,6 @@ class RowFormat:
     def header(self, fields):
         """ return a header for the row """
         return "\t".join(fields)
-
 
 class DefaultRowFormat(RowFormat):
 
@@ -148,8 +146,6 @@ class CarrierSummary(RowFormat):
             elif ct == False:
                 header_columns.append("unaffected")
         return header_columns
-
-
 
 class TPEDRowFormat(RowFormat):
 
@@ -268,7 +264,7 @@ class JSONRowFormat(RowFormat):
     def header(self, fields):
         return None
 
-class VCFRowFormat(RowFormat):
+'''class VCFRowFormat(RowFormat):
 
     name = "vcf"
 
@@ -329,31 +325,18 @@ class VCFRowFormat(RowFormat):
             self.gq.run('select vcf_header from vcf_header')
             return str(self.gq.next()).strip()
         except:
-            sys.exit("Your database does not contain the vcf_header table. Therefore, you cannot use --header.\n")
+            sys.exit("Your database does not contain the vcf_header table. Therefore, you cannot use --header.\n")'''
 
 class GeminiRow(object):
 
-    def __init__(self, row, gts=None, gt_types=None,
-                 gt_phases=None, gt_depths=None,
-                 gt_ref_depths=None, gt_alt_depths=None,
-                 gt_quals=None, gt_copy_numbers=None,
+    def __init__(self, row, 
                  variant_samples=None,
                  HET_samples=None, HOM_ALT_samples=None,
                  HOM_REF_samples=None, UNKNOWN_samples=None,
                  info=None, formatter=DefaultRowFormat(None)):
         self.row = row
[…]

### 2015-03-27 gemini/gemini/database_cassandra.py

@@ -19,7 +19,7 @@ def create_tables(session, gt_column_names, extra_sample_columns):
     """
         
     session.execute(SimpleStatement('''CREATE TABLE if not exists variant_impacts  (   \
-                    variant_id uuid,                               \
+                    variant_id int,                               \
                     anno_id int,                                  \
                     gene text,                                        \
                     transcript text,                                  \
@@ -86,17 +86,17 @@ def create_tables(session, gt_column_names, extra_sample_columns):
     session.execute(SimpleStatement('''CREATE TABLE if not exists variants_by_samples_gt_type ( \
                         sample_name text, \
                         gt_type int, \
-                        variant_id uuid, \
+                        variant_id int, \
                         primary key (sample_name, gt_type, variant_id))'''))
     
     session.execute(SimpleStatement('''CREATE TABLE if not exists variants_by_samples_gt_depth( \
                         sample_name text, \
                         gt_depth int, \
-                        variant_id uuid, \
+                        variant_id int, \
                         primary key (sample_name, gt_depth, variant_id))'''))    
     
     session.execute(SimpleStatement('''CREATE TABLE IF NOT EXISTS samples_by_variants_gt_type ( \
-                    variant_id uuid, \
+                    variant_id int, \
                     gt_type int, \
                     sample_name text, \
                     PRIMARY KEY (variant_id, gt_type, sample_name))'''))
@@ -104,7 +104,7 @@ def create_tables(session, gt_column_names, extra_sample_columns):
     session.execute(SimpleStatement('''CREATE TABLE if not exists vcf_header (vcf_header text PRIMARY KEY)'''))
     
     session.execute(SimpleStatement('''CREATE TABLE if not exists variants_by_sub_type_call_rate ( \
-                        variant_id uuid, \
+                        variant_id int, \
                         sub_type text, \
                         call_rate float, \
                         PRIMARY KEY (sub_type, call_rate, variant_id))'''))
@@ -125,7 +125,7 @@ def create_variants_table(gt_column_names):
                     start int,                                  \
                     \"end\" int,                                \
                     vcf_id text,                                \
-                    variant_id uuid PRIMARY KEY,            \
+                    variant_id int PRIMARY KEY,            \
                     anno_id int,                                \
                     ref text,                                   \
                     alt text,                                   \

### 2015-03-27 gemini/gemini/gemini_load_chunk.py

@@ -27,7 +27,6 @@ from cassandra.cluster import Cluster
 from blist import blist
 from table_schemes import get_column_names
 from gemini.ped import get_ped_fields
-from uuid import uuid4
 import time
 
 class GeminiLoader(object):
@@ -106,11 +105,19 @@ class GeminiLoader(object):
         typed_column_names = concat(map(lambda x: map(lambda y: x[0] + '_' + y + ' ' + x[1], self.samples), gt_cols))
         
         return (column_names, typed_column_names)
+    
+    def _get_vid(self):
+        if hasattr(self.args, 'offset'):
+            v_id = int(self.args.offset)
+        else:
+            v_id = 1
+        return v_id
 
     def populate_from_vcf(self):
         """
         """
         import gemini_annotate  # avoid circular dependencies
+        self.v_id = self._get_vid()
         self.var_buffer = blist([])
         self.var_impacts_buffer = blist([])
         self.var_subtypes_buffer = blist([])
@@ -126,7 +133,7 @@ class GeminiLoader(object):
                 if self.args.passonly and (var.FILTER is not None and var.FILTER != "."):
                     self.skipped += 1
                     continue
-                (variant_id, variant, variant_impacts, sample_info, extra_fields) = self._prepare_variation(var)
+                (variant, variant_impacts, sample_info, extra_fields) = self._prepare_variation(var)
                 if extra_fields:
                     extra_handle.write("%s\n" % json.dumps(extra_fields))
                     extra_headers = self._update_extra_headers(extra_headers, extra_fields)
@@ -139,8 +146,8 @@ class GeminiLoader(object):
                 
                 for sample in sample_info:
                     #TODO: check if gt_type is None. Compare if faster to check client side i.s.o. Cassandra - side
-                    var_sample_gt_depths_buffer.append([variant_id, sample[0], sample[2]])
-                    var_sample_gt_types_buffer.append([variant_id, sample[0], sample[1]])
+                    var_sample_gt_depths_buffer.append([self.v_id, sample[0], sample[2]])
+                    var_sample_gt_types_buffer.append([self.v_id, sample[0], sample[1]])
                                     
                 batch_insert(self.session, 'variants_by_samples_gt_type', ["variant_id", "sample_name", "gt_type"], var_sample_gt_types_buffer)
                 batch_insert(self.session, 'samples_by_variants_gt_type', ["variant_id", "sample_name", "gt_type"], var_sample_gt_types_buffer)
[…]

### 2015-03-27 gemini/gemini/gemini_query.py

@@ -1,17 +1,13 @@
 #!/usr/bin/env python
 import os
-import sys
-from itertools import tee, ifilterfalse
 from collections import defaultdict
 
 # gemini imports
 import GeminiQuery
 from GeminiQuery import select_formatter
-from gemini_constants import *
 from gemini_region import add_region_to_query
-from gemini_subjects import (Subject, get_subjects, get_subjects_in_family,
+from gemini_subjects import (get_subjects, get_subjects_in_family,
                              get_family_dict)
-from gemini_utils import itersubclasses
 from dgidb import query_dgidb
 
 def all_samples_predicate(args):

### 2015-03-27 gemini/gemini/sql_utils.py

@@ -59,13 +59,22 @@ def get_query_parts(query):
     where_clause = ""
     rest = ""
     from_table = query[from_loc + 4: from_end].strip()
-    select = query[select_loc + 6:from_loc].strip()
+    select_clause = query[select_loc:from_loc]
+    
+    # remove the SELECT keyword from the query
+    select_pattern = re.compile("select", re.IGNORECASE)
+    select_clause = select_pattern.sub('', select_clause)
+
+    # now create and iterate through a list of of the SELECT'ed columns
+    selected_columns = select_clause.split(',')
+    selected_columns = [c.strip() for c in selected_columns]
+    
     if where_loc > -1:
         where_clause = query[where_loc + 5: where_end].strip()
     if where_end < len(query):
         rest = query[where_end:].strip()
     
-    return select, from_table, where_clause, rest   
+    return selected_columns, from_table, where_clause, rest   
 
 def ensure_columns(query, cols):
     """

### 2015-03-27 gemini/update_load.sh

@@ -3,6 +3,7 @@
 sudo python setup.py install 
 
 
-python droptables.py
-gemini load --skip-gene-tables --test-mode -v test/test.query.vcf --cores 5 --skip-cadd --skip-gerp-bp -t snpEff test/my.db
+#python droptables.py
+#gemini load --skip-gene-tables --test-mode -v test/test.query.vcf --cores 5 --skip-cadd --skip-gerp-bp -t snpEff test/my.db
 
+gemini load --skip-gene-tables --test-mode -v test/test.comp_het.vcf --skip-gerp-bp --skip-cadd -t snpEff -p test/test.comp_het.ped test/test.comp_het.db

### 2015-03-27 gemini/update_query.sh

@@ -1,7 +1,10 @@
 #!/bin/bash
 
-#sudo python setup.py install
+sudo python setup.py install
 
-gemini query -q "select variant_id, chrom, sub_type from variants" \
-		--gt-filter "[gt_depth].[sex='1'].[>100].[all] && gt_type.child_1 = HET" \
-		test/my.db
+#gemini query -q "select variant_id, chrom, sub_type from variants" \
+#		--gt-filter "[gt_depth].[sex='1'].[>100].[all] && gt_type.child_1 = HET" \
+#		--header --show-samples test/my.db
+
+gemini query -q "select variant_id, (gt_types).(sex='1') from variants limit 1" \
+		--header test/my.db

### 2015-03-28 gemini/gemini/GeminiQuery.py
reporting result columns cleaned up, works

@@ -506,7 +506,8 @@ class GeminiQuery(object):
                 self.gt_filter_exp = self._correct_genotype_filter()
                 self.query_type = "filter-genotypes"
 
-        (self.all_selected_columns, self.from_table, where_clause, self.rest_of_query) = get_query_parts(self.query)
+        (self.requested_columns, self.from_table, where_clause, self.rest_of_query) = get_query_parts(self.query)
+        self.extra_columns = []
         
         if where_clause != '':
             self.where_exp = self.parse_where_clause(where_clause, self.from_table)
@@ -531,7 +532,7 @@ class GeminiQuery(object):
         Return a header describing the columns that
         were selected in the query issued to a GeminiQuery object.
         """
-        h = [col for col in self.all_selected_columns]
+        h = [col for col in self.report_cols]
         if self.show_variant_samples:
             h += ["variant_samples", "HET_samples", "HOM_ALT_samples"]
         if self.show_families:
@@ -666,7 +667,7 @@ class GeminiQuery(object):
         if self.matches == []:
             sys.exit("No results!")
         try:
-            query = "SELECT %s FROM %s" % (','.join(self.all_selected_columns), self.from_table)
+            query = "SELECT %s FROM %s" % (','.join(self.requested_columns + self.extra_columns), self.from_table)
             if self.matches != "*":
                 if type(self.matches[0]) is UnicodeType:
                     in_clause = "','".join(self.matches)            
@@ -676,7 +677,11 @@ class GeminiQuery(object):
                     query += " WHERE %s IN (%s)" % (self.get_partition_key(self.from_table), in_clause)
             query += " " + self.rest_of_query
             self.session.row_factory = ordered_dict_factory
-            self.result = iter(self.session.execute(query))
+            
+            res = self.session.execute(query)            
+            self.report_cols = filter(lambda x: not x in self.extra_columns, res[0].keys())
+            self.result = iter(res)
+            
         except cassandra.protocol.SyntaxException as e:
             print "Cassandra error: {0}".format(e)
             sys.exit("The query issued (%s) has a syntax error." % query)
@@ -687,7 +692,7 @@ class GeminiQuery(object):
         replace sample names with indices where necessary.
         """
         if self.needs_genes:
-            self.all_selected_columns.append("gene")
+            self.requested_columns.append("gene")
 
         '''if self.needs_vcf_columns:
             self.query = self._add_vcf_cols_to_query()'''
@@ -702,29 +707,13 @@ class GeminiQuery(object):
[…]

### 2015-03-28 gemini/gemini/gemini_main.py

@@ -224,24 +224,6 @@ def main():
         import gemini_load_chunk
         gemini_load_chunk.load(parser, args)
     parser_loadchunk.set_defaults(func=loadchunk_fn)
-    
-    #########################################
-    # $ gemini merge_chunks
-    #########################################
-    parser_mergechunks = subparsers.add_parser('merge_chunks',
-            help='combine intermediate db files into the final gemini ')
-    parser_mergechunks.add_argument('--db',
-            dest='db',
-            help='The name of the final database to be loaded.')
-    parser_mergechunks.add_argument('--chunkdb',
-            nargs='*',
-            dest='chunkdbs',
-            action='append')
-
-    def mergechunk_fn(parser, args):
-        import gemini_merge_chunks
-        gemini_merge_chunks.merge_chunks(parser, args)
-    parser_mergechunks.set_defaults(func=mergechunk_fn)
 
     #########################################
     # $ gemini query

### 2015-03-28 gemini/gemini/gemini_merge_chunks.py

@@ -1,190 +1,0 @@
-#!/usr/bin/env python
-import sqlite3
-import os
-import database as gemini_db
-import gemini_utils as util
-
-
-def append_variant_info(main_curr, chunk_db):
-    """
-    Append the variant and variant_info data from a chunk_db
-    to the main database.
-    """
-
-    cmd = "attach ? as toMerge"
-    main_curr.execute(cmd, (chunk_db, ))
-
-    main_curr.execute("BEGIN TRANSACTION")
-    cmd = "INSERT INTO variants SELECT * FROM toMerge.variants"
-    main_curr.execute(cmd)
-
-    cmd = \
-        "INSERT INTO variant_impacts SELECT * FROM toMerge.variant_impacts"
-    main_curr.execute(cmd)
-    main_curr.execute("END TRANSACTION")
-
-    cmd = "detach toMerge"
-    main_curr.execute(cmd)
-
-
-def append_sample_genotype_counts(main_curr, chunk_db):
-    """
-    Append the sample_genotype_counts from a chunk_db
-    to the main database.
-    """
-    cmd = "attach ? as toMerge"
-    main_curr.execute(cmd, (chunk_db, ))
-
-    cmd = "INSERT INTO sample_genotype_counts \
-           SELECT * FROM toMerge.sample_genotype_counts"
-    main_curr.execute(cmd)
-
-    cmd = "detach toMerge"
-    main_curr.execute(cmd)
-
-
-def append_sample_info(main_curr, chunk_db):
-    """
-    Append the sample info from a chunk_db
-    to the main database.
-    """
-    cmd = "attach ? as toMerge"
-    main_curr.execute(cmd, (chunk_db, ))
-
-    cmd = "create table samples as select * from toMerge.samples where 1=0"
-    main_curr.execute(cmd)
-
-    cmd = "INSERT INTO samples SELECT * FROM toMerge.samples"
-    main_curr.execute(cmd)
-
-    cmd = "detach toMerge"
-    main_curr.execute(cmd)
-
-
-def append_resource_info(main_curr, chunk_db):
-    """
-    Append the resource info from a chunk_db
-    to the main database.
-    """
-    cmd = "attach ? as toMerge"
-    main_curr.execute(cmd, (chunk_db, ))
-
-    cmd = "INSERT INTO resources SELECT * FROM toMerge.resources"
-    main_curr.execute(cmd)
-
-    cmd = "detach toMerge"
-    main_curr.execute(cmd)
-
-
[…]

### 2015-03-28 gemini/update_query.sh

@@ -2,9 +2,11 @@
 
 sudo python setup.py install
 
-#gemini query -q "select variant_id, chrom, sub_type from variants" \
-#		--gt-filter "[gt_depth].[sex='1'].[>100].[all] && gt_type.child_1 = HET" \
-#		--header --show-samples test/my.db
+gemini query -q "select variant_id, chrom, sub_type from variants" \
+		--gt-filter "[gt_depth].[sex='1'].[>100].[all] && gt_type.child_1 = HET" \
+		--header --show-samples test/my.db
 
-gemini query -q "select variant_id, (gt_types).(sex='1') from variants limit 1" \
+gemini query -q "select variant_id, (gt_types).(sex='1') from variants" \
 		--header test/my.db
+
+gemini query -q "select * from samples" --header test/my.db

### 2015-03-28 gemini/gemini/GeminiQuery.py
multicore gt-filter wildcard evaluation

@@ -12,11 +12,11 @@ import sys
 import compression
 from gemini_constants import HOM_ALT, HOM_REF, HET, UNKNOWN
 from gemini_subjects import get_subjects
-from gemini_utils import (OrderedSet, OrderedDict, itersubclasses, partition_by_fn)
-from sql_utils import ensure_columns, get_select_cols_and_rest
+from gemini_utils import (OrderedDict, itersubclasses, partition_by_fn)
+from sql_utils import ensure_columns
 from collections import namedtuple
 from gemini.query_expressions import Simple_expression, AND_expression,\
-    NOT_expression, OR_expression, rows_as_list
+    NOT_expression, OR_expression, rows_as_list, GT_wildcard_expression
 from gemini.sql_utils import get_query_parts
 from types import UnicodeType
 from cassandra.query import ordered_dict_factory, tuple_factory
@@ -459,7 +459,7 @@ class GeminiQuery(object):
     def run(self, query, gt_filter=None, show_variant_samples=False,
             variant_samples_delim=',', predicates=None,
             needs_genotypes=False, needs_genes=False,
-            show_families=False):
+            show_families=False, nr_cores = 1):
         """
         Execute a query against a Gemini database. The user may
         specify:
@@ -469,6 +469,8 @@ class GeminiQuery(object):
         """
         self.query = self.formatter.format_query(query)
         self.gt_filter = gt_filter
+        print self.query + '; gt-filter = %s' % gt_filter
+        self.nr_cores = nr_cores
         if self._is_gt_filter_safe() is False:
             sys.exit("ERROR: unsafe --gt-filter command.")
         if not self.gt_filter == None:
@@ -560,14 +562,6 @@ class GeminiQuery(object):
             if 'info' in self.report_cols:
                 info = compression.unpack_ordereddict_blob(row['info'])
 
-            if self._query_needs_genotype_info():
-                
-                het_names = self._get_variant_samples(row['variant_id'], HET)
-                hom_alt_names = self._get_variant_samples(row['variant_id'], HOM_ALT)
-                hom_ref_names = self._get_variant_samples(row['variant_id'], HOM_REF)
-                unknown_names = self._get_variant_samples(row['variant_id'], UNKNOWN)
-                variant_names = het_names + hom_alt_names
-
             fields = OrderedDict()
 
             for idx, col in enumerate(self.report_cols):
@@ -579,12 +573,20 @@ class GeminiQuery(object):
[…]

### 2015-03-28 gemini/gemini/gemini_main.py

@@ -270,6 +270,12 @@ def main():
                               metavar='STRING',
                               help='The delimiter to be used with the --show-samples option.',
                               default=',')
+    
+    parser_query.add_argument('--cores',
+                             dest='cores',
+                             default=1,
+                             type=int,
+                             help="Number of cores to use to interpret gt-filter wildcards in parallel.")
 
     parser_query.add_argument('--header',
                               dest='use_header',

### 2015-03-28 gemini/gemini/gemini_query.py

@@ -119,7 +119,7 @@ def run_query(args):
     gq = GeminiQuery.GeminiQuery(args.db, out_format=formatter)
     gq.run(args.query, args.gt_filter, args.show_variant_samples,
            args.sample_delim, predicates, genotypes_needed,
-           gene_needed, args.show_families)
+           gene_needed, args.show_families, args.cores)
 
     if args.use_header and gq.header:
         print gq.header
@@ -141,7 +141,7 @@ def run_query(args):
         gq = GeminiQuery.GeminiQuery(args.db, out_format=formatter)
         gq.run(args.query, args.gt_filter, args.show_variant_samples,
            args.sample_delim, predicates, genotypes_needed,
-           gene_needed, args.show_families)
+           gene_needed, args.show_families, args.cores)
 
         # report the query results with DGIdb info added at the end.
         for row in gq:

### 2015-03-28 gemini/gemini/query_expressions.py

@@ -5,6 +5,10 @@ Created on Mar 16, 2015
 '''
 import abc
 from types import UnicodeType
+from multiprocessing import Pipe
+from multiprocessing.process import Process
+from cassandra.cluster import Cluster
+import array
 
 class Expression(object):
     
@@ -20,7 +24,7 @@ class Expression(object):
 
     @abc.abstractmethod
     def can_prune(self):
-        return
+        return True
 
 class Simple_expression(Expression):
     
@@ -28,7 +32,7 @@ class Simple_expression(Expression):
         self.table = from_table
         self.select_column = select_column
         self.where_clause = where_clause
-        
+  
     def evaluate(self, socket, starting_set):
         
         if starting_set == []:
@@ -46,10 +50,10 @@ class Simple_expression(Expression):
                 query += " AND %s IN (%s)" % (self.select_column, in_clause)
         
         return rows_as_list(socket.execute(query))
-    
+
     def to_string(self):
         return self.where_clause
-    
+
     def can_prune(self):
         return not any (op in self.where_clause for op in ["<", ">"])
     
@@ -58,7 +62,7 @@ class AND_expression(Expression):
     def __init__(self, left, right):
         self.left = left
         self.right = right
-        
+
     def evaluate(self, session, starting_set):
         
         if starting_set == []:
@@ -73,11 +77,11 @@ class AND_expression(Expression):
         else:
             temp = self.left.evaluate(session, starting_set)
             return intersect(temp, self.right.evaluate(session, temp))
-    
+
     def to_string(self):
         res = "(" + self.left.to_string() + ")" + " AND " + "(" + self.right.to_string() + ")"
         return res
-    
+
     def can_prune(self):
         return True
     
@@ -86,18 +90,18 @@ class OR_expression(Expression):
     def __init__(self, left, right):
         self.left = left
         self.right = right
-        
+ 
     def evaluate(self, session, starting_set):
         
         if starting_set == []:
             return []
         
         return union(self.left.evaluate(session, starting_set), self.right.evaluate(session, starting_set))
-    
+
     def to_string(self):
         res = "(" + self.left.to_string() + ")" + " OR " + "(" + self.right.to_string() + ")"
         return res
-    
+
     def can_prune(self):
         return True
     
@@ -107,7 +111,7 @@ class NOT_expression(Expression):
         self.exp = exp
         self.table = table
         self.select_column = select_column
-        
+ 
     def evaluate(self, session, starting_set):
         
[…]

### 2015-03-28 gemini/update_query.sh

@@ -9,4 +9,12 @@ gemini query -q "select variant_id, chrom, sub_type from variants" \
 gemini query -q "select variant_id, (gt_types).(sex='1') from variants" \
 		--header test/my.db
 
-gemini query -q "select * from samples" --header test/my.db
+gemini query -q "select * from samples where sex = '1'" --header test/my.db
+
+gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<240].[any]" --header test/my.db
+
+gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<240].[none]" --cores 2 test/my.db
+
+gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<100].[all]" --cores 2 test/my.db
+
+gemini query -q "select variant_id from variants" --gt-filter "[gt_type].[*].[!=HET].[any]" --cores 2 test/my.db

### 2015-04-01 thesis/gemini_beschrijving.tex
tekst datamodel cassandra begin

@@
 \section{Inladen}
-%TODO ref nr IPython dink
 Het inladen van de data uit VCF-bestanden is een computationeel intensieve operatie, enerzijds omwille van de enorme grootte van deze bestanden, en anderzijds omdat in deze fase ook alle variants geannoteerd moeten worden.
 Om het proces te versnellen, biedt GEMINI de mogelijkheid het werk te paralleliseren door het VCF-bestand de comprimeren, op het bestaande bestand een index te defini\"eren en zo het werk te verdelen.
-Dit kan over meerdere processoren binnen 1 computer zijn, maar via de IPython-interface ook over volledige clusters van computers.\\
+Dit kan over meerdere processoren binnen 1 computer zijn, maar via de IPython-interface ook over volledige clusters van computers \cite{PER-GRA:2007}.\\
 
@@
 Enerzijds via gewone SQL-queries, maar omdat het onderzoeken van individuele genotypes van cruciaal belang is in het onderzoek naar ziektes en SQL standaard niet de individuele genotypes in array-kolommen kan opvragen, biedt GEMINI bovendien een verrijkte SQL-syntax.\\
-Zo laat het de gebruiker toe via filters en wildcards de gewenste genotype-eigenschappen en samples te defini\"eren.
+Zo laat het de gebruiker toe via filters en wildcards de gewenste genotype-eigenschappen en samples te specifi\"eren.
 
@@
 \subsection{Genotype filters/wildcards}
-Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL WHERE-clauses uitbreiden met zogenaamde genotype filters.
+Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL WHERE-clausules uitbreiden met zogenaamde genotype filters.
 Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor Jan heterozygoot is en Mieke alles behalve heterozygoot is, kan dit met:\\
@@
 
-Al de bovenstaande filters kunnen met elkaar gecombineerd worden, evenals met gewone WHERE-clauses op de andere kolommen van de variants-tabel.
+\noindent Al de bovenstaande filters kunnen met elkaar gecombineerd worden, evenals met gewone WHERE-clausules op de andere kolommen van de variants-tabel.
+
+\subsection{\texttt{--show-samples}}
+
+GEMINI biedt ook de mogelijkheid, bij een query op de \texttt{variants}-tabel, voor elke variant die aan de gestelde eisen voldoet, de namen van alle samples weer te geven die voor gegeven variant homozygoot of heterozygoot zijn.
+Dit gebeurt door de \texttt{--show-samples} flag mee te geven aan het \texttt{query} commando.

### 2015-04-02 thesis/datamodel_cassandra.tex
datamodel_cassandra tekst

\chapter{Datamodel Cassandra}
\label{cassandra_datamodel}

Het datamodel van Apache Cassandra vertoont enkele sterke verschillen met het relationele datamodel. Dit vereist enkele grondige aanpassingen aan het database schema van GEMINI. In deze sectie komen de belangrijkste eigenschappen van Cassandra aan bod, hun gevolgen voor de belangrijkste database-functionaliteiten, en de uitgevoerde aanpassingen aan het onderliggende schema van GEMINI ten op zichte van de SQLite implementatie.

\section{Datamodel Cassandra}

Zoals eerder vermeld bewaart Cassandra de cellen in een tabel als een 2 dimensionele map van enerzijds een per rij gedefinieerde primary key, en de naam van een kolom. De inhoud van cellen die niet in de primary key van een rij liggen, heeft voor Cassandra geen enkele betekenis.\\
Die primary key bestaat uit 2 delen: het eerste is de partition key, deze bestaat uit minstens 1 kolom en de waarde hiervan die via het consistent-hashing mechanisme bepaalt op welke knopen in het cluster de rij terechtkomt. Het tweede, optionele, deel is de clustering key, en bepaalt in welke volgorde rijen met dezelfde partition key op 1 node bewaard worden. Dit is standaard in oplopende volgorde.\\\\

%TODO: ALLOW FILTERING
Omdat het inspecteren van cellen indruist tegen de principes van Cassandra, zijn de query-mogelijkheden eerder beperkt: zonder het defini\"eren van indices is het in \texttt{WHERE}-clausules enkel mogelijk beperkingen op te leggen aan kolommen in de primary key. Hierbij moet er een gelijkheidsbeperking opgelegd worden aan de volledige partition key, en mogen er zowel gelijk- als ongelijkheidsbeperkingen opgelegd worden aan de kolommen in de clustering key, maar enkel op voorwaarde dat de voorgaande kolom in de clustering key ook met een gelijkheidsbeperking gespecifieerd is. Op deze manier kan Cassandra queries zeer snel uitvoeren door ze a.d.h.v. de hash van de partition key naar een juiste node te routeren, en vervolgens via de overige opgegeven kolommen de locatie van de juiste rijen te berekenen (die omwille van de clustering allemaal opelkaar volgend opgeslagen zijn). Dit gebeurt zonder de waarden van individuele cellen te bekijken, maar dus enkel door het berekenen van een simpele hashfunctie. Range-queries zijn dus enkel mogelijk op kolommen in de clustering key en het is niet mogelijk \texttt{!=}-beperkingen te gebruiken in \texttt{WHERE}-clausules. Bovendien laat Cassandra enkel toe beperkingen met elkaar te combineren via conjuncties, dus niet via \texttt{OR}- of \texttt{NOT}-operatoren. Een uitzondering op dit laatste is de \texttt{IN}-operator: op deze manier kan de gebruiker meegeven in welke set van waarden een bepaalde kolom moet liggen, maar dit is ook enkel mogelijk op de laatste kolom in de partition key of de laatste kolom in de clustering key (maar weer op voorwaarde dat de voorgaande kolommen reeds beperkt zijn).\\
Cassandra laat toe om indices op kolommen te defini\"eren, maar deze zijn niet bijzonder nuttig. Ze laten enkel gelijkheidsbeperkingen toe (dus geen range-queries) en bovendien raadt Datastax het gebruik van indices op kolommen met zowel een zeer lage als een zeer hoge kardinaliteit af, dit omdat in het eerste geval de index tabel zal bestaan uit zeer weinig zeer lange rijen voor elk van de ge\"indexeerde waarden en in het tweede geval Cassandra bij een query op de ge\"indexeerde kolom door zeer veel verscheidene waarden zal moeten zoeken om een klein aantal resultaten te vinden \cite{when_to_use_index}.

\section{Database schema GEMINI}
Zoals blijkt uit bovenstaande beschrijving van het datamodel van Cassandra, leent het systeem zich niet zomaar tot ad-hoc querying: het is niet mogelijk op een performante manier zomaar aan willekeurige kolommen in een tabel voorwaarden op te leggen en deze voorwaarden met elkaar te combineren. Dit betekent dat bij het ontwerpen van het database schema al rekening gehouden moet worden met de queries die achteraf op de data mogelijk moeten zijn.

### 2015-04-02 db_schemas/gemini_schemas.tex
dink
cassandra indices + begin dataschema

@@
 
-1 \texttt{variants} tabel met array columns.\\
+1 \texttt{variants} tabel met arrays (binary blob in SQLite, array columns in PostgreSQL).\\
 
-\texttt{SELECT * \\FROM variants \\WHERE (gts.brecht == hom\_ref \&\& gts.aaron == hom\_alt)}
+\texttt{SELECT * \\FROM variants \\WHERE (gts.bruce == hom\_ref \&\& gts.bob == hom\_alt)}
 \begin{table}[h]
@@
 
-1 \texttt{variants} tabel zonder array columns, met kolom voor elke \texttt{individual}.\\
+1 \texttt{variants} tabel zonder array columns, met kolom voor elke \texttt{sample}.\\
 
@@
 FROM variants \\
-WHERE brecht == hom\_ref \&\& aaron == hom\_alt}
+WHERE bruce == hom\_ref \&\& bob == hom\_alt}
 \begin{table}[!htbp]
@@
 \toprule
-variant\_id & ref & alt & \ldots & aaron & brecht & \ldots \\ \bottomrule
+variant\_id & ref & alt & \ldots & bob & bruce & \ldots \\ \bottomrule
 \end{tabular}
@@
 \begin{table}[!htbp]
-\begin{tabular}{@{}|l|l|l|l|@{}}
+\begin{tabular}{@{}|l|l|l|l|l|@{}}
 \toprule
-variant\_id & sample & genotype & \ldots \\ \bottomrule
+variant\_id & sample\_name & genotype & gt\_depth & \ldots \\ \bottomrule
 \end{tabular}
@@
 \texttt{SELECT *}
-\\\texttt{FROM aaron JOIN (}\=\texttt{SELECT variant\_id}
-\\ \> \texttt{FROM (brecht JOIN variants)}
-\\ \> \texttt{WHERE hom\_ref) }
-\\ \texttt{WHERE hom\_alt }
+\\\texttt{FROM bob JOIN (}\=\texttt{SELECT variant\_id}
+\\ \> \texttt{FROM (bruce JOIN variants)}
+\\ \> \texttt{WHERE genotype = hom\_ref) }
+\\ \texttt{WHERE genotype = hom\_alt }
 \end{tabbing}
@@
 \newpage
-\subsection{\texttt{genotypes} tabel}
+\subsection{\texttt{genotype}-eigenschap tabellen}
 
@@
 
-1 \texttt{genotypes} tabel met rij voor elke \texttt{(variant, sample, genotype)}
+1 \texttt{variants\_by\_samples\_gt\_type} tabel met rij voor elke \texttt{(variant, sample, genotype)}\\ Met als primary key \texttt{((sample\_name, gt\_type), (variant\_id))}\\
+
+1 \texttt{variants\_by\_samples\_gt\_depth} tabel met rij voor elke \texttt{(variant, sample, gt\_depth)}\\ Met als primary key \texttt{((sample\_name), (depth, variant\_id))}\\
 
@@
 \toprule
-variant\_id & sample & genotype & \color{red} individual\_id & \ldots \\ \bottomrule
+variant\_id & sample\_name & genotype & \ldots \\ \bottomrule
+\end{tabular}\\
+\begin{tabular}{@{}|l|l|l|l|l|@{}}
+\toprule
+variant\_id & sample\_name & gt\_depth & \ldots \\ \bottomrule
 \end{tabular}
 \end{table}
-
-Query onmogelijk zonder \texttt{individual\_id}. Anders:\\
+En vergelijkbare tabellen voor de andere genotype eigenschappen van de samples.
 
@@
 \\ \>\>\texttt{FROM genotypes }
-\\ \>\>\texttt{WHERE individual\_id == brecht)) }
+\\ \>\>\texttt{WHERE sample\_name == bruce)) }
 \\ \> \texttt{WHERE hom\_ref) JOIN (}\=\= \texttt{SELECT * }
 					\\ \>\>\> \texttt{FROM genotypes }
-					\\ \>\>\> \texttt{WHERE individual\_id == aaron))}
-\\ \texttt{WHERE hom\_alt}
+					\\ \>\>\> \texttt{WHERE sample\_name == bob))}
+\\ \texttt{WHERE genotype = hom\_alt}
 

### 2015-04-02 thesis/datamodel_cassandra.tex
cassandra indices + begin dataschema

@@
 %TODO: ALLOW FILTERING
-Omdat het inspecteren van cellen indruist tegen de principes van Cassandra, zijn de query-mogelijkheden eerder beperkt: zonder het defini\"eren van indices is het in \texttt{WHERE}-clausules enkel mogelijk beperkingen op te leggen aan kolommen in de primary key.
-Hierbij moet er een gelijkheidsbeperking opgelegd worden aan de volledige partition key, en mogen er zowel gelijk- als ongelijkheidsbeperkingen opgelegd worden aan de kolommen in de clustering key, maar enkel op voorwaarde dat de voorgaande kolom in de clustering key ook met een gelijkheidsbeperking gespecifieerd is.
+Omdat het inspecteren van cellen indruist tegen de principes van Cassandra, zijn de query-mogelijkheden eerder beperkt: zonder het defini\"eren van indices is het in \texttt{WHERE}-clausules enkel mogelijk beperkingen op te leggen aan kolommen in de primary key, en dan nog zo dat de bedoelde rijen binnen 1 partitie liggen, en opeenvolgend opgeslagen zijn.
+Daarom moet er een gelijkheidsbeperking opgelegd worden aan de volledige partition key, en mogen er zowel gelijk- als ongelijkheidsbeperkingen opgelegd worden aan de kolommen in de clustering key, maar enkel op voorwaarde dat de voorgaande kolom in de clustering key ook met een gelijkheidsbeperking gespecifieerd is.
 Op deze manier kan Cassandra queries zeer snel uitvoeren door ze a.d.h.v.
@@
 Dit betekent dat bij het ontwerpen van het database schema al rekening gehouden moet worden met de queries die achteraf op de data mogelijk moeten zijn.
+Omdat het soort queries dat een tabel ondersteunt sterk samenhangt met de keuze van de primary key van de tabel en GEMINI meerdere, uiteenlopende soorten queries op elke tabel vereist, zal dit onvermijdelijk leiden tot duplicatie van data.\\\\
+
+Het belangrijkste vraagstuk is hoe de genotype-kolommen uit het oorspronkelijke relationele model in Cassandra op te slaan. Hier zijn enkele opties voor:
+
+\begin{itemize}
+
+\item[Collection columns] Cassandra biedt zogenaamde collection types, zoals sets, lists of maps.
+Dit is vergelijkbaar met de bestaande implementatie in SQLite (buiten dat ze in Cassandra niet als binary blobs bewaard worden).
+Het nadeel is echter dat deze collections niet meer dan 65536 ($2^{16}$) entries kunnen bevatten, wat het hele nut van de migratie naar Cassandra zou ondermijnen.
+
+\item[Super-\texttt{variants}-tabel] Een andere mogelijkheid is de \texttt{variants}-tabel uit te breiden met een kolom voor elke genotype-eigenschap van elke sample.
+Deze aanpak heeft als voornaamste voordeel dat de kolommen met genotype-eigenschappen van specifieke samples zonder omwegen uit de \texttt{variants}-tabel gehaald kunnen worden.
+Dit is vooral van belang voor de in (\ref{gemini_beschrijving}) beschreven sample-filters.
+Bovendien kan Cassandra tot 2 miljard cellen opslaan op 1 partitie, dus vormt het grote aantal kolommen dat dit model met zich meebrengt, geen probleem.\\
+Het nadeel is echter dat het onmogelijk is ad-hoc queries te defini\"eren op genotypes van willekeurige eigenschappen zonder het gebruik van secundaire indices.
+Elke query die in een \texttt{WHERE}-clausule andere kolommen of samples betrekt, vereist om de hierboven beschreven redenen een andere keuze van de primary key om effici\"ent de juiste rijen te kunnen opzoeken.
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|l|l|l|l|l|l|l|l|@{}}
+\toprule
+variant\_id & ref & alt & \ldots & gt\_type\_bob & gt\_type\_bruce & \ldots & gt\_depth\_bob & gt\_depth\_bruce & \ldots \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\end{itemize}

### 2015-04-02 thesis/gemini_beschrijving.tex
cassandra indices + begin dataschema

@@
 \chapter{GEMINI overzicht}
-\label{cha3}
+\label{gemini_beschrijving}
 
@@
 Wanneer de gebruiker ge\"interesseerd is in bijvoorbeeld het genotype van Jan, wordt dit:\\
-\texttt{SELECT gt\_types.Jan FROM variants}\\\\
+\texttt{SELECT gt\_types.Bob FROM variants}\\\\
 Als de gebruiker de diepte van de genotypes alle vrouwelijke samples wil zien, wordt dit:\\\\
@@
 Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL WHERE-clausules uitbreiden met zogenaamde genotype filters.
-Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor Jan heterozygoot is en Mieke alles behalve heterozygoot is, kan dit met:\\
+Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor Bob heterozygoot is en Bruce alles behalve heterozygoot is, kan dit met:\\
 
-\noindent\texttt{\$ gemini query -q} "\texttt{SELECT * FROM variants}"\textbackslash \\\texttt{--gt-filter }"\texttt{gt\_types.Jan == HET and gt\_types.Mieke != HET}"\\
+\noindent\texttt{\$ gemini query -q} "\texttt{SELECT * FROM variants}"\textbackslash \\\texttt{--gt-filter }"\texttt{gt\_types.Bob == HET and gt\_types.Bruce != HET}"\\
 

### 2015-04-03 thesis/datamodel_cassandra.tex
datamodel cassandra details

@@
 Dit betekent dat bij het ontwerpen van het database schema al rekening gehouden moet worden met de queries die achteraf op de data mogelijk moeten zijn.
-Omdat het soort queries dat een tabel ondersteunt sterk samenhangt met de keuze van de primary key van de tabel en GEMINI meerdere, uiteenlopende soorten queries op elke tabel vereist, zal dit onvermijdelijk leiden tot duplicatie van data.\\\\
+Omdat het soort queries dat een tabel ondersteunt sterk samenhangt met de keuze van de primary key van de tabel en GEMINI meerdere, uiteenlopende soorten queries op elke tabel vereist, zal dit onvermijdelijk leiden tot duplicatie van data.
+
+\subsection{\texttt{variants}-tabel vs. genotype-informatie}
 
@@
 
-\item[Collection columns] Cassandra biedt zogenaamde collection types, zoals sets, lists of maps.
+\item \textbf{Collection columns} Cassandra biedt zogenaamde collection types, zoals sets, lists of maps.
 Dit is vergelijkbaar met de bestaande implementatie in SQLite (buiten dat ze in Cassandra niet als binary blobs bewaard worden).
@@
 
-\item[Super-\texttt{variants}-tabel] Een andere mogelijkheid is de \texttt{variants}-tabel uit te breiden met een kolom voor elke genotype-eigenschap van elke sample.
+\item \textbf{Super-\texttt{variants}-tabel} Een andere mogelijkheid is de \texttt{variants}-tabel uit te breiden met een kolom voor elke genotype-eigenschap van elke sample.
 Deze aanpak heeft als voornaamste voordeel dat de kolommen met genotype-eigenschappen van specifieke samples zonder omwegen uit de \texttt{variants}-tabel gehaald kunnen worden.
@@
 
+\item \textbf{\texttt{genotype}-tabellen} Een derde optie is een \texttt{variants}-tabel zonder genotype-informatie, gecombineerd met een tabel voor elke eigenschap van de genotypes van de samples, met een rij voor elke \texttt{(variant, sample)}.
+Bijvoorbeeld:\\
+
+\begin{itemize}
+
+\item Een \texttt{variants\_by\_samples\_gt}-tabel met als primary key \texttt{((sample\_name, gt\_type), (variant\_id))}.
+De primary key is zo gekozen dat alle variants waarvoor een sample eenzelfde genotype heeft, bij elkaar op 1 node liggen, zodat deze gemakkelijk opgevraagd kunnen worden.
+Hetzelfde was mogelijk geweest met enkel \texttt{sample\_name} als partition key, maar met de bovenstaande, granulairdere partition key zijn de variants beter over het cluster verspreid.
+Bovendien houdt het vanuit een semantisch oogpunt geen steek range queries uit te voeren op de \texttt{gt\_type}-kolom, dus moet deze niet per se in de clustering key staan.
+De \texttt{variant\_id}-kolom ten slotte maakt enkel deel uit van de primary key om deze uniek te maken voor elk tupel \texttt{(variant, sample, genotype)}.\\
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|l|@{}}
+\toprule
+variant\_id & sample\_name & genotype \\ \bottomrule
+\end{tabular}\\
+\end{table}
+
+\item Een \texttt{variants\_by\_samples\_gt\_depth}-tabel met als primary key \texttt{((sample\_name), (gt\_depth, variant\_id))}.
+Omdat range queries op de \texttt{gt\_depth}-kolom wel een vereiste zijn, moeten alle variants voor eenzelfde sample volgens gt\_depth geclusterd, en dus gerangschikt staan.
+Vandaar dat de \texttt{gt\_depth}-kolom in dit geval in de clustering key, en niet in de partition key staat.
+Voor de rest is de keuze van de primary key volledig analoog met die in de \texttt{variants\_by\_samples\_gt}-tabel.
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|l|@{}}
+\toprule
+variant\_id & sample\_name & gt\_depth \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\item Vergelijkbare tabellen voor de overige genotype-eigenschappen.
+
 \end{itemize}
+
+Het grote voordeel van deze aanpak is dat het, dankzij de keuze van de primary key, zeer eenvoudig is variants op te vragen waarvoor het genotype van \'e\'en sample aan specifieke gelijkheids- of ongelijkheidsvoorwaarden voldoet.
+Het nadeel van dit schema is drieledig: ten eerste scheidt het de informatie over de genotypes van samples van de andere informatie over de variants.
+Dit betekent dat, om deze samen weer te geven, een \texttt{JOIN} van de \texttt{variants}-tabel met een \texttt{genotype}-tabel nodig is, en Cassandra biedt zoals geweten geen \texttt{JOIN}.
+Ten tweede is het onmogelijk om met een sample-filter voor meerdere samples de genotypes tegelijkertijd op te vragen, en ten laatste is het onmogelijk in een query voorwaarden op te leggen aan de genotypes van meerdere samples.
+
+\end{itemize}
+
+De uiteindelijke keuze is gevallen op een combinatie van de tweede en derde optie.
+Dit om de volledige oorspronkelijke functionaliteit van GEMINI te kunnen bieden: enerzijds het opvragen van de genotype-informatie van willekeurige samples door middel van de sample-filters, en anderzijds het query'en van de \texttt{variants}-tabel met, dankzij gt-filters en -wildcards, beperkingen op de genotypes van specifiek gekozen samples.
+Zoals hierboven beschreven is dit laatste niet voor de hand liggend, en zal door het ontbreken van \texttt{JOIN}s en de layout van de \texttt{genotype}-tabellen het combineren van beperkingen op de genotypes van meerdere samples nog extra aanpassingen vergen.
+Hier komt sectie \ref{??} uitgebreid op terug.
+%TODO
+
+\subsection{\texttt{variants-, samples-}tabel vs. arbitraire queries}
+
+Een tweede belangrijke vraagstuk is hoe de \texttt{variants-} en \texttt{samples-}tabellen zo te ontwerpen dat ze ook op andere, arbitraire kolommen effici\"ent doorzoekbaar zijn.\\
+Om queries op kolommen te ondersteunen die logisch gezien enkel met gelijkheden beperkt zullen worden, zoals \texttt{chrom} of respectievelijk \texttt{sex}, zou het in theorie volstaan op deze kolommen een secundaire index te defini\"eren.
+Dit is eenvoudig bij de creatie van de tabel, veroorzaakt geen duplicatie van data en is zeer rechttoe-rechtaan bij het query'en, maar heeft zoals hierboven vermeld onvoorspelbare performantie afhankelijk van de kardinaliteit van de kolommen.
+Bovendien werkt dit mechanisme niet voor kolommen die vaker met ongelijkheidsbeperkingen gequeried zullen worden, zoals \texttt{depth}, \texttt{start}, of in het geval van de \texttt{samples}-tabel (hypothetisch, deze tabellen zitten niet standaard in GEMINI) leeftijd of generatie.\\
+
+Een aanpak die in beide gevallen op een voorspelbare en betrouwbare manier werkt, is om net als voor de genotype-informatie extra tabellen te defini\"eren, met een specifiek gekozen primary key die de gewenste query mogelijk maakt.
+Deze oplossing zorgt natuurlijk voor veel gedupliceerde data, maar heeft als voordeel dat ze \'e\'en coherente en performante aanpak van het probleem mogelijk maakt.
+Bij het opstellen van deze tabellen moet goed overwogen worden welke vaakgebruikte queries een eigen tabel vereisen en verdienen.
+Minder frequente, uitgebreidere queries, kunnen mits een goede keuze van de basistabellen immers gesplitst worden in subqueries op deze basistabellen.
+Hetzelfde mechanisme dat voor de gt-filters en -wildcards uitgebreide queries zal uitvoeren is ook hier van toepassing.
+Dit zal natuurlijk minder effici\"ent zijn dan een aangepaste tabel voor de uitgebreide, originele query, maar laat toe de data-duplicatie enigszins binnen de perken te houden.
+Voorbeelden van zulke basistabellen zijn:
+
+\begin{itemize}
+\item Een \texttt{variants\_by\_chrom\_start}-tabel, die het eenvoudig maakt alle varianten op te zoeken die op een bepaald chromosoom binnen een range van startposities liggen.
+De primary key is in dit geval: \texttt{((chrom), (start, variant\_id))}.
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|l|@{}}
+\toprule
+ chrom & start & variant\_id \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\item Een \texttt{variants\_by\_gene}-tabel, die het eenvoudig maakt alle varianten binnen \'e\'en gen op te zoeken. De primary key is in dit geval: \texttt{((gene), (variant\_id))}.
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|@{}}
+\toprule
+ gene & variant\_id \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\item Een \texttt{samples\_by\_sex}-tabel, die het eenvoudig maakt alle samples van een bepaald geslacht op te vragen. De primary key is in dit geval: \texttt{((sex), (sample\_name))}.
+
+\begin{table}[!htbp]
+\begin{tabular}{@{}|l|l|@{}}
+\toprule
+ sex & sample\_name \\ \bottomrule
+\end{tabular}
+\end{table}
+\end{itemize}
+
+Ten slotte is het ook nog interessant een feature aan de GEMINI command line interface toe te voegen waarmee de gebruiker zelf nieuwe basistabellen kan defini\"eren wanneer hij dit nodig acht.

### 2015-04-06 gemini/gemini/GeminiQuery.py
unit tests for loading and querying

@@ -425,10 +425,11 @@ class GeminiQuery(object):
             print row, gts[idx]
     """
 
-    def __init__(self, db_contact_points, include_gt_cols=False,
+    def __init__(self, db_contact_points, keyspace, include_gt_cols=False,
                  out_format=DefaultRowFormat(None)):
 
         self.db_contact_points = map(strip, db_contact_points.split(','))
+        self.keyspace = keyspace
         self.query_executed = False
         self.for_browser = False
         self.include_gt_cols = include_gt_cols
@@ -454,7 +455,7 @@ class GeminiQuery(object):
     def run(self, query, gt_filter=None, show_variant_samples=False,
             variant_samples_delim=',', predicates=None,
             needs_genotypes=False, needs_genes=False,
-            show_families=False, nr_cores = 1):
+            show_families=False, sort_results=False, nr_cores = 1):
         """
         Execute a query against a Gemini database. The user may
         specify:
@@ -462,9 +463,9 @@ class GeminiQuery(object):
             1. (reqd.) an SQL `query`.
             2. (opt.) a genotype filter.
         """
-        self.query = self.formatter.format_query(query)
+        self.query = self.formatter.format_query(query).replace('==','=')
         self.gt_filter = gt_filter
-        print self.query + '; gt-filter = %s' % gt_filter
+        #print self.query + '; gt-filter = %s \n' % gt_filter
         self.nr_cores = nr_cores
         if self._is_gt_filter_safe() is False:
             sys.exit("ERROR: unsafe --gt-filter command.")
@@ -473,7 +474,7 @@ class GeminiQuery(object):
         
         self.show_variant_samples = show_variant_samples
         self.variant_samples_delim = variant_samples_delim
-
+        self.sort_results = sort_results
         self.needs_genotypes = needs_genotypes
         self.needs_vcf_columns = False
         if self.formatter.name == 'vcf':
@@ -619,7 +620,7 @@ class GeminiQuery(object):
         # open up a new database
         
         self.cluster = Cluster(self.db_contact_points)
-        self.session = self.cluster.connect('gemini_keyspace')
+        self.session = self.cluster.connect(self.keyspace)
 
     def _is_gt_filter_safe(self):
         """
@@ -669,8 +670,13 @@ class GeminiQuery(object):
                         query += " WHERE %s IN ('%s')" % (self.get_partition_key(self.from_table), in_clause)
                 query += " " + self.rest_of_query
                 self.session.row_factory = ordered_dict_factory
[…]

### 2015-04-06 gemini/gemini/database_cassandra.py

@@ -93,7 +93,7 @@ def create_tables(session, gt_column_names, extra_sample_columns):
                         sample_name text, \
                         gt_depth int, \
                         variant_id int, \
-                        primary key (sample_name, gt_depth, variant_id))'''))    
+                        primary key (sample_name, gt_depth, variant_id))'''))  
     
     session.execute(SimpleStatement('''CREATE TABLE IF NOT EXISTS samples_by_variants_gt_type ( \
                     variant_id int, \
@@ -108,6 +108,12 @@ def create_tables(session, gt_column_names, extra_sample_columns):
                         sub_type text, \
                         call_rate float, \
                         PRIMARY KEY (sub_type, call_rate, variant_id))'''))
+     
+    session.execute(SimpleStatement('''CREATE TABLE if not exists variants_by_chrom_start( \
+                        chrom text,
+                        start int,
+                        variant_id int, \
+                        primary key (chrom, start, variant_id))'''))     
     
     session.execute(SimpleStatement('''CREATE TABLE if not exists variants_by_gene ( \
                         variant_id int, \

### 2015-04-06 gemini/gemini/gemini_load.py

@@ -140,8 +140,9 @@ def load_chunks_multicore(grabix_file, args):
     if args.skip_info_string is True:
         skip_info_string = "--skip-info-string"
         
-    contact_points = "-db \"" + args.contact_points + "\""
-
+    contact_points = "-db " + args.contact_points 
+    keyspace = "-ks " + args.keyspace
+    
     submit_command = "{cmd}"
     vcf, _ = os.path.splitext(grabix_file)
     chunk_steps = get_chunk_steps(grabix_file, args)
@@ -201,6 +202,7 @@ def load_chunks_ipython(grabix_file, args, view):
         passonly = "--passonly"
         
     contact_points = "-db \"" + args.contact_points + "\""
+    keyspace = "-ks \"" + args.keyspace + "\""
 
     skip_info_string = ""
     if args.skip_info_string is True:
@@ -223,7 +225,8 @@ def load_chunks_ipython(grabix_file, args, view):
                  "test_mode": test_mode,
                  "passonly": passonly,
                  "skip_info_string": skip_info_string,
-                 "contact_points": contact_points}    
+                 "contact_points": contact_points,
+                 "keyspace": keyspace}    
     chunk_dbs = view.map(load_chunk, chunk_steps, [load_args] * total_chunks)
 
     print "Done loading variants in {0} chunks.".format(total_chunks)
@@ -247,7 +250,7 @@ def gemini_pipe_load_cmd():
                        " {no_genotypes} {no_load_genotypes} {no_genotypes}"
                        " {skip_gerp_bp} {skip_gene_tables} {skip_cadd}"
                        " {passonly} {skip_info_string} {test_mode}"
-                       " -o {start} {contact_points}")
+                       " -o {start} {contact_points} {keyspace}")
     return " | ".join([grabix_cmd, gemini_load_cmd])
 
 def get_chunk_steps(grabix_file, args):

### 2015-04-06 gemini/gemini/gemini_load_chunk.py

@@ -43,11 +43,15 @@ class GeminiLoader(object):
         
         self.buffer_size = buffer_size
         self._get_anno_version()
-        self.contact_points = map(strip, args.contact_points.split(','))   
+        self.contact_points = map(strip, args.contact_points.split(','))
+        self.keyspace = args.keyspace
+        
+        self.typed_gt_column_names = []
+        self.gt_column_names = []
         
         if not self.args.no_genotypes:
             self.samples = self.vcf_reader.samples
-            (self.gt_column_names, self.typed_gt_column_names) = self._get_typed_gt_column_names()
+            self.gt_column_names, self.typed_gt_column_names = self._get_typed_gt_column_names()
             
         NUM_BUILT_IN = 6
         self.extra_sample_columns = get_ped_fields(args.ped_file)[NUM_BUILT_IN:]
@@ -124,6 +128,7 @@ class GeminiLoader(object):
         self.var_impacts_buffer = blist([])
         self.var_subtypes_buffer = blist([])
         self.var_gene_buffer = blist([])
+        self.var_chrom_start_buffer = blist([])
         buffer_count = 0
         self.skipped = 0
         #extra_file, extraheader_file = gemini_annotate.get_extra_files(self.args.db)
@@ -144,7 +149,9 @@ class GeminiLoader(object):
             # add the core variant info to the variant buffer
             self.var_buffer.append(variant)
             self.var_subtypes_buffer.append([self.v_id, variant[11], variant[12]])
-            self.var_gene_buffer.append([self.v_id, variant[55]])
+            if variant[55] != None:
+                self.var_gene_buffer.append([self.v_id, variant[55]])
+            self.var_chrom_start_buffer.append([self.v_id, variant[0], variant[1]])
                 
             var_sample_gt_types_buffer = blist([])
             var_sample_gt_depths_buffer = blist([])
@@ -170,12 +177,14 @@ class GeminiLoader(object):
                 batch_insert(self.session, 'variants_by_sub_type_call_rate',\
                                                  get_column_names('variants_by_sub_type_call_rate'), self.var_subtypes_buffer)
                 batch_insert(self.session, 'variants_by_gene', ['variant_id', 'gene'], self.var_gene_buffer)
+                batch_insert(self.session, 'variants_by_chrom_start', ['variant_id', 'chrom', 'start'], self.var_chrom_start_buffer)
                     # binary.genotypes.append(var_buffer)
[…]

### 2015-04-06 gemini/gemini/gemini_main.py

@@ -73,6 +73,12 @@ def main():
     parser_load.add_argument('-db', dest='contact_points',
                              default = "127.0.0.1",
                              help='The IP adresses at which the Cassandra cluster is reachable.')
+    parser_load.add_argument('-ks', 
+                            dest='keyspace',
+                            default = 'gemini_keyspace',
+                            metavar = 'STRING',
+                            help="The Cassandra keyspace in which the data should be stored.")
+                             
     parser_load.add_argument('-v', dest='vcf',
                              help='The VCF file to be loaded.')
     parser_load.add_argument('-t', dest='anno_type',
@@ -162,6 +168,11 @@ def main():
     parser_loadchunk.add_argument('-db', dest='contact_points',
                              default = "127.0.0.1",
                              help='The IP adresses at which the Cassandra cluster is reachable.')
+    parser_loadchunk.add_argument('-ks', 
+                                 dest='keyspace',
+                                 default = 'gemini_keyspace',
+                                 metavar = 'STRING',
+                                 help="The Cassandra keyspace in which the data should be stored.")
     parser_loadchunk.add_argument('-v',
                                   dest='vcf',
                                   help='The VCF file to be loaded.')
@@ -234,6 +245,9 @@ def main():
     parser_query.add_argument('-db', dest='contact_points',
                              default = "127.0.0.1",
                              help='The IP adresses at which the Cassandra cluster is reachable.')
+    parser_query.add_argument('-ks', dest='keyspace',
+                             default = "gemini_keyspace",
+                             help='The Cassandra keyspace in which the data is stored.')
     parser_query.add_argument('-q',
             dest='query',
             metavar='QUERY_STR',
@@ -313,6 +327,11 @@ def main():
                               dest='dgidb',
                               action='store_true',
                               help='Request drug-gene interaction info from DGIdb.',
+                              default=False)
+    parser_query.add_argument('--test-mode',
+                              dest='testing',
+                              action='store_true',
+                              help='Sort variants by start, samples by sample_id. ONLY TO BE USED FOR UNIT TESTS',
                               default=False)
     def query_fn(parser, args):
         import gemini_query

### 2015-04-06 gemini/gemini/gemini_query.py

@@ -116,13 +116,14 @@ def run_query(args):
     formatter = select_formatter(args)
     genotypes_needed = needs_genotypes(args)
     gene_needed = needs_gene(args)
-    gq = GeminiQuery.GeminiQuery(args.contact_points, out_format=formatter)
+    gq = GeminiQuery.GeminiQuery(args.contact_points, args.keyspace, out_format=formatter)
     gq.run(args.query, args.gt_filter, args.show_variant_samples,
            args.sample_delim, predicates, genotypes_needed,
-           gene_needed, args.show_families, args.cores)
+           gene_needed, args.show_families, args.testing, args.cores)
 
     if args.use_header and gq.header:
         print gq.header
+        #print
 
     if not args.dgidb:
         for row in gq:
@@ -138,10 +139,10 @@ def run_query(args):
         dgidb_info = query_dgidb(genes)
 
         # rerun the query (the cursor is now consumed)
-        gq = GeminiQuery.GeminiQuery(args.contact_points, out_format=formatter)
+        gq = GeminiQuery.GeminiQuery(args.contact_points, args.keyspace, out_format=formatter)
         gq.run(args.query, args.gt_filter, args.show_variant_samples,
            args.sample_delim, predicates, genotypes_needed,
-           gene_needed, args.show_families, args.cores)
+           gene_needed, args.show_families, args.testing, args.cores)
 
         # report the query results with DGIdb info added at the end.
         for row in gq:
@@ -152,7 +153,7 @@ def query(parser, args):
     start_time = time.time()
     run_query(args)
     end_time = time.time()
-    print "query completed in %s s. \n" % (end_time - start_time)
+    #print "query completed in %s s. \n" % (end_time - start_time)
 
 if __name__ == "__main__":
     main()

### 2015-04-06 gemini/gemini/query_expressions.py

@@ -46,8 +46,7 @@ class Simple_expression(Expression):
                 query += " AND %s IN ('%s')" % (self.select_column, in_clause)
             else:
                 in_clause = ",".join(map(lambda x: str(x), starting_set))            
-                query += " AND %s IN (%s)" % (self.select_column, in_clause)
-        
+                query += " AND %s IN (%s)" % (self.select_column, in_clause)       
         return rows_as_set(socket.execute(query))
 
     def to_string(self):
@@ -129,7 +128,7 @@ class NOT_expression(Expression):
     
 class GT_wildcard_expression(Expression):
     
-    def __init__(self, column, wildcard_rule, rule_enforcement, sample_names, db_contact_points, cores_for_eval = 1):
+    def __init__(self, column, wildcard_rule, rule_enforcement, sample_names, db_contact_points, keyspace, cores_for_eval = 1):
         self.column = column
         self.wildcard_rule = wildcard_rule
         if rule_enforcement.startswith('count'):
@@ -140,6 +139,7 @@ class GT_wildcard_expression(Expression):
         self.names = sample_names
         self.nr_cores = cores_for_eval
         self.db_contact_points = db_contact_points
+        self.keyspace = keyspace
         
     def to_string(self):
         return "[%s].[%s].[%s].[%s]" % (self.column, "?", self.wildcard_rule, self.rule_enforcement)
@@ -185,7 +185,7 @@ class GT_wildcard_expression(Expression):
             parent_conn, child_conn = Pipe()
             conns.append(parent_conn)
             p = Process(target=eval(target_rule +'_query'), args=(child_conn, self.column, corrected_rule,\
-                                                                   correct_starting_set, self.db_contact_points))
+                                                                   correct_starting_set, self.db_contact_points, self.keyspace))
             procs.append(p)
             p.start()
             
@@ -251,10 +251,10 @@ def rows_as_set(rows):
 def rows_as_list(rows):
     return map(lambda x: x[0], rows)
  
-def all_query(conn, field, clause, initial_set, contact_points):
+def all_query(conn, field, clause, initial_set, contact_points, keyspace):
         
     cluster = Cluster(contact_points)
-    session = cluster.connect('gemini_keyspace')
+    session = cluster.connect(keyspace)
     
     names = conn.recv()
     
@@ -284,10 +284,10 @@ def all_query(conn, field, clause, initial_set, contact_points):
     conn.send(results)
     conn.close()
 
[…]

### 2015-04-06 gemini/gemini/sql_utils.py

@@ -50,7 +50,7 @@ def get_query_parts(query):
     
     where_end = len(query)
     
-    for keyword in ["order_by", "limit", "allow_filtering"]:
+    for keyword in ["order by", "limit", "allow_filtering"]:
         stop = query.find(keyword)
         if stop > -1:
             from_end = min(stop, from_end)

### 2015-04-06 gemini/master_test_cassandra.sh

@@ -1,0 +1,17 @@
+check()
+{
+        if diff $1 $2; then
+        echo ok
+        else
+        echo fail
+        fi
+}
+export -f check
+
+cd test
+
+bash cassandra-test-setup.sh
+
+bash cassandra-test-query.sh
+
+bash test-load.sh

### 2015-04-06 gemini/test/cassandra-test-query.sh

@@ -1,0 +1,337 @@
+
+####################################################################
+# 1. Test the samples table
+####################################################################
+echo "    query.t01...\c"
+echo "1094PC0005	1	0	0	0	-9	-9
+1094PC0009	2	0	0	0	-9	-9
+1094PC0012	3	0	0	0	-9	-9
+1094PC0013	4	0	0	0	-9	-9
+1094PC0016	5	0	0	0	-9	-9
+1094PC0017	6	0	0	0	-9	-9
+1094PC0018	7	0	0	0	-9	-9
+1094PC0019	8	0	0	0	-9	-9
+1094PC0020	9	0	0	0	-9	-9
+1094PC0021	10	0	0	0	-9	-9" > exp
+gemini query -q "select * from samples" --test-mode -ks test_query_db | head > obs
+check obs exp
+rm obs exp
+
+####################################################################
+# 2. Test a basic query of the variants table
+####################################################################
+echo "    query.t02...\c"
+echo "chr1	30547	30548	T	G
+chr1	30859	30860	G	C
+chr1	30866	30869	CCT	C
+chr1	30894	30895	T	C
+chr1	30922	30923	G	T
+chr1	69269	69270	A	G
+chr1	69427	69428	T	G
+chr1	69510	69511	A	G
+chr1	69760	69761	A	T
+chr1	69870	69871	G	A" > exp
+gemini query -q "select chrom, start, end, ref, alt from variants" --test-mode -ks test_query_db | head \
+       > obs
+check obs exp
+rm obs exp
+
+####################################################################
+# 3. Test a basic query of the variants table with a where clause
+####################################################################
+echo "    query.t03...\c"
+echo "chr1	1219381	1219382	C	G	SCNN1D
+chr1	1219476	1219477	T	G	SCNN1D
+chr1	1219486	1219487	T	G	SCNN1D
+chr1	1219488	1219489	A	G	SCNN1D
+chr1	1219494	1219496	GT	G	SCNN1D
+chr1	1219502	1219505	GTT	G	SCNN1D
+chr1	1219507	1219511	GTGA	G	SCNN1D
+chr1	1219521	1219524	GTC	G	SCNN1D
+chr1	1219533	1219536	GTT	G	SCNN1D
+chr1	1219555	1219558	GTT	G	SCNN1D" > exp
+gemini query -q "select chrom, start, end, ref, alt, gene \
+                 from variants \
[…]

### 2015-04-06 gemini/test/cassandra-test-setup.sh

@@ -1,0 +1,2 @@
+gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_query_db --cores 2 
+

### 2015-04-06 gemini/test/test-load.sh

@@ -1,8 +1,9 @@
+
 ###########################################################################################
 #1. Test loading an unannotated file without genotypes
 ###########################################################################################
 gemini load -v ALL.wgs.phase1_release_v3.20101123.snps_indels_sv.sites.snippet.vcf \
-	        --skip-gene-tables --skip-gerp-bp --skip-cadd --no-genotypes 1000G.snippet.db
+	        --skip-gene-tables --skip-gerp-bp --skip-cadd --no-genotypes -ks thousand_g_snippet_db
 
 echo "    load.t1...\c"
 echo "chr1	10582	G	A	None
@@ -16,8 +17,8 @@ chr1	46401	C	CTGT	None
 chr1	47189	G	GA	None
 chr1	51475	T	C	None" > exp
 
-gemini query -q "select chrom, start, ref, alt, gene from variants limit 10" \
-	1000G.snippet.db > obs
+gemini query -q "select chrom, start, ref, alt, gene from variants" \
+	-ks thousand_g_snippet_db --test-mode | head > obs
 check obs exp
 rm obs exp
 
@@ -27,7 +28,7 @@ rm obs exp
 ###########################################################################################
 gemini load -v ALL.wgs.phase1_release_v3.20101123.snps_indels_sv.sites.snippet.snpEff.vcf \
     --skip-gene-tables --skip-gerp-bp --skip-cadd --no-genotypes \
-    -t snpEff 1000G.snippet.snpEff.db
+    -t snpEff -ks thousand_g_snippet_snpeff_db
 
 echo "    load.t2...\c"
 echo "chr1	10582	G	A	WASH7P
@@ -41,8 +42,8 @@ chr1	46401	C	CTGT	None
 chr1	47189	G	GA	None
 chr1	51475	T	C	None" > exp
 
-gemini query -q "select chrom, start, ref, alt, gene from variants limit 10" \
-	1000G.snippet.snpEff.db > obs
+gemini query -q "select chrom, start, ref, alt, gene from variants" \
+	-ks thousand_g_snippet_snpeff_db --test-mode | head > obs
 check obs exp
 rm obs exp
 
@@ -50,15 +51,15 @@ rm obs exp
 #3. Test loading an extended ped file
 ###########################################################################################
 gemini load -p test_extended_ped.ped -v test4.vep.snpeff.vcf \
[…]

### 2015-04-06 gemini/update_load.sh

@@ -4,7 +4,7 @@ sudo python setup.py install
 
 clear
 
-#python droptables.py
-#gemini load --skip-gene-tables --test-mode -v test/test.query.vcf --cores 5 --skip-cadd --skip-gerp-bp -t snpEff
+python droptables.py
+#gemini load --skip-gene-tables --test-mode -ks 'test_query' -v test/test.query.vcf --cores 4 --skip-cadd --skip-gerp-bp -t snpEff
 
-gemini load --skip-gene-tables --test-mode -v test/test.comp_het.vcf --cores 3 --skip-gerp-bp --skip-cadd -t snpEff -p test/test.comp_het.ped
+gemini load --skip-gene-tables --test-mode -ks load_test_keyspace -v test/test.comp_het.vcf --cores 3 --skip-gerp-bp --skip-cadd -t snpEff -p test/test.comp_het.ped

### 2015-04-06 gemini/update_query.sh

@@ -6,19 +6,19 @@ clear
 
 gemini query -q "select variant_id, chrom, sub_type from variants" \
 		--gt-filter "[gt_depth].[sex='1'].[>100].[all] && gt_type.child_1 = HET" \
-		--header --show-samples
+		--header --show-samples -ks load_test_keyspace
 
 gemini query -q "select variant_id, (gt_types).(sex='1') from variants" \
-		--header
+		--header -ks load_test_keyspace
 
-gemini query -q "select * from samples where sex = '1'" --header
+#gemini query -q "select * from samples where sex = '1'" --header
 
-gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<240].[any]" --header
+#gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<240].[any]" --header
 
-gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<240].[none]" --cores 2
+#gemini query -q "select variant_id from variants" --header --gt-filter "[gt_depth].[*].[<240].[none]" --cores 2
 
-gemini query -q "select variant_id from variants" --gt-filter "[gt_depth].[*].[<100].[count < 3]" --header --cores 1
+#gemini query -q "select variant_id from variants" --header --gt-filter "[gt_depth].[*].[<100].[count < 3]" --header --cores 1
 
-gemini query -q "select variant_id from variants" --gt-filter "[gt_type].[*].[=HET].[count <= 2]"
+#gemini query -q "select variant_id from variants" --header --gt-filter "[gt_type].[*].[=HET].[count <= 2]"
 
-gemini query -q "select variant_id from variants" --gt-filter "[gt_type].[*].[!=HET].[any]" --cores 2
+#gemini query -q "select variant_id from variants" --header --gt-filter "[gt_type].[*].[!=HET].[any]" --cores 2

### 2015-04-06 thesis/NoSQL_overzicht.tex
nosql methodologie + vergelijkende studie

@@
 \begin{itemize}
-\item[Key-Value stores] Deze zijn vergelijkbaar met dictionaries en mappen unieke keys op values.
+\item \textbf{Key-Value stores} Deze zijn vergelijkbaar met dictionaries en mappen unieke keys op values.
 Deze values zijn voor de databank volledig betekenisloze byte-arrays en de enige manier om ze op te vragen, is via de bijhoren key.
@@
 
-\item[Columnar stores] Deze zijn gebaseerd op het datamodel dat Google's BigTable heeft ge\"introduceerd en slaan data op in een spaarse, gedistribueerde, persistente en multidimensionele gesorteerde map\cite{chang2008bigtable}.
+\item \textbf{Columnar stores} Deze zijn gebaseerd op het datamodel dat Google's BigTable heeft ge\"introduceerd en slaan data op in een spaarse, gedistribueerde, persistente en multidimensionele gesorteerde map\cite{chang2008bigtable}.
 In het geval van BigTable zijn dit drie dimensies: row key, column key en een timestamp.
@@
 
-\item[Document stores] Deze bewaren data als key-value paren en encapsuleren deze in documenten.
+\item \textbf{Document stores} Deze bewaren data als key-value paren en encapsuleren deze in documenten.
 Values kunnen van een brede waaier aan types zijn, zoals geneste documenten, lijsten of scalars.
@@
 
-\item[Graph databases] Zoals de naam doet vermoeden, stammen deze uit grafentheorie en maken ze gebruik van grafen als datamodel.
+\item \textbf{Graph databases} Zoals de naam doet vermoeden, stammen deze uit grafentheorie en maken ze gebruik van grafen als datamodel.
 Ze zijn uitermate geschikt om sterk verweven data te beheren, van bronnen zoals sociale netwerken of location based services.

### 2015-04-06 thesis/gemini_beschrijving.tex
nosql methodologie + vergelijkende studie

@@
 
-\section{Datamodel}
+\section{Database schema}
 
@@
 Deze informatie komt in het eerste geval eveneens uit de annotatiebestanden, en in het tweede uit tekstbestanden met referentie-informatie over het menselijk genoom die GEMINI, indien gewenst mee inlaadt.\\
-Ten laatste zijn er nog enkele kleine tabellen met meta-informatie, zoals de \texttt{resources}-tabel die de gebruikte annotatie-files bevat, en de \texttt{version}-tabel die bevat doorn welke versie van GEMINI de data ingeladen is.\\\\
+Ten laatste zijn er nog enkele kleine tabellen met meta-informatie, zoals de \texttt{resources}-tabel die de gebruikte annotatie-files bevat, en de \texttt{version}-tabel die bevat door welke versie van GEMINI de data ingeladen is.\\\\
 
@@
 \subsection{Genotype filters/wildcards}
-Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL WHERE-clausules uitbreiden met zogenaamde genotype filters.
-Is hij bij voorbeeld enkel ge\"interesseerd in variants waarvoor Bob heterozygoot is en Bruce alles behalve heterozygoot is, kan dit met:\\
+Om beperkingen op te leggen aan de variants waarin hij ge\"interesseerd is, kan de gebruiker SQL \texttt{WHERE}-clausules uitbreiden met zogenaamde genotype filters.
+Is de gebruiker bij voorbeeld enkel ge\"interesseerd in variants waarvoor Bob heterozygoot is en Bruce alles behalve heterozygoot is, kan dit met:\\
 

### 2015-04-06 thesis/nosql_survey.tex
nosql methodologie + vergelijkende studie

\chapter{NoSQL: vergelijkende studie}
\label{nosql_survey}

\section{Methodologie}

Omwille van het enorme aanbod aan NoSQL en NewSQL systemen was een exhaustieve studie niet haalbaar. Deze vergelijkende studie behandelt de populairste datastores in een aantal relevante categorie\"en, namelijk document stores, columnar stores en NewSQL stores. Key-value stores en graph databases komen niet aan bod, aangezien hun datamodellen niet geschikt zijn voor de toepassing in kwestie. De uiteindelijke selectie werd gemaakt volgens criteria vergelijkbaar met die in \cite{grolinger2013data}, met de ranking van DB-Engine Ranking \cite{db_engine_rank} als maat voor de populariteit.\\

Deze ranglijst tracht populariteit te meten op basis van enkele parameters, zoals aantal vermeldingen op websites, algemene interesses volgens Google Trends, frequentie van technische discussies op fora zoals StackOverflow, vacatures i.v.m. de technologie en vermeldingen in professionele profielen op sites zoals LinkedIn. De resulterende selectie bestaat uit de document stores MongoDB en CouchBase Server, wide columnar stores Cassandra en HBase en NewSQL database VoltDB. Er bestaat al een uitbreiding van de DNA sequencing pijplijn die het ExaScience Life Lab gebruikt om MongoDB  databanken als in- en/of uitvoer te gebruiken voor de pijplijn. Dit maakt MongoDB uiteraard nog relevanter. Ten laatste werd ook NewSQL query engine Cloudera Impala in de studie betrokken wegens expliciete interesse van onderzoekers in het eerder vernoemde lab. Deze 6 systemen vergeleek ik 
%TODO ik?
vervolgens op een aantal voor high performance computing relevante eigenschappen zoals indexeringsmechanismen, interfaces naar de gebruiker en API's, distributiestrategie, concurrency controle en consistentiemodel.

\subsection{Document stores}

\paragraph{MongoDB}

MongoDB slaat gegevens op in BSON (binary JSON) documenten. Het systeem biedt krachtige ondersteuning voor indices, vooral door de mogelijkheid om secundaire indices van een brede waaier van types te defini\"eren op alle attributen, zoals in het relationele model. Deze indices zijn gebouwd op B-trees\cite{mongodb_indexes}. Om denormalizatie te bevorderen, kunnen documenten geneste documenten en arrays bevatten. Zo zijn joins ook overbodig in de query taal.\\
De bestandsgrootte is beperkt tot 16 MB, om te voorkomen dat \'e\'en enkel document buitensporig veel RAM of bandbreedte opeist. Om grotere bestanden te bewaren, kan het ingebouwde GridFS (dat integenstelling tot wat de naam doet vermoeden, geen volwaardig file system is) automatisch bestanden opsplitsen in kleinere delen en deze delen als aparte documenten bewaren, zonder dat de gebruiker zich hierom moet bekommeren.\\
MongoDB biedt API's in zeer vele programmeertalen en de functionaliteit om het equivalent van SQL \texttt{WHERE}-clausules te defini\"eren als javascript uitdrukkingen. MongoDB vertaalt deze vervolgens naar een eigen, interne en afgeschermde query taal \cite{grolinger2013data}. De query optimizer van MongoDB verwerkt queries en kiest voor elke query een zo effici\"ent mogelijk uitvoeringsplan gegeven de beschikbare indices. Deze plannen worden gecached als er meerdere goede alternatieven zijn en kunnen geherevalueerd worden naarmate de gegevensset in de databank evolueert.\\
Qua consistentie laat MongoDB de keuze tussen uiteindelijke en strikte consistentie. Strikte consistentie is mogelijk door ofwel enkel te lezen van de master node (die de meest up-to-date versie van de data heeft) of na schrijfopdrachten te wachten tot alle replica's bevestigd hebben alvorens verder te gaan. De eerste optie introduceert een bottleneck bij het lezen van data, de tweede verhoogt de latentie van schrijfopdrachten.\\
MongoDB repliceert data asynchroon en partitioneert in ranges: nodes zijn verantwoordelijk voor ranges van keys. Dit zorgt voor snelle range queries, maar kan hotspots en load-balancingproblemen veroorzaken. Dankzij een master-slave struktuur kan MongoDB updates gemakkelijk naar de juiste replica's doorverwijzen.
Op het gebied van concurrency controle biedt MongoDB atomiciteit binnen documenten en reader-writer locks. 
%TODO beter woord dan locken
Bij schrijfopdrachten de databank locken heeft een zware impact op de performantie in scenario's waar veel geschreven moet worden.\\\\
Kortom, MongoDB bewaart BSON-bestanden op een zeer toegankelijke manier met flexibele query- en indexeringsmechanismes. De concurrency- en consistentiemodellen daarentegen vertonen enkele nadelen.

\paragraph{Couchbase Server}

Couchbase, het resultaat van de fusie tussen CouchDB en Membase, slaat gegevens op in JSON documenten. Het hanteert het memcached protocol om een gedistribueerde cache en is bedoeld voor zeer interactieve toepassingen met hoge vereisten op gebied van latentie \cite{grolinger2013data}\cite{couchbase_about}.
De JSON documenten kunnen genest zijn en kunnen doorzocht worden met een uitgebreide, SQL-achtige taal, N1QL (op het moment van schrijven is dit wel nog steeds een developer preview, uitgebracht in januari 2015) \cite{couchbase_n1ql}.
Net als MongoDB kunnen primaire en secundaire indices gedefinieerd worden en zijn deze gestoeld op B-trees \cite{couchbase_index}.\\
Binnen \'e\'en cluster zijn transacties strikt consistent, maar tussen meerder clusters slechts uiteindelijk consistent.\\
CouchBase biedt gebruikers de keuze tussen optimistische (m.b.v. compare-and-swap) en pessimistische (m.b.v. 'finegrained locking') concurreny controle.\\\\
Dankzij zijn flexibele datamodel, caching en concurrency controle, past CouchBase goed voor toepassingen die snelle en intensieve interactieve vergen tussen gebruiker en data.

\subsection{Columnar stores}

\paragraph{Cassandra}

Cassandra werd oorspronkelijk ontwikkeld voor intern gebruik bij Facebook maar is later als Apache opensourceproject publiekelijk beschikbaar gemaakt. Het combineert het datamodel van Google's BigTable systeem met de architectuur en distributiestrategie van Amazons DynamoDB. Het is gericht op flexibele, quasipermanent beschikbare opslag van zeer grote datasets op goedkope standaardhardware, met daarenboven hoge througput voor schrijfopdrachten zonder effici\"entie bij leesopdrachten op te offeren \cite{borthakur2011apache}.\\
Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model, in die zin dat het nu tabellen en samengestelde kolommen
%TODO betere vertaling voor composite columns?
biedt, evenals een eigen query taal, CQL \cite{cassandra_CQL}. CQL vertoont op het gebied van syntax en functionaliteit sterke gelijkenissen met SQL, maar is toch sterk beperkt. Zo biedt het bijvoorbeeld geen \texttt{JOIN}-clausule, en zijn \texttt{WHERE}-clausules aan sterke voorwaarden onderhevig. Cassandra moedigt het samenbewaren van gegevens die samen opgevraagd worden sterk aan en ondersteunt denormalizatie met features zoals collection types.\\
Cassandra heeft indexeringsmechanismes and implementeert deze met log-structured merge trees, met hogere schrijfthroughput als gevolg.
%TODO hier uitleggen of eerder meer nr boven?
Net als BigTable biedt Cassandra ook Bloom filters.

### 2015-04-07 thesis/nosql_survey.tex
nosql survey cassandra

@@
 Het is gericht op flexibele, quasipermanent beschikbare opslag van zeer grote datasets op goedkope standaardhardware, met daarenboven hoge througput voor schrijfopdrachten zonder effici\"entie bij leesopdrachten op te offeren \cite{borthakur2011apache}.\\
-Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model, in die zin dat het nu tabellen en samengestelde kolommen
+Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model\cite{cassandra_there&now}, in die zin dat het nu tabellen en samengestelde kolommen
 %TODO betere vertaling voor composite columns?
@@
 %TODO hier uitleggen of eerder meer nr boven?
-Net als BigTable biedt Cassandra ook Bloom filters.
+Net als BigTable biedt Cassandra ook Bloom filters.\\
+Om lineair in het aantal ingeschakelde nodes te kunnen schalen naar zeer grote datasets, opereert Cassandra op een volledig hi\"erarchieloze wijze.
+Vanuit het perspectief van het CAP-theorema, spitst Cassandra zich toe op availability en partition tolerance, ten koste van onmiddellijke consistentie.
+Het consistentieniveau kan wel per query door de gebruiker bepaald worden, zoals later verduidelijkt wordt.
+Hoge beschikbaarheid en tolerantie voor fouten bereikt Cassandra door asynchroon data te repliceren over verschillende nodes, met consistent hashing en virtuele nodes om frequent komen-en-gaan en incrementeel toevoegen van nodes op te vangen.
+Het aantal replica's kan de gebruiker zelf kiezen.
+Bovendien voorziet Cassandra ook interdatacenterreplicatie, om zelfs het falen van volledige datacenters op te vangen\cite{decandia2007dynamo} \cite{lakshman2010cassandra} \cite{cassandra_then&now}.\\
+Bij lees- en schrijfopdrachten kan de gebruiker een quorum specifi\"eren.
+Hoewel Cassandra met uiteindelijke consistentie voor het oog ontworpen werd, is onmiddellijke consistentie mits een juiste keuze van de quorum dus ook een optie.\\
+Op het gebied van concurrency controle garandeert Cassandra atomiciteit binnen rijen en serializeerbare \textit{lightweight transactions}, eigenlijk compare-and-set functionaliteit, voor grotere operaties.
+\\\\
+Samengevat biedt Cassandra redelijk flexibele datamodellering met (licht beperkte) query- en indexeringsmechanismen via de CQL-interface.
+Het sterkste punt is echter dat Cassandra incrementeel schaalt naar enorme datasets, dankzij uitvoerige replicatie- en foutverwerkingsfeatures.

### 2015-04-07 thesis/nosql_survey.tex
HBase, VoltDB, Impala in verg studie

@@
 Het is gericht op flexibele, quasipermanent beschikbare opslag van zeer grote datasets op goedkope standaardhardware, met daarenboven hoge througput voor schrijfopdrachten zonder effici\"entie bij leesopdrachten op te offeren \cite{borthakur2011apache}.\\
-Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model\cite{cassandra_there&now}, in die zin dat het nu tabellen en samengestelde kolommen
+Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model\cite{cassandra_then&now}, in die zin dat het nu tabellen en samengestelde kolommen
 %TODO betere vertaling voor composite columns?
@@
 Het sterkste punt is echter dat Cassandra incrementeel schaalt naar enorme datasets, dankzij uitvoerige replicatie- en foutverwerkingsfeatures.
+
+\paragraph{HBase}
+
+Apache HBase is een opensource datastore gebaseerd op het datamodel van Google BigTable, die draait bovenop het Hadoop Distributed File System (HDFS) in plaats van het Google File System (GFS).\\
+Sinds z'n lancering heeft HBase verschillende secundaire indexeringsmechanismes verworven.
+Deze zijn ook gebaseerd op LSM trees en daarnaast biedt ook HBase Bloom filters\cite{borthakur2011apache}\cite{hbase_schema}.
+HBase heeft een Java API, maar zonder SQL-achtige geavanceerde querytaal.
+Hierbij moet wel opgemerkt worden dat een omweg via Apache Hive, een ander data-opslag en -analyseproject \cite{apache_hive}, en de bijhorende querytaal HiveQL dit probleem kan verhelpen.
+Dankzij de HDFS-fundering can HBase vlot fungeren als in- en output voor MapReduce taken.\\
+HBase partitioneert data net als BigTable in ranges en repliceert gegevens op ofwel master-slave of multi-master wijze.
+Leesopdrachten worden echter niet gedistribueerd: er is slechts 1 server die instaat voor elke rij.
+De replica's zijn enkel bestemd voor het herstellen van fouten.\\
+De sterke punten van HBase zijn de sterke consistentie, een rariteit in de NoSQL-wereld, en concurrency-model: ACID transactions binnen rijen en optimistische, multi-version concurrency controle voor grootschaligere operaties \cite{hbase_acid}\cite{grolinger2013data}\cite{borthakur2011apache}.\\\\
+
+Kort samengevat laat HBase gebruikers toe flexibel data te modelleren en is het systeem vooral nuttig wanneer het startpunt een dataset in HDFS is en MapReduce-compatibiliteit een prioriteit is.
+HBase schaalt goed naar zeer grote datasets en beschikt over uitstekende concurrency- en consistentie-eigenschappen.
+
+\subsection{NewSQL}
+
+\paragraph{VoltDB}
+
+VoltDB is een relationele, gedistribueerde in-memory databank, die als doel heeft de garanties van klassieke SQL-stores te koppelen met de schaalbaarheid van NoSQL-systemen, en bovendien aan zeer hoge snelheden te functioneren\cite{stonebraker2013voltdb}.\\
+VoltDB bewaart gegevens in het traditionele relationele model, maar gerepliceerd en gepartitioneerd (met consistent hashing) over verschillende nodes\cite{grolinger2013data}.
+De data is doorzoekbaar via een (groeiende) subset van SQL-92\cite{voltdb2010voltdb}.
+Queries worden bij voorkeur gedefinieerd als stored procedures in Java, met daarin ingebed de SQL uitdrukkingen.
+VoltDB ondersteunt primaire en secundaire indices en laat de gebruiker de keuze tussen hash- en boomindices\cite{voltdb_indexes}.
+VoltDB plant en optimaliseert de queries in stored procedures offline tijdens het compileren\cite{voltdb_query_plans}.\\
+Omdat VoltDB volledig op RAM geheugen vertrouwt, is het duur om te schalen naar datavolumes in de grootorde van meerdere petabytes, maar VoltDB kan data exporteren naar andere, meer geschikte databanksystemen zoals columnaire NoSQL systemen.\\
+Vooral op het gebied van concurrency controle onderscheidt VoltDB zich van andere systemen: transacties voldoen aan de ACID-eigenschappen en worden simultaan op alle replica's uitgevoerd.
+Het geheugen is in blokken verdeeld, die elk statisch toegewezen zijn aan \'e\'en enkele, single-threaded core.
+Een globale controller serializeert alle transacties waarbij meerdere nodes betrokken zijn tot een sequentie van enkelvoudige transacties en voegt deze in de transaction queues van de betrokken nodes.
+Op deze manier maakt VoltDB locking en latching technieken overbodig.
+De durability uit ACID bereikt VoltDB door op regelmatige basis snapshots van de databank in het geheugen te nemen en deze op schijf op te slaan.\\\\
+
+Kort samengevat is VoltDB een relationele databank in main memory die schaalt tot relatief grote datasets, met zeer snelle SQL-query-capaciteiten en ACID-transacties.
+Het is bijgevolg meer geschikt voor rekenintensieve toepassingen die niet overdreven veel data verwerken, maar wel zeer lage latentie vereisen.
+
+\paragraph{Cloudera Impala}
+
+Cloudera Impala is een gedistribueerde SQL-machine die draait bovenop de Hadoopstack, ofwel op HDFS of op HBase \cite{cloudera_impala}.
+Ze is specfiek bedoeld voor analytisch gebruik, met een focus op het leveren van real-time query-capaciteiten eerder dan op hoge throughput bij schrijfopdrachten.\\
+Opgeslagen data is toegankelijk via een subset van SQL-92.\\
+Impala's architectuur is bijna perfect symmetrisch gedistribueerd: alle nodes voeren hetzelfde \texttt{impalad}-proces uit, dat de belangrijkste databankfunctionaliteiten verzorgt.
+Elke node kan als startpunt fungeren voor een query en zal vervolgens de query in kwestie co\"ordineren.
+Twee processen lopen daarentegen op \'e\'en enkele (niet noodzakelijk dezelfde) node in het cluster; zij staan in voor boekhoudkundige taken en doorgeven van wijzingen aan metadata doorheen het cluster \cite{impala_components}.\\
+In tegenstelling tot voorgaande opties partitioneert Impala rijen niet automatisch.
+De gebruiker krijgt wel de keuze hiertoe, voor in het geval de hoeveelheid gegevens hierom zou vragen \cite{impala_partitioning}.\\
+Impala beschikt over een zeer effici\"ente I/O-laag die schijf- en CPU-gebruik te allen tijde hoog houdt, wat resulteert in aanzienlijk snellere prestaties dan andere SQL-on-Hadoop oplossingen zoals Apache Hive\cite{floratou2014sql}.
+Het inherente nadeel is echter dat de volledige dataset moet passen in het totale werkgeheugen van het cluster waarin Impala draait.
+Dit beperkt enigszins de grootte van datasets diet Impala kan verwerken, ondanks de schaalbaarheid van de onderliggende Hadooplaag.\\
+Omwille van z'n analytische doeleinden heeft Impala geen uitvoerige mechanismes voor concurrency controle, maar vertrouwt hiervoor op het onderliggende opslagsysteem.
+Gezien de uitstekende concurrency-eigenschappen van HBase vormt dit niet noodzakelijk een probleem.

### 2015-04-08 gemini/gemini/GeminiQuery.py
unit tests for wildcards. Fixed some bugs and now they all run smoothly!

@@ -469,8 +469,8 @@ class GeminiQuery(object):
         self.nr_cores = nr_cores
         if self._is_gt_filter_safe() is False:
             sys.exit("ERROR: unsafe --gt-filter command.")
-        if not self.gt_filter == None:
-            self.gt_filter = self.gt_filter.replace('==','=')
+        '''if not self.gt_filter == None:
+            self.gt_filter = self.gt_filter.replace('==','=')'''
         
         self.show_variant_samples = show_variant_samples
         self.variant_samples_delim = variant_samples_delim
@@ -862,9 +862,11 @@ class GeminiQuery(object):
     
     def gt_base_parser(self, clause):
         if (clause.find("gt") >= 0 or clause.find("GT") >= 0) and not '[' in clause:
-            return self.gt_filter_to_query_exp(clause)
+            return self.gt_filter_to_query_exp(clause.replace('==', '='))
         elif (clause.find("gt") >= 0 or clause.find("GT") >= 0) and '[' in clause:
-            return self.parse_gt_wildcard(clause)
+            dink = self.parse_gt_wildcard(clause)
+            #print dink.to_string()
+            return dink
         else:
             sys.exit("ERROR: invalid --gt-filter command")   
         
@@ -906,9 +908,12 @@ class GeminiQuery(object):
     
         (column, wildcard, wildcard_rule, wildcard_op) = token.split('.')
         column = column.strip('[').strip(']').strip()
-        wildcard = wildcard.strip('[').strip(']').strip()
+        wildcard = wildcard.strip('[').strip(']').strip().replace('==', '=')
         wildcard_rule = wildcard_rule.strip('[').strip(']').strip()
         wildcard_op = wildcard_op.strip('[').strip(']').strip()
+        
+        if not (wildcard_op.lower().strip() in ['any', 'all', 'none'] or wildcard_op.lower().strip().startswith('count')):
+            sys.exit("Unsupported wildcard operation: (%s). Exiting." % wildcard_op)
                     
         sample_names = self._get_matching_sample_ids(wildcard)
     
@@ -951,16 +956,14 @@ class GeminiQuery(object):
         # distinguish the genotype-specific columns from the base columns
         if "from" not in self.query.lower():
             sys.exit("Malformed query: expected a FROM keyword.")
-
-
+        
+        tokens_to_be_removed = set()
         for token in self.requested_columns:
-
             # it is a WILDCARD
             if (token.find("gt") >= 0 or token.find("GT") >= 0) \
                 and '.(' in token and ').' in token:
[…]

### 2015-04-08 gemini/gemini/database_cassandra.py

@@ -93,7 +93,14 @@ def create_tables(session, gt_column_names, extra_sample_columns):
                         sample_name text, \
                         gt_depth int, \
                         variant_id int, \
-                        primary key (sample_name, gt_depth, variant_id))'''))  
+                        primary key (sample_name, gt_depth, variant_id))''')) 
+     
+    session.execute(SimpleStatement('''CREATE TABLE if not exists variants_by_samples_gt( \
+                        sample_name text, \
+                        gt text, \
+                        variant_id int, \
+                        primary key (sample_name, gt, variant_id))'''))
+    
     
     session.execute(SimpleStatement('''CREATE TABLE IF NOT EXISTS samples_by_variants_gt_type ( \
                     variant_id int, \

### 2015-04-08 gemini/gemini/gemini_load_chunk.py

@@ -155,15 +155,18 @@ class GeminiLoader(object):
                 
             var_sample_gt_types_buffer = blist([])
             var_sample_gt_depths_buffer = blist([])
+            var_sample_gt_buffer = blist([])
                 
             for sample in sample_info:
                     #TODO: check if gt_type is None. Compare if faster to check client side i.s.o. Cassandra - side
                 var_sample_gt_depths_buffer.append([self.v_id, sample[0], sample[2]])
                 var_sample_gt_types_buffer.append([self.v_id, sample[0], sample[1]])
+                var_sample_gt_buffer.append([self.v_id, sample[0], sample[3]])
                                     
             batch_insert(self.session, 'variants_by_samples_gt_type', ["variant_id", "sample_name", "gt_type"], var_sample_gt_types_buffer)
             batch_insert(self.session, 'samples_by_variants_gt_type', ["variant_id", "sample_name", "gt_type"], var_sample_gt_types_buffer)
             batch_insert(self.session, 'variants_by_samples_gt_depth', ["variant_id", "sample_name", "gt_depth"], var_sample_gt_depths_buffer)
+            batch_insert(self.session, 'variants_by_samples_gt', ["variant_id", "sample_name", "gt"], var_sample_gt_buffer)
                 # add each of the impact for this variant (1 per gene/transcript)
             for var_impact in variant_impacts:
                 self.var_impacts_buffer.append(var_impact)
@@ -481,7 +484,7 @@ class GeminiLoader(object):
             gt_columns = concat([gt_bases, gt_types, gt_phases, gt_depths, gt_ref_depths, gt_alt_depths, gt_quals, gt_copy_numbers])
 
             for entry in var.samples:
-                sample_info.append((entry.sample, entry.gt_type, entry.gt_depth))
+                sample_info.append((entry.sample, entry.gt_type, entry.gt_depth, entry.gt_bases))
 
             # tally the genotypes
             #TODO: perhapds uncomment? Don't understand the use just yet.

### 2015-04-08 gemini/gemini/query_expressions.py

@@ -142,7 +142,7 @@ class GT_wildcard_expression(Expression):
         self.keyspace = keyspace
         
     def to_string(self):
-        return "[%s].[%s].[%s].[%s]" % (self.column, "?", self.wildcard_rule, self.rule_enforcement)
+        return "[%s].[%s].[%s].[%s]" % (self.column, ','.join(self.names), self.wildcard_rule, self.rule_enforcement)
     
     def can_prune(self):
         return True
@@ -174,7 +174,7 @@ class GT_wildcard_expression(Expression):
             target_rule = self.rule_enforcement
             corrected_rule = self.wildcard_rule
             
-        if starting_set == "*" and (invert or target_rule == 'none'):
+        if starting_set == "*" and (invert or target_rule == 'none' or target_rule == 'count'):
             correct_starting_set = array.array('i', rows_as_set(session.execute("SELECT variant_id FROM variants")))
         elif starting_set != "*":
             correct_starting_set = array.array('i', starting_set)
@@ -218,9 +218,10 @@ class GT_wildcard_expression(Expression):
             res = diff(set(correct_starting_set), res)
         
         if target_rule == 'count':
-            res_dict = dict()
+            res_dict = {x: 0 for x in correct_starting_set}
             for d in results:
-                res_dict = add_sub_res_to_count_dict(res_dict, d)            
+                for var, count in d.iteritems():
+                    res_dict[var] += count     
             if invert_count:
                 #TODO: if starting_set == "*", retrieve nr of variants somewhere
                 total = len(starting_set)
@@ -258,7 +259,10 @@ def all_query(conn, field, clause, initial_set, contact_points, keyspace):
     
     names = conn.recv()
     
-    results = set(initial_set)
+    if initial_set != "*":
+        results = set(initial_set)
+    else:
+        results = initial_set
     
     for name in names:
         
@@ -266,18 +270,15 @@ def all_query(conn, field, clause, initial_set, contact_points, keyspace):
             break
         
         query = "select variant_id from variants_by_samples_%s WHERE sample_name = '%s' AND %s %s " % (field, name, field, clause)
-        
-        if not any (op in clause for op in ["<", ">"]):  
-            if initial_set != "*":     
-                in_clause = ",".join(map(lambda x: str(x), results))
-                query += " AND variant_id IN (%s)" % in_clause
-            results = rows_as_set(session.execute(query))            
+                
+        if results == "*":
+            results = rows_as_set(session.execute(query))
[…]

### 2015-04-08 gemini/master_test_cassandra.sh

@@ -15,3 +15,5 @@ bash cassandra-test-setup.sh
 bash cassandra-test-query.sh
 
 bash test-load.sh
+
+bash test-wildcards.sh

### 2015-04-08 gemini/test/cassandra-test-setup.sh

@@ -1,2 +1,2 @@
 gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_query_db --cores 2 
-
+gemini load --skip-gene-tables --test-mode -p test_extended_ped.ped -v test4.vep.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks extended_ped_db

### 2015-04-08 gemini/test/test-wildcards.sh

@@ -1,27 +1,18 @@
-check()
-{
-  if diff $1 $2; then
-      echo ok
-  else
-      echo fail
-  fi
-}
-export -f check
 
 ########################################################################
 # 1. Test select * wildcard
 ########################################################################
 echo "    wildcard.t1...\c"
-echo "C/C	C/C	C/C	C/C
-T/T	C/T	C/T	C/C
+echo "G/G	G/A	G/G	G/G
+C/C	C/C	C/C	C/C
+T/T	T/T	T/C	T/C
+T/T	C/C	C/T	C/T
 ./.	./.	C/C	C/C
-./.	C/C	C/C	./.
-G/G	G/G	G/G	G/A
-T/T	C/C	C/C	T/T
 ./.	A/A	./.	A/A
-T/T	T/C	T/C	T/T
-C/T	C/C	C/C	C/C" > exp
-gemini query -q "select (gts).(*)from variants" extended_ped.db > obs
+T/T	T/T	C/C	C/C
+C/T	C/C	C/C	C/C
+./.	C/C	C/C	./." > exp
+gemini query -q "select (gts).(*)from variants" -ks extended_ped_db > obs
 check obs exp
 rm obs exp
 
@@ -30,49 +21,47 @@ rm obs exp
 # 2. Test select * wildcard with other columns and a header
 ########################################################################
 echo "    wildcard.t2...\c"
-echo "chrom	start	end	ref	alt	gts.M10475	gts.M10478	gts.M10500	gts.M128215
+echo "chrom	start	end	ref	alt	gts_m10475	gts_m128215	gts_m10500	gts_m10478
 chr10	1142207	1142208	T	C	C/C	C/C	C/C	C/C
-chr10	48003991	48003992	C	T	T/T	C/T	C/T	C/C
-chr10	52004314	52004315	T	C	./.	./.	C/C	C/C
-chr10	52497528	52497529	G	C	./.	C/C	C/C	./.
-chr10	126678091	126678092	G	A	G/G	G/G	G/G	G/A
-chr10	135210790	135210791	T	C	T/T	C/C	C/C	T/T
+chr10	48003991	48003992	C	T	T/T	C/C	C/T	C/T
+chr10	52004314	52004315	T	C	./.	C/C	C/C	./.
+chr10	52497528	52497529	G	C	./.	./.	C/C	C/C
[…]

### 2015-04-08 gemini/gemini/query_expressions.py
extra tests for count wildcards, some bugfixes

@@ -224,7 +224,7 @@ class GT_wildcard_expression(Expression):
                     res_dict[var] += count     
             if invert_count:
                 #TODO: if starting_set == "*", retrieve nr of variants somewhere
-                total = len(starting_set)
+                total = len(self.names)
                 for variant, count in res_dict.iteritems():
                     res_dict[variant] = total - count
             res = set([v for v, c in res_dict.iteritems() if eval(str(c) + self.count_comp)])

### 2015-04-08 gemini/test/test-wildcards.sh

@@ -243,3 +243,40 @@ gemini query --header -q "select chrom, start, end, ref, alt, (gts).(*) from var
 check obs exp
 rm obs exp
 
+########################################################################
+# 19. Test multiple wildcard on same column
+########################################################################
+echo "    wildcard.t19...\c"
+echo "chrom	start	end	ref	alt	gene	gts_m10475	gts_m128215	gts_m10500	gts_m10478
+chr10	1142207	1142208	T	C	WDR37	C/C	C/C	C/C	C/C
+chr10	48003991	48003992	C	T	ASAH2C	T/T	C/C	C/T	C/T
+chr10	52004314	52004315	T	C	ASAH2	./.	C/C	C/C	./.
+chr10	52497528	52497529	G	C	ASAH2B	./.	./.	C/C	C/C
+chr10	135336655	135336656	G	A	SPRN	./.	A/A	./.	A/A" > exp
+gemini query --header -q "select chrom, start, end, ref, alt, gene, (gts).(*) from variants" \
+             --gt-filter "([gt_type].[phenotype=='1'].[!=HOM_REF].[count>=1] && [gt_type].[phenotype=='2'].[!=HOM_REF].[count>=1])" --test-mode -ks extended_ped_db > obs
+check obs exp
+rm obs exp
+
+########################################################################
+# 20. Test multiple wildcard on same column
+########################################################################
+echo "    wildcard.t20...\c"
+echo "No results!" > exp
+gemini query --header -q "select chrom, start, end, ref, alt, gene, (gts).(*) from variants" \
+             --gt-filter "([gt_type].[phenotype=='1'].[!=HOM_REF].[all] && [gt_type].[phenotype=='2'].[==HOM_REF].[all])" -ks extended_ped_db > obs
+check obs exp
+rm obs exp
+
+########################################################################
+# 21. Test multiple wildcard on same column
+########################################################################
+echo "    wildcard.t21...\c"
+echo "chrom	start	end	ref	alt	gene	gts_m10475	gts_m128215	gts_m10500	gts_m10478
+chr16	72057434	72057435	C	T	DHODH	C/T	C/C	C/C	C/C
+chr10	126678091	126678092	G	A	CTBP2	G/G	G/A	G/G	G/G" > exp
+gemini query --header -q "select chrom, start, end, ref, alt, gene, (gts).(*) from variants" \
[…]

### 2015-04-09 gemini/gemini/GeminiQuery.py
genotypes tests - all succeed

@@ -708,7 +708,8 @@ class GeminiQuery(object):
         if self.show_families or self.show_variant_samples:
             if (not 'variant_id' in self.requested_columns) and (not "*" in self.requested_columns):
                 self.extra_columns.append('variant_id')
-
+        if self.sort_results and self.from_table == 'variants' and not 'start' in self.requested_columns:
+            self.extra_columns.append('start')
         self._execute_query()
         
 

### 2015-04-09 gemini/master_test_cassandra.sh

@@ -17,3 +17,5 @@ bash cassandra-test-query.sh
 bash test-load.sh
 
 bash test-wildcards.sh
+
+bash test-genotypes.sh

### 2015-04-09 gemini/test/cassandra-test-setup.sh

@@ -1,2 +1,4 @@
-gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_query_db --cores 2 
-gemini load --skip-gene-tables --test-mode -p test_extended_ped.ped -v test4.vep.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks extended_ped_db
+gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_query_db --cores 3 
+gemini load --skip-gene-tables --test-mode -p test_extended_ped.ped -v test4.vep.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks extended_ped_db --cores 3
+gemini load --skip-gene-tables --test-mode -v test.snpeff.vcf --skip-gerp-bp --skip-cadd     -t snpEff -ks test_snpeff_vcf_db --cores 3
+

### 2015-04-09 gemini/test/test-genotypes.sh

@@ -1,12 +1,4 @@
-check()
-{
-	if diff $1 $2; then
-    	echo ok
-	else
-    	echo fail
-	fi
-}
-export -f check
+
 ####################################################################
 # 1. Test accessing individual genotype alleles
 ####################################################################
@@ -21,9 +13,9 @@ T/T	T/T	T/T	T/T
 ./.	./.	A/G	A/G
 A/A	A/T	A/A	A/A
 ./.	G/G	G/G	G/G" > exp
-gemini query -q "select gts.1094PC0005, gts.1094PC0009, \
-				gts.1094PC0012, gts.1094PC0013 \
-				from variants" test.snpeff.vcf.db \
+gemini query -q "select gts_1094pc0005, gts_1094pc0009, \
+				gts_1094PC0012, gts_1094PC0013 \
+				from variants" --test-mode -ks test_snpeff_vcf_db \
        > obs
 check obs exp
 rm obs exp
@@ -43,9 +35,9 @@ echo "2	2	2	2
 2	2	1	1
 0	1	0	0
 2	0	0	0" > exp
-gemini query -q "select gt_types.1094PC0005, gt_types.1094PC0009, \
-	                    gt_types.1094PC0012, gt_types.1094PC0013 \
-	             from variants" test.snpeff.vcf.db \
+gemini query -q "select gt_types_1094pc0005, gt_types_1094pc0009, \
+	                    gt_types_1094pc0012, gt_types_1094pc0013 \
+	             from variants" --test-mode -ks test_snpeff_vcf_db \
        > obs
 check obs exp
 rm obs exp
@@ -57,11 +49,11 @@ rm obs exp
 echo "    genotypes.t03...\c"
 echo "0	0	1	0
 2	2	1	1" > exp
-gemini query -q "select gt_types.1094PC0005, gt_types.1094PC0009, \
-	                    gt_types.1094PC0012, gt_types.1094PC0013 \
+gemini query -q "select gt_types_1094pc0005, gt_types_1094pc0009, \
+	                    gt_types_1094pc0012, gt_types_1094pc0013 \
 	             from variants" \
-			 --gt-filter "gt_types.1094PC0012 == HET" \
[…]

### 2015-04-09 thesis/implementatie.tex
tekst: gemini load implementatie

--- 2015-04-03 thesis/datamodel_cassandra.tex
@@
-\chapter{Datamodel Cassandra}
+\chapter{Implementatie}
+\label{implementatie}
+
+Dit hoofdstuk belicht aan de hand van een case-study de praktische realisatie van een schaalbare tool voor software-analyse.
+De case-study in kwestie is een implementatie van GEMINI die werkt met een Cassandra i.p.v.
+SQLite databank.
+Achtereenvolgens komen een motivatie voor de keuze voor Cassandra, de nodige aanpassingen aan het dataschema van GEMINI, en de nodige aanpassingen aan de implementatie van de querying-functionaliteit van GEMINI aan bod.
+
+\section{Keuze database}
+
+\subsection{Vereisten}
+
+Zoals beschreven in \ref{gemini_beschrijving} bewaart GEMINI de data over genetische varianten en proefpersonen in enkele zeer grote tabellen.
+Een eerste vereiste voor een database is dus om deze tabulaire data goed te kunnen voorstellen en beheren.
+Omdat de nadruk van dit onderzoek specifiek ligt op het schaalbaar maken van de applicatie, gebeurt dit bij voorkeur d.m.v.
+automatische verspreiding over verschillende nodes in een cluster.\\
+Omdat GEMINI meerdere processoren of zels computers kan gebruiken bij het inladen van de gegevens uit VCF-bestanden, is goede concurrency controle en hoge schrijf-throughput eveneens belangrijk.
+Na het inladen van de VCF bestanden voert GEMINI enkel nog lees-queries uit, dus zijn de belangrijkste verdere vereisten voor een database hoge lees-throughput, goede query-mogelijkheden en indexeringsmechanismes.
+\\Ten laatste is een Python-API ook nuttig, gezien GEMINI in Python ge\"implementeerd is.
+
+\subsection{Keuze}
+
+Denk denk denk: Cassandra niet beste optie, maar toch gekozen. Eerste prioriteit schaalbaarheid?
+
+\section{Dataschema GEMINI in Cassandra}
 \label{cassandra_datamodel}
@@
 
-\section{Datamodel Cassandra}
+\subsection{Datamodel Cassandra}
 
@@
 
-\section{Database schema GEMINI}
+\subsection{Databaseschema GEMINI}
 Zoals blijkt uit bovenstaande beschrijving van het datamodel van Cassandra, leent het systeem zich niet zomaar tot ad-hoc querying: het is niet mogelijk op een performante manier zomaar aan willekeurige kolommen in een tabel voorwaarden op te leggen en deze voorwaarden met elkaar te combineren.
@@
 
-\subsection{\texttt{variants}-tabel vs. genotype-informatie}
+\paragraph{\texttt{variants}-tabel vs. genotype-informatie}
 
@@
 
-\subsection{\texttt{variants-, samples-}tabel vs. arbitraire queries}
+\paragraph{\texttt{variants-, samples-}tabel vs. arbitraire queries}
 
@@
 Ten slotte is het ook nog interessant een feature aan de GEMINI command line interface toe te voegen waarmee de gebruiker zelf nieuwe basistabellen kan defini\"eren wanneer hij dit nodig acht.
+
+\section{Implementatie GEMINI vs. Cassandra}
+
+\paragraph{\texttt{gemini load}}
+
+Het inladen van genoomdata in GEMINI gebeurt in twee stappen.
+De reden voor deze opsplitsing is dat een klein deel van het inladen en initialiseren van de databank niet parallel op meerdere processoren kan verlopen.
+\\In de eerste en kleinste van de twee, initialiseert GEMINI de databank, maakt de nodige tabellen aan en laadt de samples-informatie in.
+GEMINI haalt de namen van de samples uit de VCF-file, en vult deze aan met informatie uit de ped-file.
+Indien er geen ped-file voorhanden is, initialiseert GEMINI de \texttt{samples}-tabel met default-waarden voor de kolommen zoals \texttt{sex} en \texttt{phenotype}.
+Meteen worden ook de extra, aan de \texttt{samples}-tabel verbonden tabellen zoals \texttt{samples\_by\_sex} en \texttt{samples\_by\_phenotype} aangemaakt en gevuld.
+Ook de \texttt{resources, gene\_detailed, gene\_summary, version} worden in deze fase gevuld.
+Dit omdat de pedigree-files en de bestanden met data over de genen niet effici\"ent opgesplitst kunnen worden.
+Bovendien is de duur van het inladen van de \texttt{samples}-tabel zelfs bij 10000'en samples verwaarloosbaar tegenover het inladen en annoteren van de \texttt{variants}-tabel.\\
+
+In de tweede fase gebeurt het leeuwendeel van het werk: het inladen van de variants-informatie uit de (verplicht meegegeven) VCF-file.
+Dankzij de bgzip \cite{bgzip} en grabix \cite{grabix} tools kan GEMINI effici\"ent VCF-bestanden in blokken opsplitsen, comprimeren, hierop een index defini\"eren en vervolgens daarmee verderwerken.
+Zo kan GEMINI eenvoudig het inladen van de variants, die elk \'e\'en lijn in de VCF-file in beslag nemen, gemakkelijk over meerdere cores verdelen.
+Het is uiteraard ook in deze fase dat GEMINI de aan de \texttt{variants}-tabel verwante tabellen zoals \texttt{variants\_by\_samples\_gt\_type}, \texttt{variants\_by\_samples\_gt\_depth} en \texttt{variants\_by\_chrom\_start} opvult.
+Alle cores of nodes in het cluster, voeren rechtstreeks rijen in in \'e\'en en dezelfde tabel in de Cassandra-databank.
+Omdat Cassandra atomiciteit binnen rijen garandeert en alle workers strikt disjuncte sets van variants invoeren in het systeem, is er geen enkel risico op conflicten.\\\\
+De grote winst ten opzichte van de SQLite-implementatie van GEMINI is dat in het parallelle gedeelte, alle workers naar de zelfde databank kunnen schrijven.
+In de SQLite-versie van GEMINI schreven alle workers naar een aparte SQLite-database, die vervolgens in een kostelijke operatie nog allemaal samengevoegd moeten worden tot \'e\'en grote SQLite-databank.
+Deze stap (het \texttt{gemini merge}-commando) kan de Cassandra-implementatie volledig overslaan.

### 2015-04-09 gemini/gemini/gemini_load_chunk.py
unittest test-exac

@@ -778,9 +778,9 @@ def parse_float(s):
         return float(s)
     except ValueError:
         #TODO: sensible value?
-        return -42.0
+        return None
     except TypeError:
-        return -43.0
+        return None
     
 def parse_int(s):
     try:

### 2015-04-09 gemini/master_test_cassandra.sh

@@ -19,3 +19,5 @@ bash test-load.sh
 bash test-wildcards.sh
 
 bash test-genotypes.sh
+
+bash test-exac.sh

### 2015-04-09 gemini/test/cassandra-test-setup.sh

@@ -1,4 +1,5 @@
 gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_query_db --cores 3 
 gemini load --skip-gene-tables --test-mode -p test_extended_ped.ped -v test4.vep.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks extended_ped_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test.snpeff.vcf --skip-gerp-bp --skip-cadd     -t snpEff -ks test_snpeff_vcf_db --cores 3
+gemini load --skip-gene-tables --test-mode -v test.exac.vcf --skip-gerp-bp --skip-cadd -ks test_exac_db
 

### 2015-04-09 gemini/test/test-exac.sh

@@ -1,20 +1,21 @@
+
 #############################################
 # 1. Test ExAC allele frequencies
 #############################################
 echo "    exac.t01...\c"
 echo "chrom	start	end	ref	alt	in_exac	aaf_exac_all	aaf_adj_exac_all	aaf_adj_exac_afr	aaf_adj_exac_amr	aaf_adj_exac_eas	aaf_adj_exac_fin	aaf_adj_exac_nfe	aaf_adj_exac_oth	aaf_adj_exac_sas
-chr1	985954	985955	G	C	0	None	None	None	None	None	None	None	None	None
-chr1	1199488	1199489	G	A	0	None	None	None	None	None	None	None	None	None
-chr1	1959698	1959699	G	A	1	0.017	0.0175534349801	0.00481877226063	0.00973012667523	0	0.0410552061495	0.0237265929571	0.0232843137255	0.00655049605698
-chr1	161276552	161276553	G	T	0	None	None	None	None	None	None	None	None	None
-chr1	247587092	247587093	C	T	0	None	None	None	None	None	None	None	None	None
-chr9	112185055	112185056	C	G	1	1.626e-05	1.62665105082e-05	0.000189000189	0	0	0	0	0	0
-chr12	121432116	121432118	GC	G	1	0.0002284	0.000234270414993	0.000539374325782	0.000371839365394	0.000825354902608	0	9.89472017731e-05	0	0.000300571085062
-chr14	105420589	105420590	C	T	1	0.001746	0.00175203039036	0.000301083902047	0.000775862068966	0	0.000296384113811	0.0024867521241	0.00437636761488	0.00168431183831" > exp
+chr1	985954	985955	G	C	False	None	None	None	None	None	None	None	None	None
+chr1	1199488	1199489	G	A	False	None	None	None	None	None	None	None	None	None
+chr1	1959698	1959699	G	A	True	0.0170000009239	0.0175534356385	0.00481877243146	0.00973012670875	0.0	0.0410552062094	0.023726593703	0.0232843142003	0.00655049597844
+chr14	105420589	105420590	C	T	True	0.00174600002356	0.00175203033723	0.000301083899103	0.000775862077717	0.0	0.000296384125249	0.00248675211333	0.00437636766583	0.00168431189377
+chr9	112185055	112185056	C	G	True	1.6260000848e-05	1.62665110111e-05	0.000189000187675	0.0	0.0	0.0	0.0	0.0	0.0
[…]

### 2015-04-09 gemini/master_test_cassandra.sh
test-genome

@@ -21,3 +21,5 @@ bash test-wildcards.sh
 bash test-genotypes.sh
 
 bash test-exac.sh
+
+bash test-genome.sh

### 2015-04-09 gemini/test/cassandra-test-setup.sh

@@ -2,4 +2,5 @@ gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -
 gemini load --skip-gene-tables --test-mode -p test_extended_ped.ped -v test4.vep.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks extended_ped_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test.snpeff.vcf --skip-gerp-bp --skip-cadd     -t snpEff -ks test_snpeff_vcf_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test.exac.vcf --skip-gerp-bp --skip-cadd -ks test_exac_db
+gemini load --skip-gene-tables --test-mode -v test3.snpeff.vcf --skip-gerp-bp --skip-cadd         -ks test3_snpeff_db
 

### 2015-04-09 gemini/test/test-genome.sh

@@ -1,3 +1,4 @@
+
 ################################################################################
 #1. Test genome annotations: rmsk
 ################################################################################
@@ -10,7 +11,7 @@ None
 None
 None
 None" > exp
-gemini query -q "select rmsk from variants" test3.snpeff.db \
+gemini query -q "select rmsk from variants" --test-mode -ks test3_snpeff_db \
  > obs
 check obs exp
 rm obs exp
@@ -20,15 +21,15 @@ rm obs exp
 ################################################################################
 
 echo "    genome.t2...\c"
-echo "0
-0
-0
-0
-0
-1
-0
-0" > exp
-gemini query -q "select in_cpg_island from variants" test3.snpeff.db \
+echo "False
+False
+False
+False
+False
+True
+False
+False" > exp
+gemini query -q "select in_cpg_island from variants" --test-mode -ks test3_snpeff_db \
 > obs
 check obs exp
 rm obs exp
@@ -37,15 +38,15 @@ rm obs exp
 #3. Test genome annotations: in_segdup
 ################################################################################
 echo "    genome.t3...\c"
-echo "1
-1
-1
-0
-0
-0
-0
-0" > exp
-gemini query -q "select in_segdup from variants" test3.snpeff.db \
+echo "True
+True
+True
+False
+False
+False
+False
+False" > exp
+gemini query -q "select in_segdup from variants" --test-mode -ks test3_snpeff_db \
  > obs
 check obs exp
 rm obs exp
@@ -55,15 +56,15 @@ rm obs exp
 ################################################################################
 
 echo "    genome.t4...\c"
-echo "0
-0
-0
-0
-0
-0
-1
-1" > exp
-gemini query -q "select is_conserved from variants" test3.snpeff.db \
+echo "False
+False
+False
+False
+False
+False
+True
+True" > exp
+gemini query -q "select is_conserved from variants" --test-mode -ks test3_snpeff_db \
 > obs
 check obs exp
 rm obs exp
@@ -81,7 +82,7 @@ chr1p36.33
 chr1p36.33
 chr1p36.33
 chr1p35.2" > exp
-gemini query -q "select cyto_band from variants" test3.snpeff.db \
+gemini query -q "select cyto_band from variants" --test-mode -ks test3_snpeff_db \
 > obs
 check obs exp
 rm obs exp
@@ -90,15 +91,15 @@ rm obs exp
[…]

### 2015-04-09 gemini/master_test_cassandra.sh
test-columns

@@ -23,3 +23,5 @@ bash test-genotypes.sh
 bash test-exac.sh
 
 bash test-genome.sh
+
+bash test-columns.sh

### 2015-04-09 gemini/test/cassandra-test-setup.sh

@@ -1,6 +1,8 @@
 gemini load --test-mode -v test.query.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_query_db --cores 3 
 gemini load --skip-gene-tables --test-mode -p test_extended_ped.ped -v test4.vep.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks extended_ped_db --cores 3
-gemini load --skip-gene-tables --test-mode -v test.snpeff.vcf --skip-gerp-bp --skip-cadd     -t snpEff -ks test_snpeff_vcf_db --cores 3
-gemini load --skip-gene-tables --test-mode -v test.exac.vcf --skip-gerp-bp --skip-cadd -ks test_exac_db
-gemini load --skip-gene-tables --test-mode -v test3.snpeff.vcf --skip-gerp-bp --skip-cadd         -ks test3_snpeff_db
+gemini load --skip-gene-tables --test-mode -v test.snpeff.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test_snpeff_vcf_db --cores 3
+gemini load --skip-gene-tables --test-mode -v test.exac.vcf --skip-gerp-bp --skip-cadd -ks test_exac_db --cores 3 
+gemini load --skip-gene-tables --test-mode -v test3.snpeff.vcf --skip-gerp-bp --skip-cadd -ks test3_snpeff_db --cores 3
+gemini load --skip-gene-tables --test-mode -v test.vcf_id.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks test_vcf_id_snpeff_vcf_db --cores 3
+gemini load --skip-gene-tables --test-mode -v test2.snpeff.vcf --skip-gerp-bp --skip-cadd -ks test2_snpeff_db --cores 3
 

### 2015-04-09 gemini/test/test-columns.sh

@@ -1,3 +1,4 @@
+
 ####################################################################
 # 1. Make sure we are representing the correct REF and ALT alleles.
 ####################################################################
@@ -13,7 +14,7 @@ chr1	69510	69511	A	G
 chr1	69760	69761	A	T
 chr1	69870	69871	G	A" > exp
 gemini query -q "select chrom, start, end, ref, alt from variants" \
-	   test.snpeff.vcf.db \
+	   --test-mode -ks test_snpeff_vcf_db \
        > obs
 check obs exp
 rm obs exp
@@ -34,7 +35,7 @@ snp	ts
 snp	tv
 snp	ts" > exp
 gemini query -q "select type, sub_type from variants" \
-	   test.snpeff.vcf.db \
+	   --test-mode -ks test_snpeff_vcf_db \
        > obs
 check obs exp
 rm obs exp
@@ -55,7 +56,7 @@ echo "7
 574
 154" > exp
 gemini query -q "select depth from variants" \
-	   test.snpeff.vcf.db \
+	   --test-mode -ks test_snpeff_vcf_db \
        > obs
 check obs exp
 rm obs exp
@@ -65,18 +66,18 @@ rm obs exp
 # 4. Test qual
 ####################################################################
 echo "    columns.t04...\c"
-echo "50.09
-54.3
-49.48
-51.79
-601.49
-2971.02
-938.56
-101944.44
-1026.71
-82.97" > exp
+echo "50.0900001526
+54.2999992371
+49.4799995422
+51.7900009155
+601.489990234
+2971.02001953
+938.559997559
+101944.4375
+1026.70996094
+82.9700012207" > exp
 gemini query -q "select qual from variants" \
-	   test.snpeff.vcf.db \
+	   --test-mode -ks test_snpeff_vcf_db \
        > obs
 check obs exp
 rm obs exp
@@ -96,7 +97,7 @@ ABFilter;LowQual;QDFilter;QUALFilter;SBFilter
 ABFilter;LowQual;QUALFilter
 None" > exp
 gemini query -q "select filter from variants" \
-	   test2.snpeff.db \
+	   --test-mode -ks test2_snpeff_db \
        > obs
 check obs exp
 rm obs exp
@@ -117,7 +118,7 @@ echo "0	0	0	4
 3	1	0	0
 3	0	0	1" > exp
 gemini query -q "select num_hom_ref, num_het, num_hom_alt, num_unknown \
-	             from variants" test.snpeff.vcf.db \
[…]

### 2015-04-10 gemini/gemini/GeminiQuery.py
tests for multicore evaluation of wilcards. Found a bug. Smashed it.

@@ -922,7 +922,9 @@ class GeminiQuery(object):
         wildcard_rule = self._swap_genotype_for_number(wildcard_rule)
         wildcard_rule = wildcard_rule.replace('==', '=')
         
-        return GT_wildcard_expression(column, wildcard_rule, wildcard_op, sample_names, self.db_contact_points, self.keyspace, self.nr_cores) 
+        actual_nr_cores = min(len(sample_names), self.nr_cores)
+        
+        return GT_wildcard_expression(column, wildcard_rule, wildcard_op, sample_names, self.db_contact_points, self.keyspace, actual_nr_cores) 
     
     def _swap_genotype_for_number(self, token):
                 

### 2015-04-10 gemini/master_test_cassandra.sh

@@ -10,13 +10,17 @@ export -f check
 
 cd test
 
-bash cassandra-test-setup.sh
+#bash cassandra-test-setup.sh
 
 bash cassandra-test-query.sh
+
+bash test-query-multicore.sh
 
 bash test-load.sh
 
 bash test-wildcards.sh
+
+bash test-wildcards-multicore.sh
 
 bash test-genotypes.sh
 

### 2015-04-10 gemini/master_test_cassandra.sh
clinvar unittest

@@ -29,3 +29,5 @@ bash test-exac.sh
 bash test-genome.sh
 
 bash test-columns.sh
+
+bash test-clinvar.sh

### 2015-04-10 gemini/test/cassandra-test-setup.sh

@@ -5,4 +5,5 @@ gemini load --skip-gene-tables --test-mode -v test.exac.vcf --skip-gerp-bp --ski
 gemini load --skip-gene-tables --test-mode -v test3.snpeff.vcf --skip-gerp-bp --skip-cadd -ks test3_snpeff_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test.vcf_id.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks test_vcf_id_snpeff_vcf_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test2.snpeff.vcf --skip-gerp-bp --skip-cadd -ks test2_snpeff_db --cores 3
+gemini load --skip-gene-tables --test-mode -v test.clinvar.vcf --skip-gerp-bp --skip-cadd -ks test_clinvar_db
 

### 2015-04-10 gemini/test/test-clinvar.sh

@@ -1,12 +1,4 @@
-check()
-{
-        if diff $1 $2; then
-        echo ok
-        else
-        echo fail
-        fi
-}
-export -f check
+
 ####################################################################
 # 1. Test clinvar annotations
 ####################################################################
@@ -14,12 +6,12 @@ echo "    clinvar.t01...\c"
 echo "1	pathogenic	MYASTHENIC_SYNDROME,_CONGENITAL,_WITH_PRE-_AND_POSTSYNAPTIC_DEFECTS	ClinVar|OMIM_Allelic_Variant	NM_198576.3:c.5125G>C|103320.0001	germline	MedGen:OMIM:Orphanet	CN186031:615120:ORPHA590	RCV000019902.27	1	0	C
 1	other	Lung_cancer	ClinVar	NM_058167.2:c.132-723C>T	somatic	GeneReviews:MedGen:OMIM:SNOMED_CT	NBK1108:C0684249:211980:187875007	RCV000089747.1	1	0	A
 1	other	Generalized_epilepsy_with_febrile_seizures_plus_type_5|Epilepsy,_juvenile_myoclonic_7|Epilepsy,_idiopathic_generalized_10	ClinVar|OMIM_Allelic_Variant	NM_000815.4:c.659G>A|137163.0002	germline	MedGen|MedGen|MedGen:OMIM	C3150401|CN043549|C2751603:613060	RCV000017599.1|RCV000017600.1|RCV000022558.1	1	0	A
+0	None	AllHighlyPenetrant,AllHighlyPenetrant	.,.	.,.	None	.,.	.,.	RCV000080866.1,RCV000080867.1	1	0	A,AT
+1	other	Juvenile_GM>1<_gangliosidosis	ClinVar	NM_000404.2:c.1150A>T	somatic	MedGen:OMIM:Orphanet:Orphanet:SNOMED_CT	C0268272:230600:ORPHA354:ORPHA79256:18756002	RCV000056404.1	1	0	A
+1	other	Lung_cancer	.	.	somatic	GeneReviews:MedGen:OMIM:SNOMED_CT	NBK1108:C0684249:211980:187875007	RCV000104117.1	1	0	T
 1	pathogenic	Roussy-Levy_syndrome	ClinVar|OMIM_Allelic_Variant	NM_000530.6:c.393C>A|159440.0021	germline	MedGen:OMIM:SNOMED_CT	C0205713:180800:45853006	RCV000015250.24	1	0	T
[…]

### 2015-04-12 thesis/implementatie.tex
tekst implementatie queries

@@
 
-\paragraph{\texttt{gemini load}}
+\subsection{\texttt{gemini load}}
 
@@
 Meteen worden ook de extra, aan de \texttt{samples}-tabel verbonden tabellen zoals \texttt{samples\_by\_sex} en \texttt{samples\_by\_phenotype} aangemaakt en gevuld.
-Ook de \texttt{resources, gene\_detailed, gene\_summary, version} worden in deze fase gevuld.
+Ook de \texttt{resources-, gene\_detailed-, gene\_summary-} en \texttt{version-}tabellen worden in deze fase gevuld.
 Dit omdat de pedigree-files en de bestanden met data over de genen niet effici\"ent opgesplitst kunnen worden.
@@
 Omdat Cassandra atomiciteit binnen rijen garandeert en alle workers strikt disjuncte sets van variants invoeren in het systeem, is er geen enkel risico op conflicten.\\\\
-De grote winst ten opzichte van de SQLite-implementatie van GEMINI is dat in het parallelle gedeelte, alle workers naar de zelfde databank kunnen schrijven.
-In de SQLite-versie van GEMINI schreven alle workers naar een aparte SQLite-database, die vervolgens in een kostelijke operatie nog allemaal samengevoegd moeten worden tot \'e\'en grote SQLite-databank.
+De grote winst ten opzichte van de SQLite-implementatie van GEMINI is dat in het parallelle gedeelte, alle workers naar dezelfde databank kunnen schrijven.
+In de SQLite-versie van GEMINI schrijven alle workers naar een aparte SQLite-database, die vervolgens in een kostelijke operatie nog allemaal samengevoegd moeten worden tot \'e\'en grote SQLite-databank.
 Deze stap (het \texttt{gemini merge}-commando) kan de Cassandra-implementatie volledig overslaan.
+Een ander opzicht waarin de Cassandra-implementatie effici\"enter is dan de SQLite-versie, is dat in SQLite voor elke variant de genotype-kolommen nog tot binary blobs gecomprimeerd worden.
+In Cassandra is dit, dankzij de keuze van het databaseschema, niet nodig.
+
+\subsection{\texttt{gemini query}}
+
+Het uitvoeren van de gewoonlijke GEMINI-queries tegen een Cassandra-databank vergt enkele ingrijpende aanpassingen ten opzichte van de SQLite-versie van Cassandra.
+Het doel is natuurlijk deze aanpassingen zoveel mogelijk te verbergen voor de gebruiker, wat ook in grote mate gelukt is.
+In deze sectie komt achtereenvolgens aan bod hoe ingewikkelde queries gesplitst worden in eenvoudige subqueries, hoe de resultaten hiervan vervolgens gecombineerd worden en wat dit betekent voor de meest geavanceerde feature van GEMINI, namelijk de genotype-filter wildcards.
+
+\paragraph{Reductie tot subqueries}
+
+Zoals beschreven in \ref{cassandra_datamodel}, kan Cassandra enkel effici\"ent queries uitvoeren die beperkingen opleggen aan de kolommen in de primary key van een tabel, en dit ook slechts onder bepaalde voorwaarden.
+Bovendien is de enige logische operator die Cassandra ondersteunt om beperkingen te combineren, de conjunctie.\\
+Om arbitraire queries en \texttt{WHERE}-clausules te ondersteunen, kan GEMINI niet zoals in de SQLite-versie de \texttt{WHERE}-clausule zomaar onveranderd aan Cassandra doorgeven.
+Om dit te illustreren, zal ik de onderstaande query als voorbeeld hanteren:\\\\
+\texttt{SELECT chrom, start, subtype FROM variants WHERE chrom = 'chromX' AND start > 5600 AND gene = 'gene\_A'}.\\\\
+Deze query valt uiteen in 2 subqueries: een eerste om uit de \texttt{variants\_by\_chrom\_start}-tabel alle variants op te vragen die voldoen aan de voorwaarde \texttt{chrom = 'chromX' AND start > 5600} en een tweede om uit de \texttt{variants\_by\_gene}-tabel alle variants die voldoen aan \texttt{gene = 'gene\_A'} op te vragen.\\
+
+Om beslissingen zoals in het voorbeeld dynamisch te kunnen maken, is het nodig de \texttt{WHERE}-clausule te inspecteren en hieruit de benodigde hulptabellen voor de subqueries te identificeren.
+Om dit proces te vereenvoudigen, heb ik ervoor gekozen de query syntax licht aan te passen: beperkingen die binnen 1 hulptabel liggen, kunnen nog steeds met elkaar gecombineerd worden met het \texttt{AND}-keyword, maar beperkingen die niet binnen eenzelfde hulptabel liggen, moeten met de \texttt{\&\&, ||} of \texttt{NOT}-operatoren van elkaar gescheiden worden.
+Zo kan GEMINI bij het parsen van de \texttt{WHERE}-clausule deze onmiddellijk opsplitsen in subqueries en voor elk van deze subqueries de nodige hulptabel bepalen.
+Zo wordt de voorbeeldquery:\\
+
+\texttt{SELECT chrom, start, subtype FROM variants WHERE chrom = 'chromX' AND start > 5600 \&\& gene = 'gene\_A'}.\\
+
+Dit vereist van de gebruiker uiteraard dat hij op de hoogte is van welke hulptabellen er bestaan.
+Gebruikers moeten sowieso weten op welke kolommen ze queries kunnen defini\"eren, en bijgevolg ook welke hulptabellen er bestaan.
+Bovendien bestaat het doelpubliek van GEMINI uit onderzoekers die het programma intensief gebruiken en dus voldoende gelegenheden hebben om dergelijke aanpassingen snel onder de knie te krijgen.\\
+%TODO
+Een andere optie is de syntax onveranderd te laten en uit de in de \texttt{WHERE}-clausule vermelde kolommen benodigde hulptabellen af te leiden.
+Deze aanpak heeft als voordeel dat de interface naar de gebruiker niet verandert, maar gezien nog steeds enkel queries mogelijk zijn op kolommen uit de hoofdtabel waarvoor hulptabellen bestaan, is achterwaartse compatibiliteit hiermee nog niet verzekerd.
+Daarbovenop levert deze aanpak ook nieuwe problemen op, zoals: wat met kolommen die in meerdere hulptabellen voorkomen, welke combinatie van tabellen wordt er dan best geselecteerd?
+Dergelijke vraagstukken neigen naar query-optimization en vallen buiten het ...
+Daarom verkoos ik de macht bij de gebruiker te leggen en te vertrouwen op diens oordeelkundigheid.\\\\
+De gebruiker dient er ook rekening mee te houden dat Cassandra range-queries enkel onder strikte voorwaarden ondersteunt, namelijk (zoals eerder uitgelegd) enkel op clustering columns en enkel wanneer de voorgaande kolommen in de clustering key met gelijkheden beperkt zijn.
+Dit impliceert dat er slechts een range query op 1 kolom in een tabel mogelijk is.
+Daarnaast biedt Cassandra ook geen \texttt{!=}-operator.
+Dit euvel valt te verhelpen door een \texttt{!=}-clausule om te zetten naar een gelijkheidsbeperking en deze vervolgens te negeren (zie volgende paragraaf voor evaluatie van negaties).
+Een clausule als \texttt{phenotype != 2} wordt dan \texttt{NOT (phenotype = 2)}.
+Dit vereist dat de \texttt{!=}-clausule een aparte subquery vormt.\\\\
+
+Het uiteindelijke algoritme om de juiste hulptabel te identificeren op basis van de kolommen en gebruikte operatoren in een subquery, gebruikt de metadata van het Cassandra-cluster om alle hulptabellen voor een gegeven hoofdtabel op te vragen, en vervolgens aan de hand van de kolommen in de primary keys van de kandidaat-hulptabellen de geschikte eruit te kiezen.
+%TODO code?
+
+\paragraph{Combineren subqueries}
+
+%recursive datatypes
+\texttt{query\_expression}s are ze weej toe goow
+
+\paragraph{Genotype-filter wildcards}
+
+Ter herinnering, de structuur van een dergelijke wildcard:\\\\
+\texttt{(column).(sample\_wildcard).(gt\_wildcard\_rule).(rule\_enforcement)}\\\\
+Gegeven het bovenstaande query-mechanisme is de meest voor de hand liggende manier om genotype-filter wildcards te implementeren, deze op te splitsen in subqueries voor elke sample die aan de voorwaarden in de \texttt{sample\_wildcard} voldoet, en deze subqueries te combineren met een keten van geneste \texttt{query\_expression}s afhankelijk van de gegeven \texttt{rule\_enforcement}.
+Zo wordt een \texttt{all}-wildcard een keten van geneste \texttt{AND\_expression}s en een \texttt{any}-wildcard een keten van geneste \texttt{OR\_expression}s.
+\texttt{none}-wildcards vereisen eerst dat de subqueries in \texttt{NOT\_expression}d ingebed worden, die vervolgens allemaal in een keten van geneste \texttt{AND\_expression}s gecombineerd worden.
+Het is moeilijker \texttt{count}-wildcards tot dergelijke \texttt{expressions} om te vormen.
+Deze vergen een aparte strategie, die later aan bod komt.\\
+Deze aanpak is vergelijkbaar met wat er in de SQLite- en PostgresQL-versies van GEMINI gebeurt: in SQLite vormt GEMINI de wildcards om tot lange Python-expressions die uiteindelijk voor elke variant met de Python \texttt{eval()}-functie ge\"evalueerd worden.
+Dit uiteraard omdat de genotype-kolommen bestaan uit binary blobs die voor de SQLite-databank geen enkele betekenis hebben.
+De PostgreSQL van GEMINI pakt het slimmer aan en vormt de wildcards om tot lange SQL \texttt{WHERE}-clausules en laat zo het evaluatiewerk over aan de databank.\\
+In dit opzicht lijkt de Cassandra-implementatie op de SQLite-implementatie, gezien de client het meeste werk verricht, in tegenstelling tot de databank.
+Waar Cassandra wel wint, is dat het minder variants zal opvragen dan SQLite
+%TODO argumenteren!
+, en niets meer hoeft te decomprimeren.

### 2015-04-13 gemini/gemini/gemini_subjects.py
sample-filters implemented + some unit-tests

@@ -2,10 +2,9 @@
 import sqlite3
 import os
 import sys
-import numpy as np
 from collections import defaultdict
 
-from gemini_constants import *
+from gemini_constants import HET, HOM_ALT, HOM_REF
 import GeminiQuery
 
 class Subject(object):
@@ -27,7 +26,7 @@ class Subject(object):
         self.mother = True
 
     def _set_fields_from_row(self, row):
-        [setattr(self, k, v) for (k, v) in zip(row.keys(), list(row))]
+        [setattr(self, k, v) for (k, v) in row.iteritems()]
         self.phenotype = int(self.phenotype) if self._has_phenotype() else None
         self._set_affected_status(row)
 
@@ -560,25 +559,6 @@ class Family(object):
 
         return mask
 
-
-    def get_genotype_columns(self):
-        """
-        Return the indices into the gts numpy array for the parents
-        and the children.
-        """
-        columns = []
-
-        if not self.find_parents():
-            for subject in self.subjects:
-                columns.append('gts[' + str(subject.sample_id - 1) + ']')
-        else:
-            columns.append('gts[' + str(self.father.sample_id - 1) + ']')
-            columns.append('gts[' + str(self.mother.sample_id - 1) + ']')
-            for child in self.children:
-                columns.append('gts[' + str(child.sample_id - 1) + ']')
-
-        return columns
-
     def get_genotype_depths(self):
         """
         Return the indices into the gt_depths numpy array for the parents
@@ -711,14 +691,14 @@ def get_subjects(args, skip_filter=False):
     return a dictionary of subjects, optionally using the
     subjects_query argument to filter them.
     """
-    gq = GeminiQuery.GeminiQuery(args.db)
+    gq = GeminiQuery.GeminiQuery(args.contact_points, args.keyspace)
     query = "SELECT * FROM samples"
     if not skip_filter:
         if hasattr(args, 'sample_filter') and args.sample_filter:
             query += " WHERE " + args.sample_filter
-    gq.c.execute(query)
+    res = gq.run_simple_query(query)
     samples_dict = {}
-    for row in gq.c:
+    for row in res:
         subject = Subject(row)
         samples_dict[subject.name] = subject
     return samples_dict

### 2015-04-13 gemini/gemini/query_expressions.py

@@ -16,7 +16,7 @@ class Expression(object):
     @abc.abstractmethod
     def evaluate(self, session, starting_set):
         return
-
+    
     @abc.abstractmethod
     def to_string(self):
         return
@@ -41,12 +41,12 @@ class Simple_expression(Expression):
         if self.where_clause != "":
             query += " WHERE %s" % self.where_clause            
         if self.can_prune() and not starting_set == "*":
-            if self.table == 'samples':
+            if self.table.startswith('samples'):
                 in_clause = "','".join(starting_set)            
                 query += " AND %s IN ('%s')" % (self.select_column, in_clause)
             else:
                 in_clause = ",".join(map(lambda x: str(x), starting_set))            
-                query += " AND %s IN (%s)" % (self.select_column, in_clause)       
+                query += " AND %s IN (%s)" % (self.select_column, in_clause)     
         return rows_as_set(socket.execute(query))
 
     def to_string(self):
@@ -117,7 +117,7 @@ class NOT_expression(Expression):
             correct_starting_set = rows_as_set(session.execute("SELECT %s FROM %s" % (self.select_column, self.table)))
         else:
             correct_starting_set = starting_set
-            
+        
         return diff(correct_starting_set, self.exp.evaluate(session, correct_starting_set))
 
     def to_string(self):

### 2015-04-13 gemini/test/cassandra-test-setup.sh

@@ -6,4 +6,6 @@ gemini load --skip-gene-tables --test-mode -v test3.snpeff.vcf --skip-gerp-bp --
 gemini load --skip-gene-tables --test-mode -v test.vcf_id.snpeff.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks test_vcf_id_snpeff_vcf_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test2.snpeff.vcf --skip-gerp-bp --skip-cadd -ks test2_snpeff_db --cores 3
 gemini load --skip-gene-tables --test-mode -v test.clinvar.vcf --skip-gerp-bp --skip-cadd -ks test_clinvar_db
-
+gemini load --skip-gene-tables --test-mode -p test.de_novo.ped -v test.family.vcf  --skip-gerp-bp --skip-cadd -t snpEff -ks test_family_db
+gemini load --skip-gene-tables --test-mode -v test4.vep.snpeff.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test4_snpeff_db --cores 3
+gemini load --skip-gene-tables --test-mode -p test4.snpeff.ped -v test4.vep.snpeff.vcf --skip-gerp-bp --skip-cadd -t snpEff -ks test4_snpeff_ped_db

### 2015-04-14 gemini/gemini/query_expressions.py
deleted redundant indirection for set ops + to_string -> __str__

@@ -16,10 +16,6 @@ class Expression(object):
     @abc.abstractmethod
     def evaluate(self, session, starting_set):
         return
-    
-    @abc.abstractmethod
-    def to_string(self):
-        return
 
     @abc.abstractmethod
     def can_prune(self):
@@ -49,7 +45,7 @@ class Simple_expression(Expression):
                 query += " AND %s IN (%s)" % (self.select_column, in_clause)     
         return rows_as_set(socket.execute(query))
 
-    def to_string(self):
+    def __str__(self):
         return self.where_clause
 
     def can_prune(self):
@@ -74,10 +70,10 @@ class AND_expression(Expression):
             return self.left.evaluate(session, temp)
         else:
             temp = self.left.evaluate(session, starting_set)
-            return intersect(temp, self.right.evaluate(session, temp))
+            return temp & self.right.evaluate(session, temp)
 
-    def to_string(self):
-        res = "(" + self.left.to_string() + ")" + " AND " + "(" + self.right.to_string() + ")"
+    def __str__(self):
+        res = "(" + str(self.left) + ")" + " AND " + "(" + str(self.right) + ")"
         return res
 
     def can_prune(self):
@@ -93,10 +89,10 @@ class OR_expression(Expression):
         
         if len(starting_set) == 0:
             return set()        
-        return union(self.left.evaluate(session, starting_set), self.right.evaluate(session, starting_set))
+        return self.left.evaluate(session, starting_set) | self.right.evaluate(session, starting_set)
 
-    def to_string(self):
-        res = "(" + self.left.to_string() + ")" + " OR " + "(" + self.right.to_string() + ")"
+    def __str__(self):
+        res = "(" + str(self.left) + ")" + " OR " + "(" + str(self.right) + ")"
         return res
 
     def can_prune(self):
@@ -118,10 +114,10 @@ class NOT_expression(Expression):
         else:
             correct_starting_set = starting_set
         
-        return diff(correct_starting_set, self.exp.evaluate(session, correct_starting_set))
+        return correct_starting_set - self.exp.evaluate(session, correct_starting_set)
 
-    def to_string(self):
-        return "NOT (" + self.exp.to_string() + ")"
+    def __str__(self):
+        return "NOT (" + str(self.exp) + ")"
 
     def can_prune(self):
         return True
@@ -141,7 +137,7 @@ class GT_wildcard_expression(Expression):
         self.db_contact_points = db_contact_points
         self.keyspace = keyspace
         
-    def to_string(self):
+    def __str__(self):
[…]

### 2015-04-14 gemini/gemini/GeminiQuery.py
query_expressions opgekuist, geformat voor tex-layout

@@ -14,7 +14,7 @@ from gemini_subjects import get_subjects
 from gemini_utils import (OrderedDict, itersubclasses, partition_by_fn)
 from sql_utils import ensure_columns
 from collections import namedtuple
-from gemini.query_expressions import Simple_expression, AND_expression,\
+from gemini.query_expressions import Basic_expression, AND_expression,\
     NOT_expression, OR_expression, rows_as_set, GT_wildcard_expression
 from gemini.sql_utils import get_query_parts
 from cassandra.query import ordered_dict_factory, tuple_factory
@@ -537,7 +537,7 @@ class GeminiQuery(object):
                         in_clause = "','".join(matches)            
                         dink_query += " WHERE %s IN ('%s')" % (self.get_partition_key(from_table), in_clause)
                     else:
-                        in_clause = ",".join(map(lambda x: str(x), matches))            
+                        in_clause = ",".join(map(str, matches))            
                         dink_query += " WHERE %s IN (%s)" % (self.get_partition_key(from_table), in_clause)
                 dink_query += " " + rest_of_query
                 self.session.row_factory = ordered_dict_factory
@@ -690,7 +690,7 @@ class GeminiQuery(object):
                 query = "SELECT %s FROM %s" % (','.join(self.requested_columns + self.extra_columns), self.from_table)
                 if self.matches != "*":
                     if self.from_table != 'samples':
-                        in_clause = ",".join(map(lambda x: str(x), self.matches))            
+                        in_clause = ",".join(map(str, self.matches))            
                         query += " WHERE %s IN (%s)" % (self.get_partition_key(self.from_table), in_clause)
                     else:
                         in_clause = "','".join(self.matches)            
@@ -745,7 +745,7 @@ class GeminiQuery(object):
         if wildcard.strip() != "*":        
             query = self.parse_where_clause(wildcard, 'samples')
         else:
-            query = Simple_expression('samples', 'name', "")
+            query = Basic_expression('samples', 'name', "")
         return list(query.evaluate(self.session, '*'))
 
 
@@ -835,7 +835,7 @@ class GeminiQuery(object):
     def where_clause_to_exp(self, table, cols, clause):
         
         target_table = self.get_table_from_where_clause(table, clause)
-        exp = Simple_expression(target_table, cols, clause)
+        exp = Basic_expression(target_table, cols, clause)
         return exp
     
     def get_table_from_where_clause(self, table, where_clause):
[…]

### 2015-04-14 gemini/gemini/query_expressions.py

@@ -21,7 +21,7 @@ class Expression(object):
     def can_prune(self):
         return True
 
-class Simple_expression(Expression):
+class Basic_expression(Expression):
     
     def __init__(self, from_table, select_column, where_clause):
         self.table = from_table
@@ -33,23 +33,27 @@ class Simple_expression(Expression):
         if len(starting_set) == 0:
             return set()
         
-        query = "SELECT %s FROM %s" % (self.select_column, self.table)
+        query = "SELECT %s FROM %s" % \
+            (self.select_column, self.table)
         if self.where_clause != "":
             query += " WHERE %s" % self.where_clause            
         if self.can_prune() and not starting_set == "*":
             if self.table.startswith('samples'):
                 in_clause = "','".join(starting_set)            
-                query += " AND %s IN ('%s')" % (self.select_column, in_clause)
+                query += " AND %s IN ('%s')" % \
+                    (self.select_column, in_clause)
             else:
-                in_clause = ",".join(map(lambda x: str(x), starting_set))            
-                query += " AND %s IN (%s)" % (self.select_column, in_clause)     
+                in_clause = ",".join(map(str, starting_set))            
+                query += " AND %s IN (%s)" % \
+                    (self.select_column, in_clause)     
         return rows_as_set(socket.execute(query))
+    
+    def can_prune(self):
+        return not any (op in self.where_clause \
+                        for op in ["<", ">"])
 
     def __str__(self):
         return self.where_clause
-
-    def can_prune(self):
-        return not any (op in self.where_clause for op in ["<", ">"])
     
 class AND_expression(Expression):
     
@@ -101,7 +105,7 @@ class OR_expression(Expression):
 class NOT_expression(Expression):
     
     def __init__(self, exp, table, select_column):
-        self.exp = exp
+        self.body = exp
         self.table = table
         self.select_column = select_column
  
@@ -110,14 +114,16 @@ class NOT_expression(Expression):
         if len(starting_set) == 0:
             return set()        
         elif starting_set == '*':
-            correct_starting_set = rows_as_set(session.execute("SELECT %s FROM %s" % (self.select_column, self.table)))
+            correct_starting_set = rows_as_set(session.execute(\
+                "SELECT %s FROM %s" % (self.select_column, self.table)))
         else:
             correct_starting_set = starting_set
         
[…]

### 2015-04-14 thesis/implementatie.tex
tekst: implementatie combinatie subqueries, query_expressions

@@
 
-%recursive datatypes
-\texttt{query\_expression}s are ze weej toe goow
+Conceptueel:
+De subqueries zullen elk een verzameling rijen als resultaat opleveren.
+Het is dan zaak deze verzamelingen met set-operaties te combineren tot de finale verzameling rijen die aan de query voldoen.
+De nodige set-operatie hangt af van de oorspronkelijke query.
+Zo zal het eindresultaat van een conjunctie van subqueries de doorsnede van de resultaatverzamelingen van de subqueries zijn.
+In het geval van een disjunctie is dit de unie van de resultaatverzamelingen, en in het geval van een negatie van een subquery is dit het verschil tussen de oorspronkelijke verzameling (voor de query) en de resultaatverzameling van de query.
+Om bijvoorbeeld alle varianten te vinden die niet op een bepaald gen x liggen, moet het verschil bepaald worden tussen de verzameling van alle varianten en de verzameling van alle varianten die op gen x liggen.
+Onderstaande tabel vat dit samen voor 2 subqueries $p$ en $r$, hun respectievelijke resultaatverzamelingen $res(p)$ en $res(r)$ en een initi\"ele verzameling rijen $I$.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|l|l|}
+\hline
+\textbf{Query} & \textbf{Resultaat}    \\ \hline
+$p$ \texttt{AND} $r$ & $res(p) \cap res(r)$   \\ \hline
+$p$ \texttt{OR} $r$ & $res(p) \cup res(r)$    \\ \hline
+\texttt{NOT} $p$  & $I \setminus res(p)$ \\ \hline
+\end{tabular}
+\end{table}
+
+Praktisch:
+
+Uit performantie-overwegingen is het een logische keuze de resultaten van subqueries op de hulptabellen te bewaren in een set datatype.
+Zo kunnen de benodigde set-operaties gebruik maken van de in Python ingebouwde set-operaties en zo veel effici\"enter verlopen dan wanneer de resultaten in bijvoorbeeld lists bewaard worden.
+Het enige nadeel van sets t.o.v.
+lists is dat de implementatie geen garanties biedt over in welke volgorde rijen in het eindresultaat zullen voorkomen.
+Dit weegt niet op tegen de tijdswinst, en bovendien biedt Cassandra zelf in het algemeen geen enkele garantie over in welke volgorde het rijen opslaat.\\\\
+Bij het praktisch evalueren van \texttt{query\_expression}s komt uiteindelijk meer kijken dan enkel set-operaties. De belangrijkste reden hiervoor is dat, zoals in ?? %TODO
+reeds beschreven, om een \texttt{NOT\_expression} te kunnen evalueren, alle kandidaat-rijen gekend moet zijn, en dus meegegeven aan de \texttt{evaluate}-functie.
+Deze extra informatie kan ook bij het evalueren van andere types van \texttt{query\_expression} goed van pas komen.
+
+\begin{itemize}
+\item \texttt{Basic\_expression}
+Het evalueren van \texttt{Basic\_expression}s gebeurt niet zo rechtoe rechtaan als de naam doet vermoeden.
+De \texttt{evaluate}-functie bouwt een CQL-query op, op basis van een gegeven \texttt{WHERE}-clausule, de relevante tabel en de gevraagde kolom uit die tabel.
+Wanneer er een zinvolle verzameling kandidaatrijen (de \texttt{starting\_set} in onderstaand codefragment) meegegeven is, wordt deze als \texttt{IN}-clausule aan de \texttt{WHERE}-clausule toegevoegd.
+Zo zoekt Cassandra enkel in rijen die effectief in aanmerking komen om aan de globale query te voldoen.
+Een voorwaarde om deze optimalisatie te kunnen toepassen, is dat de \texttt{WHERE}-clausule geen ongelijkheidsbeperkingen mag bevatten (vandaar de \texttt{can\_prune}-functie).
+
+\lstinputlisting[language=Python,firstline=31,lastline=31]{../gemini/gemini/query_expressions.py}
+\lstinputlisting[language=Python,firstline=36,lastline=53]{../gemini/gemini/query_expressions.py}
+
+\item Om een \texttt{AND\_expression} te evalueren is het strikt genomen nodig beide leden van de uitdrukking te evalueren en vervolgens de doorsnede te nemen van de resultaten.
+Als \'e\'en van de twee leden van de uitdrukking echter in staat is om dankzij een CQL \texttt{IN}-clausule enkel naar rijen te zoeken die effectief in aanmerking komen om aan de globale query te voldoen, is het mogelijk om deze operatie aan Cassandra over te laten.
+Als het rechterlid kan \textit{prunen}, dan kan GEMINI de resultaatverzameling van het linkerlid als \texttt{starting\_set} meegeven bij de evaluatie van het rechterlid, en omgekeerd.
+Wanneer geen van beide kunnen \textit{prunen}, moet GEMINI uiteraard nog steeds zelf de doorsnede berekenen.
+
+\lstinputlisting[language=Python,firstline=64,lastline=64]{../gemini/gemini/query_expressions.py}
+\lstinputlisting[language=Python,firstline=69,lastline=77]{../gemini/gemini/query_expressions.py}
+
+\item In het geval van een \texttt{OR\_expression} is er geen andere mogelijkheid dan eenvoudigweg het linker- en rechterlid te evalueren, beide met als \texttt{starting\_set} de \texttt{starting\_set} van de \texttt{OR\_expression} zelf en vervolgens de unie van de resultaten te berekenen.
+
+\item Ook in het geval van een \texttt{NOT\_expression} is er geen andere optie dan de \texttt{body}-uitdrukking te evalueren en vervolgens het complement van de \texttt{starting\_set} en het resultaat te bepalen.
+Als de gegeven \texttt{starting\_set} gelijk is aan \texttt{"*"}, d.w.z.
+alle rijen in de oorspronkelijke tabel, zit er ook niets anders op dan deze nog eerst allemaal op te vragen.
+
+\lstinputlisting[language=Python,firstline=112,lastline=112]{../gemini/gemini/query_expressions.py}
+\lstinputlisting[language=Python,firstline=116,lastline=123]{../gemini/gemini/query_expressions.py}
+
+\end{itemize}
+
+Een optimalisatie die voor alle \texttt{query\_expressions} geldt is dat in het geval van een lege \texttt{starting\_set}, GEMINI de hele evaluatie kan overslaan.
+Dit bespaart een overtollige round-trip naar de Cassandra-databank.
+Voor alle soorten \texttt{query\_expressions} behalve \texttt{Basic\_expression} geldt ook dat ze kunnen \textit{prunen}, ongeacht de aard van hun subqueries.
+In het geval dat hun subqueries een \texttt{Basic\_expression} bevatten die niet kan \textit{prunen}, zal de subquery in kwestie dit zelf afhandelen bij de evaluatie.
 
@@
 In dit opzicht lijkt de Cassandra-implementatie op de SQLite-implementatie, gezien de client het meeste werk verricht, in tegenstelling tot de databank.
-Waar Cassandra wel wint, is dat het minder variants zal opvragen dan SQLite
-%TODO argumenteren!
-, en niets meer hoeft te decomprimeren.

### 2015-04-15 gemini/gemini/query_expressions.py
all_query bugfix

@@ -265,7 +265,7 @@ def all_query(conn, field, clause, initial_set, contact_points, keyspace):
             query += " AND variant_id IN (%s)" % in_clause
             results = rows_as_set(session.execute(query))
         else:
-            results = rows_as_set(session.execute(query)) | results
+            results = rows_as_set(session.execute(query)) & results
         
     session.shutdown()   
     

### 2015-04-15 thesis/implementatie.tex
concept & implementatie gt-filter wildcards, sneuvelversie count implementatie

@@
-\chapter{Implementatie}
-\label{implementatie}
+\chapter{Oplossing}
+\label{oplossing}
 
@@
 Zo kan GEMINI bij het parsen van de \texttt{WHERE}-clausule deze onmiddellijk opsplitsen in subqueries en voor elk van deze subqueries de nodige hulptabel bepalen.
-Zo wordt de voorbeeldquery:\\
-
+Zo wordt de voorbeeldquery:\\\\
 \texttt{SELECT chrom, start, subtype FROM variants WHERE chrom = 'chromX' AND start > 5600 \&\& gene = 'gene\_A'}.\\
@@
 Een clausule als \texttt{phenotype != 2} wordt dan \texttt{NOT (phenotype = 2)}.
-Dit vereist dat de \texttt{!=}-clausule een aparte subquery vormt.\\\\
+Dit vereist dat de \texttt{!=}-clausule een aparte subquery vormt.\\
 
@@
 
-Conceptueel:
+\textbf{conceptueel}
 De subqueries zullen elk een verzameling rijen als resultaat opleveren.
@@
 
-Praktisch:
+\textbf{praktisch}
 
@@
 \begin{itemize}
-\item \texttt{Basic\_expression}
-Het evalueren van \texttt{Basic\_expression}s gebeurt niet zo rechtoe rechtaan als de naam doet vermoeden.
+\item Het evalueren van \texttt{Basic\_expression}s gebeurt niet zo rechtoe rechtaan als de naam doet vermoeden.
 De \texttt{evaluate}-functie bouwt een CQL-query op, op basis van een gegeven \texttt{WHERE}-clausule, de relevante tabel en de gevraagde kolom uit die tabel.
@@
 \lstinputlisting[language=Python,firstline=31,lastline=31]{../gemini/gemini/query_expressions.py}
+%TODO juist formatten, binnenste if, else verstoppen
 \lstinputlisting[language=Python,firstline=36,lastline=53]{../gemini/gemini/query_expressions.py}
@@
 
-\item In het geval van een \texttt{OR\_expression} is er geen andere mogelijkheid dan eenvoudigweg het linker- en rechterlid te evalueren, beide met als \texttt{starting\_set} de \texttt{starting\_set} van de \texttt{OR\_expression} zelf en vervolgens de unie van de resultaten te berekenen.
+\item In het geval van een \texttt{OR\_expression} is er geen andere mogelijkheid dan eenvoudigweg het linker- en rechterlid te evalueren, beide met als \texttt{starting\_set} de \texttt{starting\_set} van de \texttt{OR\_expression} zelf, en vervolgens de unie van de resultaten te berekenen.
 
@@
 
-Ter herinnering, de structuur van een dergelijke wildcard:\\\\
-\texttt{(column).(sample\_wildcard).(gt\_wildcard\_rule).(rule\_enforcement)}\\\\
-Gegeven het bovenstaande query-mechanisme is de meest voor de hand liggende manier om genotype-filter wildcards te implementeren, deze op te splitsen in subqueries voor elke sample die aan de voorwaarden in de \texttt{sample\_wildcard} voldoet, en deze subqueries te combineren met een keten van geneste \texttt{query\_expression}s afhankelijk van de gegeven \texttt{rule\_enforcement}.
+\textbf{conceptueel}
+Ter herinnering, de structuur van een genotype-filter wildcard:\\\\
+\texttt{(genotype\_column).(sample\_wildcard).(gt\_wildcard\_rule).(rule\_enforcement)}\\\\
+Het uitvoeren en evalueren van deze wildcards is conceptueel niet erg ingewikkeld: het volstaat de \texttt{sample\_wildcard} te evalueren en vervolgens voor elk van de resulterende samples een subquery op te stellen op de voor \texttt{genotype\_column} relevante hulptabel van de \texttt{variants}-tabel.
+Hoe deze subqueries met elkaar gecombineerd moeten worden, hangt af van de gegeven \texttt{rule\_enforcement}:
+
+\begin{itemize}
+\item Een \texttt{all}-wildcard leidt tot de conjunctie van alle subqueries.
+\item Een \texttt{any}-wildcard leidt tot de disjunctie van alle subqueries.
+\item Een \texttt{none}-wildcard leidt tot de conjunctie van de negaties van alle subqueries.
+\item \texttt{count}-wildcards vallen niet zomaar met set-algebra te evalueren: bij het combineren van de resultaten van alle subqueries moet voor elke variant geteld worden in de resultaatverzameling van hoeveel subqueries hij voorkomt.
+Nadien kan GEMINI eenvoudig op basis van deze tellingen de varianten bepalen die aan de \texttt{count}-regel voldoen.
+\end{itemize}
+
+Dit betekent dat weer veel werk van de databank naar GEMINI zelf verschuift.
+Een mogelijke manier om dit leed te verzachten is, gezien de goede concurrency-eigenschappen van NoSQL-systemen als Cassandra, om de subqueries in parallel te evalueren in plaats van sequentieel.\\
+
+\textbf{implementatie} Gegeven het bovenstaande query-mechanisme is de meest voor de hand liggende manier om genotype-filter wildcards te implementeren, deze op te splitsen in subqueries voor elke sample die aan de voorwaarden in de \texttt{sample\_wildcard} voldoet, en deze subqueries te combineren met een keten van geneste \texttt{query\_expression}s afhankelijk van de gegeven \texttt{rule\_enforcement}.
 Zo wordt een \texttt{all}-wildcard een keten van geneste \texttt{AND\_expression}s en een \texttt{any}-wildcard een keten van geneste \texttt{OR\_expression}s.
@@
 Dit uiteraard omdat de genotype-kolommen bestaan uit binary blobs die voor de SQLite-databank geen enkele betekenis hebben.
-De PostgreSQL van GEMINI pakt het slimmer aan en vormt de wildcards om tot lange SQL \texttt{WHERE}-clausules en laat zo het evaluatiewerk over aan de databank.\\
-In dit opzicht lijkt de Cassandra-implementatie op de SQLite-implementatie, gezien de client het meeste werk verricht, in tegenstelling tot de databank.
+De PostgreSQL-versie van GEMINI pakt het slimmer aan en vormt de wildcards om tot lange SQL \texttt{WHERE}-clausules en laat zo het evaluatiewerk over aan de databank, met de bemerking dat de PostgreSQL-versie \texttt{count}-wildcards vooralsnog niet ondersteunt omdat dit niet rechtstreeks in SQL vertaald kunnen worden.\\
+In dit opzicht lijkt de Cassandra-implementatie op de SQLite-implementatie, gezien de client, niet de databank, het meeste werk verricht.\\
+
+Zoals in de conceptuele uitleg van de oplossing aangehaald, kan de evaluatie van genotype-wildcards effici\"enter verlopen door de evaluatie van de subqueries in parallel uit te voeren op meerdere processoren.
+Dit heb ik ook daadwerkelijk ge\"implementeerd, door introductie van een nieuw type \texttt{query\_expression}, namelijk de \texttt{GT\_wildcard\_expression}.\\
+Bij evaluatie zal deze de namen van alle samples die aan de \texttt{sample\_wildcard} voldoen verdelen over het aantal processoren dat de gebruiker meegeeft, deze elk een tussenresultaat laten berekenen, de resultaten opvragen en samenvoegen m.b.v.
+de gekende setalgebra.
+Om het werk effici\"ent over verschillende processoren te verdelen en alle benodigde informatie te communiceren, maakt de implementatie gebruik van zogenaamde subprocesses, die elk hun eigen verbinding openen met het Cassandra-cluster, en interprocess-communicatiemechanismen uit de Python \texttt{multiprocessing}-API \cite{multiprocessing}.\\
+\texttt{all-, any-} en \texttt{none-}wildcards vereisen voor het samenvoegen van de resultaten van de verschillende subprocesses geen grote aanpassingen aan de sequenti\"ele evaluatie-strategie van wildcards.
+Omdat deze oplossingswijze er zich makkelijk toe leent af te wijken van het stramien van de geneste \texttt{query\_expressions} (immers, er is maar 1 \texttt{GT\_wildcard\_expression} voor alle samples), is het makkelijker de \texttt{count}-wildcards te implementeren: elk subprocess voert subqueries uit voor de samples waarvoor het verantwoordelijk is en telt voor elke variant hoe vaak hij in de resultaatverzamelingen van de subqueries voorkwam.
+Deze informatie slaat stuurt het subprocess terug naar zijn parent-process in een Python \textit{dictionary}, met een map van \texttt{variant\_id}s naar de telling.
+Het parent-process kan vervolgens de telling voor alle variants van alle subprocesses bij elkaar optellen, en vervolgens met gebruik van de Python \texttt{eval}-functie alle variants die aan de \texttt{count}-filter voldoen, bepalen.
+Let wel, bij het opstellen van de uiteindelijke dictionary met de tellingen voor de varianten, moet de parent-process ook de varianten die in het resultaat van geen enkel subprocess voorkwamen (waarvoor m.a.w.
+geen enkele sample aan de voorwaarden voldoet), een entry met waarde 0 voorzien.
+Dit om ook voor \texttt{count}-filters met \texttt{<, <=, ==0}-operatoren alle varianten te kunnen opvragen.

### 2015-04-15 thesis/thesis.tex
text update - begin cha2
Overzicht NoSQL & NewSQL
tekst datamodel cassandra begin
cassandra indices + begin dataschema
nosql methodologie + vergelijkende studie
tekst: gemini load implementatie
tekst implementatie queries
tekst: implementatie combinatie subqueries, query_expressions
concept & implementatie gt-filter wildcards, sneuvelversie count implementatie

@@
 \documentclass[master=cws,masteroption=gs]{kulemt}
-\setup{title={Master thesis Gemini Cassandra},
+\setup{title={Schaalbare genoomanalyse met NoSQL- databases},
 		author={Brecht Gossel\'e},
-		promotor={Prof. Roel Wuyts},
+		promotor={Prof. Dr. Roel Wuyts},
 		assessor={},
-		assistant={Prof. Roel Wuyts}}
+		assistant={Prof. Dr. Roel Wuyts}}
 \setup{filingcard,
@@
 		shortabstract={Here comes a very short abstract, containing no more than 500 words. \LaTeX\ commands can be used here.}}
+
+\usepackage{url}
+\usepackage{graphicx}
+\usepackage{booktabs}
+\graphicspath{ {afbeeldingen/} }
+\usepackage{color}
+\usepackage{listings}
+\lstset{basicstyle=\footnotesize, showspaces=false, keywordstyle=\color{blue}, stringstyle=\color[RGB]{26,196,8}}
+\usepackage[pdfusetitle,colorlinks,plainpages=false]{hyperref}
+
 \begin{document}
+
+\begin{preface}
+\end{preface}
+
+\tableofcontents
+
+\begin{abstract}
+\end{abstract}
+
 \mainmatter
-Oi
+
+\include{inleiding}
+\include{NoSQL_overzicht}
+\include{nosql_survey}
+\include{gemini_beschrijving}
+\include{implementatie}
+
+\bibliographystyle{abbrv}
+\bibliography{biblio}
 \end{document}

### 2015-04-16 thesis/implementatie.tex
beschrijving implementatie multicore wildcards

@@
 Om het werk effici\"ent over verschillende processoren te verdelen en alle benodigde informatie te communiceren, maakt de implementatie gebruik van zogenaamde subprocesses, die elk hun eigen verbinding openen met het Cassandra-cluster, en interprocess-communicatiemechanismen uit de Python \texttt{multiprocessing}-API \cite{multiprocessing}.\\
+
+\lstinputlisting[language=Python,firstline=186,lastline=210]{../gemini/gemini/query_expressions.py}
+
 \texttt{all-, any-} en \texttt{none-}wildcards vereisen voor het samenvoegen van de resultaten van de verschillende subprocesses geen grote aanpassingen aan de sequenti\"ele evaluatie-strategie van wildcards.
-Omdat deze oplossingswijze er zich makkelijk toe leent af te wijken van het stramien van de geneste \texttt{query\_expressions} (immers, er is maar 1 \texttt{GT\_wildcard\_expression} voor alle samples), is het makkelijker de \texttt{count}-wildcards te implementeren: elk subprocess voert subqueries uit voor de samples waarvoor het verantwoordelijk is en telt voor elke variant hoe vaak hij in de resultaatverzamelingen van de subqueries voorkwam.
-Deze informatie slaat stuurt het subprocess terug naar zijn parent-process in een Python \textit{dictionary}, met een map van \texttt{variant\_id}s naar de telling.
-Het parent-process kan vervolgens de telling voor alle variants van alle subprocesses bij elkaar optellen, en vervolgens met gebruik van de Python \texttt{eval}-functie alle variants die aan de \texttt{count}-filter voldoen, bepalen.
-Let wel, bij het opstellen van de uiteindelijke dictionary met de tellingen voor de varianten, moet de parent-process ook de varianten die in het resultaat van geen enkel subprocess voorkwamen (waarvoor m.a.w.
-geen enkele sample aan de voorwaarden voldoet), een entry met waarde 0 voorzien.
-Dit om ook voor \texttt{count}-filters met \texttt{<, <=, ==0}-operatoren alle varianten te kunnen opvragen.
+Omdat deze oplossingswijze er zich uitstekend toe leent af te wijken van het stramien van de geneste \texttt{query\_expressions} (immers, er is maar 1 \texttt{GT\_wildcard\_expression} voor alle samples), is het makkelijker de \texttt{count}-wildcards te implementeren: elk subprocess voert subqueries uit voor de samples waarvoor het verantwoordelijk is en telt voor elke variant hoe vaak hij in de resultaatverzamelingen van de subqueries voorkomt.
+Deze informatie stuurt het subprocess terug naar zijn parent-process in een Python \textit{dictionary}, met een map van \texttt{variant\_id}s naar de telling.
+
+\lstinputlisting[language=Python,firstline=328,lastline=357]{../gemini/gemini/query_expressions.py}
+
+Het parent-process kan vervolgens de telling voor alle variants van alle subprocesses bij elkaar optellen, en met gebruik van de Python \texttt{eval}-functie alle variants die aan de \texttt{count}-filter voldoen, bepalen.
+Let wel, bij het opstellen van de uiteindelijke \textit{dictionary} met de tellingen voor de varianten, moet het parent-process voor alle varianten in de \texttt{starting\_set}, een entry voorzien, met defaultwaarde 0.
+Zoniet zullen varianten waarvoor geen enkele sample aan de \texttt{gt\_wildcard\_rule} voldoet, niet in het resultaat voorkomen, en \texttt{count}-filters als \texttt{(count < x), (count <= x)} (voor arbitraire, positieve \texttt{x}) of \texttt{(count == 0)}  deze varianten onterecht niet in hun resultaat weergeven.
+
+\lstinputlisting[language=Python,firstline=225,lastline=235]{../gemini/gemini/query_expressions.py}
+
+De \texttt{invert\_count}-flag geeft aan of de \texttt{gt\_wildcard\_rule} een negatie is.
+Gezien Cassandra geen \texttt{!=}-clausules ondersteunt, moet GEMINI in dit geval het complement van de \texttt{gt\_wildcard\_rule} in kwestie bepalen en vervolgens de resulterende telling aftrekken van het totaal aantal samples.
+\\ Een eenvoudig voorbeeld: bij het evalueren van volgende wildcard \\\\
+\texttt{(gt\_type).(*).(!=HET).(count > 100)}\\\\
+zal GEMINI eerst voor elke variant bepalen hoeveel samples w\'el heterozygoot zijn, alvorens dit getal af te trekken van het totaal aantal samples en op dit resultaat de \texttt{count > 100}-filter toe te passen.

### 2015-04-16 gemini/gemini/query_expressions.py

@@ -23,9 +23,9 @@ class Expression(object):
 
 class Basic_expression(Expression):
     
-    def __init__(self, from_table, select_column, where_clause):
+    def __init__(self, from_table, SELECT_column, where_clause):
         self.table = from_table
-        self.select_column = select_column
+        self.SELECT_column = SELECT_column
         self.where_clause = where_clause
   
     def evaluate(self, socket, starting_set):
@@ -34,18 +34,18 @@ class Basic_expression(Expression):
             return set()
         
         query = "SELECT %s FROM %s" % \
-            (self.select_column, self.table)
+            (self.SELECT_column, self.table)
         if self.where_clause != "":
             query += " WHERE %s" % self.where_clause            
         if self.can_prune() and not starting_set == "*":
             if self.table.startswith('samples'):
                 in_clause = "','".join(starting_set)            
                 query += " AND %s IN ('%s')" % \
-                    (self.select_column, in_clause)
+                    (self.SELECT_column, in_clause)
             else:
                 in_clause = ",".join(map(str, starting_set))            
                 query += " AND %s IN (%s)" % \
-                    (self.select_column, in_clause)     
+                    (self.SELECT_column, in_clause)     
         return rows_as_set(socket.execute(query))
     
     def can_prune(self):
@@ -104,10 +104,10 @@ class OR_expression(Expression):
     
 class NOT_expression(Expression):
     
-    def __init__(self, exp, table, select_column):
+    def __init__(self, exp, table, SELECT_column):
         self.body = exp
         self.table = table
-        self.select_column = select_column
+        self.SELECT_column = SELECT_column
  
     def evaluate(self, session, starting_set):
         
@@ -115,7 +115,7 @@ class NOT_expression(Expression):
             return set()        
         elif starting_set == '*':
             correct_starting_set = rows_as_set(session.execute(\
-                "SELECT %s FROM %s" % (self.select_column, self.table)))
+                "SELECT %s FROM %s" % (self.SELECT_column, self.table)))
         else:
             correct_starting_set = starting_set
         
@@ -186,19 +186,22 @@ class GT_wildcard_expression(Expression):
         for i in range(self.nr_cores):
             parent_conn, child_conn = Pipe()
             conns.append(parent_conn)
-            p = Process(target=eval(target_rule +'_query'), args=(child_conn, self.column, corrected_rule,\
[…]

### 2015-04-17 thesis/gemini_beschrijving.tex
kleuren primary key columns vb-tabellen

@@
 Deze informatie komt in het eerste geval eveneens uit de annotatiebestanden, en in het tweede uit tekstbestanden met referentie-informatie over het menselijk genoom die GEMINI, indien gewenst mee inlaadt.\\
-Ten laatste zijn er nog enkele kleine tabellen met meta-informatie, zoals de \texttt{resources}-tabel die de gebruikte annotatie-files bevat, en de \texttt{version}-tabel die bevat door welke versie van GEMINI de data ingeladen is.\\\\
+Ten laatste zijn er nog enkele kleine tabellen met meta-informatie, zoals de \texttt{resources}-tabel die de gebruikte annotatie-files bevat, en de \texttt{version}-tabel die bevat door welke versie van GEMINI de data ingeladen is.\\
 

### 2015-04-17 thesis/implementatie.tex
kleuren primary key columns vb-tabellen

@@
 De inhoud van cellen die niet in de primary key van een rij liggen, heeft voor Cassandra geen enkele betekenis.\\
-Die primary key bestaat uit 2 delen: het eerste is de partition key, deze bestaat uit minstens 1 kolom en de waarde hiervan die via het consistent-hashing mechanisme bepaalt op welke knopen in het cluster de rij terechtkomt.
+Die primary key bestaat uit 2 delen: het eerste is de partition key, deze bestaat uit minstens 1 kolom en de waarde hiervan die via het consistent-hashing mechanisme bepaalt op welke partitie in het cluster de rij terechtkomt.
 Het tweede, optionele, deel is de clustering key, en bepaalt in welke volgorde rijen met dezelfde partition key op 1 node bewaard worden.
@@
 \subsection{Databaseschema GEMINI}
-Zoals blijkt uit bovenstaande beschrijving van het datamodel van Cassandra, leent het systeem zich niet zomaar tot ad-hoc querying: het is niet mogelijk op een performante manier zomaar aan willekeurige kolommen in een tabel voorwaarden op te leggen en deze voorwaarden met elkaar te combineren.
+Zoals blijkt uit bovenstaande beschrijving van het datamodel van Cassandra, leent het systeem zich niet goed tot ad-hoc querying: het is niet mogelijk op een performante manier zomaar aan willekeurige kolommen in een tabel voorwaarden op te leggen en deze voorwaarden met elkaar te combineren.
 Dit betekent dat bij het ontwerpen van het database schema al rekening gehouden moet worden met de queries die achteraf op de data mogelijk moeten zijn.
@@
 \toprule
-variant\_id & ref & alt & \ldots & gt\_type\_bob & gt\_type\_bruce & \ldots & gt\_depth\_bob & gt\_depth\_bruce & \ldots \\ \bottomrule
+\color{green} variant\_id & ref & alt & \ldots & gt\_type\_alex & gt\_type\_john & \ldots & gt\_depth\_alex & gt\_depth\_john & \ldots \\ \bottomrule
 \end{tabular}
@@
 \toprule
-variant\_id & sample\_name & genotype \\ \bottomrule
+\color{green} sample\_name & \color{green} genotype & \color{red} variant\_id \\ \bottomrule
 \end{tabular}\\
@@
 \toprule
-variant\_id & sample\_name & gt\_depth \\ \bottomrule
+\color{green} sample\_name & \color{red} gt\_depth & \color{red} variant\_id \\ \bottomrule
 \end{tabular}
@@
 \toprule
- chrom & start & variant\_id \\ \bottomrule
+\color{green} chrom & \color{red} start & \color{red} variant\_id \\ \bottomrule
 \end{tabular}
@@
 \toprule
- gene & variant\_id \\ \bottomrule
+\color{green} gene & \color{red} variant\_id \\ \bottomrule
 \end{tabular}
@@
 \toprule
- sex & sample\_name \\ \bottomrule
+\color{green} sex & \color{red} sample\_name \\ \bottomrule
 \end{tabular}
@@
 
+\noindent De \texttt{correct\_starting\_set} in bovenstaande code is de traditionele \texttt{starting\_set}, maar ge\"evalueerd in het geval dit de volledige verzameling varianten (dus de wildcard \texttt{"*"}) is, en gegoten in een datatype geschikt voor shared-memory access.\\
 De \texttt{invert\_count}-flag geeft aan of de \texttt{gt\_wildcard\_rule} een negatie is.
@@
 \texttt{(gt\_type).(*).(!=HET).(count > 100)}\\\\
-zal GEMINI eerst voor elke variant bepalen hoeveel samples w\'el heterozygoot zijn, alvorens dit getal af te trekken van het totaal aantal samples en op dit resultaat de \texttt{count > 100}-filter toe te passen.
+zal GEMINI eerst voor elke variant bepalen hoeveel samples w\'el heterozygoot zijn, alvorens dit getal af te trekken van het totaal aantal samples en op dit resultaat de \texttt{(count > 100)}-filter toe te passen.

### 2015-04-20 thesis/implementatie.tex
keuze database

@@
 
-Zoals beschreven in \ref{gemini_beschrijving} bewaart GEMINI de data over genetische varianten en proefpersonen in enkele zeer grote tabellen.
+Zoals beschreven in [\ref{gemini_beschrijving}] bewaart GEMINI de data over genetische varianten en proefpersonen in enkele zeer grote tabellen.
 Een eerste vereiste voor een database is dus om deze tabulaire data goed te kunnen voorstellen en beheren.
@@
 
-Denk denk denk: Cassandra niet beste optie, maar toch gekozen. Eerste prioriteit schaalbaarheid?
+De uiteindelijke keuze voor een databank viel op Apache Cassandra.
+Van de systemen besproken in [\ref{nosql_survey}] heeft Cassandra veruit de meest interessante architectuur qua schaalbaarheid en bovendien leent het columnaire model zich goed tot het modelleren van de tabellen in de oorspronkelijke versie van GEMINI.
+Vergeleken met het andere columnaire systeem, HBase, heeft Cassandra uitgebreidere query-mogelijkheden.
+Vergeleken met MongoDB, dat sterker staat op het gebied van querying en indexing heeft Cassandra het voordeel dat het flexibeler is voor concurrent schrijven naar de databank.
+Bovendien blijkt uit eerdere experimenten met MongoDB binnen het lab dat MongoDB niet space-effici\"ent BSON-documenten kan uitbreiden.
+In het geval van een incrementele versie van GEMINI, waar na verloop van tijd genoominformatie van extra proefpersonen ingeladen moet worden, leidt dit tot fragmentatie.
+Cassandra daarentegen kan eenvoudig het schema van tabellen aanpassen en naar behoefte extra kolommen en rijen toevoegen.\\
+Zoals later uitgebreid aan bod komt, heeft Cassandra ook nadelen.
+Vooral de eerder beperkte query-capaciteiten bemoeilijken het implementeren van de uitgebreide ad-hoc queryfunctionaliteiten van GEMINI.
 

### 2015-04-20 thesis/Inleiding.tex
inleiding GEMINI uitgebreid + prentjeuh sequencing cost

@@
 \chapter{Inleiding}
-\label{cha:1}
+\label{inleiding}
 
@@
 Wetenschappelijke vooruitgang heeft ervoor gezorgd dat de kost om genomen te sequencen het afgelopen decennium exponentieel gedaald is, sinds 2008 zelfs aan een hogere snelheid dan de evolutie volgens de wet van Moore \cite{wetterstrand_sequencing_cost}.
+Dit is duidelijk zichtbaar op de grafiek in figuur \ref{sequencing_cost}.
 In allerhande soorten biologisch, medisch en pharmaceutisch onderzoek worden dan ook genomen van meer en meer organismen gesequenced en dit genereert enorme hoeveelheden data.
@@
 Naarmate genomen van miljoenen mensen en andere levende wezens geanalyseerd en opgeslagen worden, vereist deze evolutie steeds betere schaalbaarheid, responstijd, en parallellisering voor de opslag en verwerking van deze data.\\
+
 Een logische stap is om deze problemen aan te pakken met grote verdeelde computersystemen, zogenaamde \textit{high performance computing systems} of \textit{supercomputers}.
 Het Exascience Life Lab van imec, Intel, Janssen Pharmaceutica en de 5 Vlaamse universiteiten verricht specifiek onderzoek naar de toepassing van supercomputers om het genoomsequencingproces te versnellen \cite{lifelab_bwa}\cite{exascience_life_lab}.\\
-De snel toegenomen populariteit van webdiensten als sociale netwerken zadelde webbedrijven op met een gelijkaardige explosie aan data.
+De snel toegenomen populariteit van webdiensten als sociale netwerken zadelde webservice-leveranciers op met een gelijkaardige explosie aan data.
 Om deze zogenaamde Big Data \cite{mashey1997big} adequaat te beheren, volstaan traditionele relationele DBMS niet meer.
@@
 
+\begin{figure}[h!]
+\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{cost_genome}
+\caption{Evolutie van de prijs om een genoom te sequencen sinds 2001}
+\label{sequencing_cost}
+\end{figure}
+
 \section{GEMINI}
 
-GEMINI (van GEnome MINIng) is een tool voor flexibele analyse van genetische variaties in grote datasets van genomen, ontwikkeld aan de University of Virginia \cite{10.1371/journal.pcbi.1003153}.
+Deze thesis bevat ook een case study over de opportuniteiten voor NoSQL-technologie\"en in de genoomanalysewereld.
+Op vraag van Janssens Pharmaceutica is het onderwerp hiervan de genoomanalysetool GEMINI (van GEnome MINIng): een software-framework voor flexibele analyse van genetische variaties in grote datasets van genomen, ontwikkeld aan de University of Utah \cite{10.1371/journal.pcbi.1003153}.
 GEMINI vertrekt daarbij van VCF (Variant Call Format) tekst-bestanden, laadt deze in in een database, voegt allerlei annotaties toe met bijkomende informatie van enkele vermaarde onderzoeksinstituten en biedt dan de mogelijkheid queries uit te voeren op deze databank.
 De huidige operationele versie van GEMINI draait op de eenvoudige SQL-databank SQLite, maar deze laat op gebied van schaalbaarheid en performantie te wensen over.
+De gewenste schaalbaarheid ligt in 2 dimensies: enerzijds het aantal genetische \textit{variants} in de dataset, d.w.z.
+stukken van genen zoals deze in het genoomsequencingproces waargenomen worden, en het aantal proefpersonen of \textit{samples} wier DNA in de dataset geanalyseerd wordt.
+Bij toenemend
+%TODO getal!
+aantal samples of variants duurt zowel het inladen van de genoomdata in de databank voor als het query'en tijdens effectief gebruik te lang om gebruiksvriendelijk te zijn.
+Het inladen van de genoomdata uit VCF-files is, enigszins contra-intu\"itief, geen eenmalige of zeldzame operatie: de huidige productieversie van GEMINI laat niet toe om na het inladen nog samples of variants toe te voegen, met als gevolg dat bij elke iteratie van een genetisch of biologisch experiment die nieuwe informatie oplevert, de volledige dataset opnieuw ingeladen dient te worden.
+Ook het incrementeel kunnen toevoegen van variants en samples aan de gegevensset, is dus een aandachtspunt bij het onderzoek naar mogelijke verbeteringen voor GEMINI.\\
 Een eerste oplossing om de performantie te verhogen was om over te schakelen op PostgreSQL, en hiervan is intussen al met succes een eerste versie ge\"implementeerd.

### 2015-04-21 thesis/Inleiding.tex
inleiding NoSQL uitgebreid

@@
 Het inladen van de genoomdata uit VCF-files is, enigszins contra-intu\"itief, geen eenmalige of zeldzame operatie: de huidige productieversie van GEMINI laat niet toe om na het inladen nog samples of variants toe te voegen, met als gevolg dat bij elke iteratie van een genetisch of biologisch experiment die nieuwe informatie oplevert, de volledige dataset opnieuw ingeladen dient te worden.
-Ook het incrementeel kunnen toevoegen van variants en samples aan de gegevensset, is dus een aandachtspunt bij het onderzoek naar mogelijke verbeteringen voor GEMINI.\\
+Ook het incrementeel kunnen toevoegen van variants en samples aan de gegevensset is dus een aandachtspunt bij het onderzoek naar mogelijke verbeteringen voor GEMINI.\\
 Een eerste oplossing om de performantie te verhogen was om over te schakelen op PostgreSQL, en hiervan is intussen al met succes een eerste versie ge\"implementeerd.

### 2015-04-21 thesis/NoSQL_overzicht.tex
inleiding NoSQL uitgebreid

@@
 
-\section{NoSQL}
+Sinds de jaren '70 zijn zogenaamde \textit{relational database management systems} (kortweg RDBMS) de voornaamste technologie voor de grootschalige opslag van gegevens.
+Ze zijn gestoeld op 2 belangrijke principes, namelijk het relationele datamodel \cite{codd1970relational} en de gestructureerde querytaal SEQUEL, beter gekend als SQL \cite{chamberlin1974sequel}.
+De architectuur van vele RDBMS is nog steeds gebaseerd op de eerste implementatie van een dergelijk systeem, namelijk het IBM onderzoeksproject System R \cite{blasgen1981system}, ook uit halverwege de jaren '70.
+System R is uiteraard ontworpen voor destijds relevante hardwarekarakteristieken en productvereisten: business data processing via een command line interface, en dit op computersystemen met trage processoren, kleine werk- en schijfgeheugens maar relatief grote bandbreedte tussen de schijfopslag en het werkgeheugen \cite{Stonebraker:2007:EAE:1325851.1325981}.
+Dit leidde tot een aantal architecturale features die nog steeds terug te vinden zijn in hedendaagse RDBMS:
+\begin{itemize}
+\item Disk-geori\"enteerde opslag- en indexstructuren
+\item Multithreading om latency te verbergen
+\item Concurrency-controlemechanismen op basis van locking
+\item Log-gebaseerd herstel van fouten
+\end{itemize} 
+Ondanks gigantische technologische vooruitgang op gebied van hardware en sterk gediversifieerde gebruiksscenario's, is er sinds hun ontstaan 40 jaar geleden weinig drastisch veranderd aan het concept van de RDBMS en zijn deze systemen de werkpaarden van de industrie geworden op het vlak van dataopslag.\\
+
+Beginnende in de jaren 2000 groeide in de IT-wereld dan ook het besef dat de rigide "one size fits all"-aanpak van RDBMS voor vele moderne toepassingen achterhaald dreigde te geraken.
+Met grote opkomende spelers uit de <eb-industrie zoals Google, Amazon en Facebook aan het roer leidde dit tot de opkomst van de NoSQL-beweging.
+NoSQL staat, in tegenstelling tot wat de naam doet vermoeden, voor "Not only SQL" en omvat een waaier van uiteenlopende alternatieve gegevensopslagsystemen die elk in bepaalde specifieke opzichten en zorgvuldig uitgekozen toepassingscenario's een meerwaarde trachten te bieden ten opzichte van het klassieke relationele model.
+
+\section{NoSQL-klassen}
 
@@
 In het geval van BigTable zijn dit drie dimensies: row key, column key en een timestamp.
-Omdat ook hier het systeem de opgeslagen data niet interpreteert, is het modelleren van relaties niet po een effici\"ente manier mogelijk.
+Omdat ook hier het systeem de opgeslagen data niet interpreteert, is het modelleren van relaties niet op een effici\"ente manier mogelijk.
 Dit wordt overgelaten aan de bovenliggende applicatie\cite{hecht2011nosql}.

### 2015-04-21 thesis/NoSQL_overzicht.tex
NoSQL achtergrond: consistentie

@@
 Met grote opkomende spelers uit de <eb-industrie zoals Google, Amazon en Facebook aan het roer leidde dit tot de opkomst van de NoSQL-beweging.
-NoSQL staat, in tegenstelling tot wat de naam doet vermoeden, voor "Not only SQL" en omvat een waaier van uiteenlopende alternatieve gegevensopslagsystemen die elk in bepaalde specifieke opzichten en zorgvuldig uitgekozen toepassingscenario's een meerwaarde trachten te bieden ten opzichte van het klassieke relationele model.
+NoSQL staat, in tegenstelling tot wat de naam doet vermoeden, voor "Not only SQL" en omvat een waaier van uiteenlopende alternatieve gegevensopslagsystemen die elk in bepaalde specifieke opzichten meerwaarde trachten te bieden ten opzichte van het klassieke relationele systemen.
+In tegenstelling tot de 'Zwitsers zakmes'-aanpak van RDMBS, leggen ze zich toe op zeer gespecialiseerde toepassingsdomeinen en proberen daarin relationele systemen te overtreffen.
+Vaak betekent dit dat NoSQL systemen vele voor hun doel onnodig geachte features van SQL systemen achterwege laten, of afzwakken.
+Een goed voorbeeld hiervan zijn de ACID-eigenschappen uit het relationele model die in vele NoSQL-systemen gereduceerd zijn tot zogenaamde BASE-eigenschappen (op het verschil tussen beide komt onderstaande sectie nog uitgebreid terug).
+
+\section{Begrippen}
+
+Deze sectie belicht de belangrijkste technische begrippen in verband met NoSQL-databanken en contrasteert waar nodig met gelijkaardige concepten in het klassieke traditionele model.
+
+\paragraph{Consistentie}
+
+Consistentie van database-transacties betekent dat transacties de databank in een consistente staat achterlaten: alle data die een applicatie kan zien, is een consistente snapshot van de databank \cite{ports2010transactional}.
+Traditionele RDBMS bieden vaak transacties met de zogenaamde ACID-eigenschappen \cite{haerder1983principles}:
+\begin{itemize}
+\item \textbf{Atomicity:} Elke transactie gebeurt ofwel volledig, ofwel helemaal niet.
+\item \textbf{Consistency:} Elke transactie laat de databank in consistente staat achter.
+\item \textbf{Isolation:} Elke transactie verloopt volledig ge\"isoleerd van elke andere transactie en be\"invloedt deze dus op geen enkele manier.
+\item \textbf{Durability:} Eens voltrokken, blijft elke transactie duurzaam bewaard in de databank, ook in het geval van stroomonderbrekingen, crashes of fouten.
+\end{itemize}
+
+Zoals Eric Brewer stelde in zijn bekende CAP-theorema \cite{brewer2000towards}, is het in een gedistribueerd systeem niet eenvoudig zowel consistentie, availability als tolerantie voor partities te bereiken en zijn 2 van deze 3 eigenschappen het hoogst haalbare\footnote{Brewer kwam hier zelf 12 jaar later op terug, stellende dat mits goede omgang met partities het toch mogelijk is een trade-off van alle drie te bereiken \cite{brewer2012cap}.}.
+Ook in NoSQL-systemen, die vaak gedistribueerd van aard zijn, is het garanderen van consistentie geen triviale opgave.
+Afhankelijk van de gehanteerde schrijfstrategie is het mogelijk dat verschillende knopen in het cluster verschillende versies van data zien, als updates nog niet in het volledige cluster doorgekomen zijn.
+Daarom is er het onderscheid tussen strikte en uiteindelijke ("\textit{eventual}") consistentie: strikte consistentie is de gekende vorm waarin updates onmiddellijk zichtbaar zijn op alle nodes in het cluster, en dus ook naar bovenliggende applicaties toe.
+In het geval van uiteindelijke consistentie garandeert het systeem enkel dat na verloop van tijd alle nodes in het cluster dezelfde, up-to-date versie van de data zullen zien.
+NoSQL-systemen bieden dan ook vaak de BASE-eigenschappen, een zwakkere versie van de ACID-garanties:
+\begin{itemize}
+\item \textbf{Basically available:} Het systeem is onder quasi alle omstandigheden beschikbaar.
+\item \textbf{Soft state:} Het systeem verkeert niet altijd in een consistente staat
+\item \textbf{Eventually consistent:} Na verloop van tijd zal het systeem in een gekende staat verkeren.
+\end{itemize}
+
+Vele NoSQL-systemen stellen de gebruiker echter niet voor een voldongen feit bij de keuze tussen strikte en uiteindelijke consistentie: dankzij zogenaamde \textit{quora} kan de gebruiker zelf configureren welke consistentie het systeem levert.
+Door lees- en schrijfquora in te stellen, kan de gebruiker tunen hoeveel replica's respectievelijk moeten returnen bij een schrijfopdracht, en het welslagen van een schrijfopdracht moeten bevestigen.
+Op deze manier kan de gebruiker zelf een trade-off maken tussen snel lezen, schrijven en de behaalde consistentie.
+Bovendien zal de gebruiker, wanneer de som van het lees- en schrijfquorum groter is dan de replicatiefactor van het cluster, steeds de meest recente versie van gegevens zien, wat hetzelfde betekent als onmiddellijke, strikte consistentie.
 

### 2015-04-21 thesis/nosql_survey.tex
NoSQL achtergrond: consistentie

@@
 Het systeem biedt krachtige ondersteuning voor indices, vooral door de mogelijkheid om secundaire indices van een brede waaier van types te defini\"eren op alle attributen, zoals in het relationele model.
-Deze indices zijn gebouwd op B-trees\cite{mongodb_indexes}.
+Deze indices zijn gebouwd op B-trees \cite{mongodb_indexes}.
 Om denormalizatie te bevorderen, kunnen documenten geneste documenten en arrays bevatten.
@@
 Het is gericht op flexibele, quasipermanent beschikbare opslag van zeer grote datasets op goedkope standaardhardware, met daarenboven hoge througput voor schrijfopdrachten zonder effici\"entie bij leesopdrachten op te offeren \cite{borthakur2011apache}.\\
-Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model\cite{cassandra_then&now}, in die zin dat het nu tabellen en samengestelde kolommen
+Sinds zijn ontstaan is Cassandra wel op enkele vlakken afgeweken van het BigTable-model \cite{cassandra_then&now}, in die zin dat het nu tabellen en samengestelde kolommen
 %TODO betere vertaling voor composite columns?
@@
 Cassandra heeft indexeringsmechanismes and implementeert deze met log-structured merge trees, met hogere schrijfthroughput als gevolg.
-%TODO hier uitleggen of eerder meer nr boven?
-Net als BigTable biedt Cassandra ook Bloom filters.\\
+Net als BigTable biedt Cassandra ook Bloom filters: een effici\"ent probabilistische mechanisme om te voorspellen of een object in een verzameling zit (in dit geval, of een key in een tabel ligt, dat het aantal nodeloze table scans beduidend kan inperken \cite{mullin1983second}.\\
 Om lineair in het aantal ingeschakelde nodes te kunnen schalen naar zeer grote datasets, opereert Cassandra op een volledig hi\"erarchieloze wijze.
@@
 Het aantal replica's kan de gebruiker zelf kiezen.
-Bovendien voorziet Cassandra ook interdatacenterreplicatie, om zelfs het falen van volledige datacenters op te vangen\cite{decandia2007dynamo} \cite{lakshman2010cassandra} \cite{cassandra_then&now}.\\
+Bovendien voorziet Cassandra ook interdatacenterreplicatie, om zelfs het falen van volledige datacenters op te vangen \cite{decandia2007dynamo} \cite{lakshman2010cassandra} \cite{cassandra_then&now}.\\
 Bij lees- en schrijfopdrachten kan de gebruiker een quorum specifi\"eren.
@@
 Sinds z'n lancering heeft HBase verschillende secundaire indexeringsmechanismes verworven.
-Deze zijn ook gebaseerd op LSM trees en daarnaast biedt ook HBase Bloom filters\cite{borthakur2011apache}\cite{hbase_schema}.
+Deze zijn ook gebaseerd op LSM trees en daarnaast biedt ook HBase Bloom filters \cite{borthakur2011apache}\cite{hbase_schema}.
 HBase heeft een Java API, maar zonder SQL-achtige geavanceerde querytaal.
@@
 
-VoltDB is een relationele, gedistribueerde in-memory databank, die als doel heeft de garanties van klassieke SQL-stores te koppelen met de schaalbaarheid van NoSQL-systemen, en bovendien aan zeer hoge snelheden te functioneren\cite{stonebraker2013voltdb}.\\
-VoltDB bewaart gegevens in het traditionele relationele model, maar gerepliceerd en gepartitioneerd (met consistent hashing) over verschillende nodes\cite{grolinger2013data}.
-De data is doorzoekbaar via een (groeiende) subset van SQL-92\cite{voltdb2010voltdb}.
+VoltDB is een relationele, gedistribueerde in-memory databank, die als doel heeft de garanties van klassieke SQL-stores te koppelen met de schaalbaarheid van NoSQL-systemen, en bovendien aan zeer hoge snelheden te functioneren \cite{stonebraker2013voltdb}.\\
+VoltDB bewaart gegevens in het traditionele relationele model, maar gerepliceerd en gepartitioneerd (met consistent hashing) over verschillende nodes \cite{grolinger2013data}.
+De data is doorzoekbaar via een (groeiende) subset van SQL-92 \cite{voltdb2010voltdb}.
 Queries worden bij voorkeur gedefinieerd als stored procedures in Java, met daarin ingebed de SQL uitdrukkingen.
-VoltDB ondersteunt primaire en secundaire indices en laat de gebruiker de keuze tussen hash- en boomindices\cite{voltdb_indexes}.
-VoltDB plant en optimaliseert de queries in stored procedures offline tijdens het compileren\cite{voltdb_query_plans}.\\
+VoltDB ondersteunt primaire en secundaire indices en laat de gebruiker de keuze tussen hash- en boomindices \cite{voltdb_indexes}.
+VoltDB plant en optimaliseert de queries in stored procedures offline tijdens het compileren \cite{voltdb_query_plans}.\\
 Omdat VoltDB volledig op RAM geheugen vertrouwt, is het duur om te schalen naar datavolumes in de grootorde van meerdere petabytes, maar VoltDB kan data exporteren naar andere, meer geschikte databanksystemen zoals columnaire NoSQL systemen.\\
@@
 De gebruiker krijgt wel de keuze hiertoe, voor in het geval de hoeveelheid gegevens hierom zou vragen \cite{impala_partitioning}.\\
-Impala beschikt over een zeer effici\"ente I/O-laag die schijf- en CPU-gebruik te allen tijde hoog houdt, wat resulteert in aanzienlijk snellere prestaties dan andere SQL-on-Hadoop oplossingen zoals Apache Hive\cite{floratou2014sql}.
+Impala beschikt over een zeer effici\"ente I/O-laag die schijf- en CPU-gebruik te allen tijde hoog houdt, wat resulteert in aanzienlijk snellere prestaties dan andere SQL-on-Hadoop oplossingen zoals Apache Hive \cite{floratou2014sql}.
 Het inherente nadeel is echter dat de volledige dataset moet passen in het totale werkgeheugen van het cluster waarin Impala draait.

### 2015-04-22 gemini/gemini_update.py
moved thesis text to separate repo

@@ -10,8 +10,8 @@ import gemini.config
 def release(parser, args):
     """Update gemini to the latest release, along with associated data files.
     """
-    url = "https://raw.github.com/arq5x/gemini/master/requirements.txt"
-    repo = "https://github.com/arq5x/gemini.git"
+    url = "https://github.com/bgossele/gemini_cassandra/master/requirements.txt"
+    repo = "https://github.com/bgossele/gemini_cassandra"
     cbl_repo = "https://github.com/chapmanb/cloudbiolinux.git"
     # update locally isolated python
     base = os.path.dirname(os.path.realpath(sys.executable))
